### 2020-11-11 paper.tex
initial commit with intro and outline

\documentclass[11pt]{article}
\usepackage[round,comma,authoryear]{natbib}
\usepackage{fullpage}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage[newfloat]{minted}
\usepackage{tcolorbox}
\usepackage{etoolbox}
\BeforeBeginEnvironment{minted}{\begin{tcolorbox}}
\AfterEndEnvironment{minted}{\end{tcolorbox}}
\usepackage{url}
\usepackage{fancyvrb}
\usepackage{caption}
\newenvironment{code}{\captionsetup{type=listing}\centering}{}
\SetupFloatingEnvironment{listing}{name=Demographic Model}

% local definitions
\newcommand{\msprime}[0]{\texttt{msprime}}
\newcommand{\stdpopsim}[0]{\texttt{stdpopsim}}
\newcommand{\demes}[0]{\texttt{demes}}
\newcommand{\Demes}[0]{\texttt{Demes}}
\newcommand{\YAML}[0]{\texttt{YAML}}
\newcommand{\moments}[0]{\texttt{moments}}
\newcommand{\dadi}[0]{\texttt{$\partial$a$\partial$i}}
\newcommand{\fwdpy}[0]{\texttt{fwdpy11}}
\newcommand{\slim}[0]{\texttt{slim}}

\newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}

\begin{document}

\title{\demes: a Python package for pain-free specification of complex demography}
\author[1,*]{Graham Gower}
\author[2,*]{Jerome Kelleher}
\author[3,*]{Aaron P. Ragsdale}
\author[4,*]{Kevin Thornton}
\affil[1]{}
\affil[2]{}
\affil[3]{}
\affil[4]{}
\affil[*]{Authors listed alphabetically}
\maketitle

\abstract{
come back to title and abstract later
}

\section*{Introduction}

The ever-increasing amount of genetic sequencing data from genetically and
geographically diverse species and populations has allowed us to
infer complex demography and study life history at fine scales. An integral
component to such population genetics studies is simulation. Software to
either simulate whole genome sequences
\citep{thornton2014c++,kelleher2016efficient,haller2019slim,adrion2020community}
or informative summary statistics of diversity
\citep{gutenkunst2009inferring,kamm2017efficient,jouganous2017inferring}
have enabled the increasing complexity of genomic studies, with many software
packages able to handle large sample sizes, many interacting populations, and
deviations from random-mating models of panmixia.

The ability to simulate complex demographic scenarios, however, does not come
without its own set of obstacles and frustrations. First of all, specifying models
with many populations and parameters can be cumbersome and error-prone
(e.g., \citep{ragsdale2020lessons}). It is
time-consuming and tedious to implement and then verify the correctness
of such demographic models. Making matters worse, each software package typically has
its own application programming interface (API), which makes it difficult to translate
models between simulation methods.

Secondly, many software API are not particularly easy to read, especially when there
are a large number of demographic events and parameters. This not only makes
debugging difficult, but it can be difficult to even notice if a mistake has been
made. A common interface for unambiguously specifying demographic
models would reduce implementation errors and promote ease-of-use for many simulation
software packages.

Finally, simulated or inferred demographic models are regularly shared in the
literature, and for someone else to be able to reproduce or reimplement some
demographic scenario, it needs to be described fully and without ambiguity. However,
it is common for reported demographic models to lack clear or even complete
descriptions, hindering reproducibility. A human-readable description for
such models would facilitate their sharing and reimplementation.

Here, we present \demes, a Python package for specifying complex demographic models
in a way that is both human readable and can be directly passed to a growing
number of simulation engines. \Demes\ supports model specification in a high-level
\YAML\ format \citep{ben2009yaml}
that is designed to be as easy as possible to implement, read, and
share. Models are then represented in an unambiguous low-level format within Python
that checks model validity and can then be used as the input for simulation software.
The initial release of \demes\ (version 1.0, described here) supports \emph{static}
demographic descriptions, that is, models with fixed parameters. Future releases
will support [drawing parameters from distributions, for example for ABC-based
inference, or describing ``dynamic`` models where parameters may be fit to data,
e.g. \moments\ or \dadi].

\section*{Defining demographic models with \demes}

\subsection*{High-level model specification in \YAML}

\begin{itemize}
\item Short description of YAML language and why it's appropriate here
\item General overview (1 paragraph) of defining demography in YAML
\item A generic, commonly used, simple demographic model for illustration (e.g. IM)
\end{itemize}

\begin{code}
\label{code:im_model}
\captionof{listing}{An IM model.}
\begin{minted}[linenos,numbersep=5pt]{yaml}
description: A two-population isolation-with-migration model.
time_units: generations
demes:
  ancestral:
    description: The ancestral deme that splits into two child demes.
    initial_size: 5000
    end_time: 2000
  deme1:
    description: The first child deme.
    ancestors: ancestral
    initial_size: 1000
    final_size: 10000
  deme2:
    description: The second child deme.
    ancestors: ancestral
    initial_size: 2000
    final_size: 8000
migrations:
  symmetric:
  - demes: deme1, deme2
    rate: 1e-4
\end{minted}
\end{code}

\subsection*{Python representation of demography as input for simulation software}

\begin{itemize}
\item The Python API, reading in a YAML, and unambiguous and manipulable representation
\end{itemize}

\subsubsection*{Software support for \demes}
\aprcomment{at time of publication. TBD...}
\begin{itemize}
\item \msprime
\item \stdpopsim
\item \fwdpy
\item \moments
\item \dadi? should try to get Ryan on board
\item \slim? I guess that comes with \stdpopsim
\end{itemize}

\subsection*{Specifying complex demography}

\aprcomment{A few examples of complex demography - e.g. 1) the very common out-of-Africa
model, 2) Denisovan admixture with a ton of populations and demographic events, 3) a
non-human model, perhaps with selfing rates and rate changes}

\aprcomment{Perhaps examples should be in the appendix, to keep the paper concise}

\section*{Prospects and future directions}

\bibliographystyle{plainnat}
\bibliography{paper}

\section*{Appendix}

\end{document}

### 2020-11-11 paper.tex
input model code from yaml

@@
 \item General overview (1 paragraph) of defining demography in YAML
-\item A generic, commonly used, simple demographic model for illustration (e.g. IM)
+\item A generic, commonly used, simple demographic model for illustration (e.g.
+  an IM in Model~\ref{code:im_model})
 \end{itemize}
@@
 \begin{code}
+\captionof{listing}{An IM model.}
+\begin{tcolorbox}
+\inputminted[linenos,numbersep=5pt]{yaml}{models/IM.yaml}
+\end{tcolorbox}
 \label{code:im_model}
-\captionof{listing}{An IM model.}
-\begin{minted}[linenos,numbersep=5pt]{yaml}
-description: A two-population isolation-with-migration model.
-time_units: generations
-demes:
-  ancestral:
-    description: The ancestral deme that splits into two child demes.
-    initial_size: 5000
-    end_time: 2000
-  deme1:
-    description: The first child deme.
-    ancestors: ancestral
-    initial_size: 1000
-    final_size: 10000
-  deme2:
-    description: The second child deme.
-    ancestors: ancestral
-    initial_size: 2000
-    final_size: 8000
-migrations:
-  symmetric:
-  - demes: deme1, deme2
-    rate: 1e-4
-\end{minted}
 \end{code}

### 2021-04-23 paper.tex
some edits
some cleanup, getting ready for editing...

@@
 \usepackage{graphicx}
+
 \usepackage[newfloat]{minted}
@@
 \SetupFloatingEnvironment{listing}{name=Demographic Model}
+% can either input code directly in a minted environment
+% or input from a source file using \inputminted
 
@@
 The ever-increasing amount of genetic sequencing data from genetically and
-geographically diverse species and populations has allowed us to
-infer complex demography and study life history at fine scales. An integral
-component to such population genetics studies is simulation. Software to
-either simulate whole genome sequences
+geographically diverse species and populations has allowed us to infer complex
+demography and study life history at fine scales. An integral component to such
+population genetics studies is simulation. Software to either simulate whole
+genome sequences
 \citep{thornton2014c++,kelleher2016efficient,haller2019slim,adrion2020community}
 or informative summary statistics of diversity
-\citep{gutenkunst2009inferring,kamm2017efficient,jouganous2017inferring}
-have enabled the increasing complexity of genomic studies, with many software
+\citep{gutenkunst2009inferring,kamm2017efficient,jouganous2017inferring} have
+enabled the increasing complexity of genomic studies, with many software
 packages able to handle large sample sizes, many interacting populations, and
@@
 The ability to simulate complex demographic scenarios, however, does not come
-without its own set of obstacles and frustrations. First of all, specifying models
-with many populations and parameters can be cumbersome and error-prone
-(e.g., \citep{ragsdale2020lessons}). It is
-time-consuming and tedious to implement and then verify the correctness
-of such demographic models. Making matters worse, each software package typically has
-its own application programming interface (API), which makes it difficult to translate
-models between simulation methods.
+without its own set of obstacles and frustrations. First, specifying models
+with many populations and parameters can be cumbersome and error-prone (e.g.,
+\citep{ragsdale2020lessons}). It is time-consuming and tedious to implement and
+then verify the correctness of such demographic models. Making matters worse,
+each software package typically has its own application programming interface
+(API), which makes it difficult to translate models between simulation
+software.
 
-Secondly, many software API are not particularly easy to read, especially when there
-are a large number of demographic events and parameters. This not only makes
-debugging difficult, but it can be difficult to even notice if a mistake has been
-made. A common interface for unambiguously specifying demographic
-models would reduce implementation errors and promote ease-of-use for many simulation
-software packages.
+Second, many software API are not particularly easy to read, especially when
+there are a large number of demographic events and parameters. This not only
+makes debugging difficult, but it can be difficult to even notice if a mistake
+has been made. A common interface for unambiguously specifying demographic
+models would reduce implementation errors and promote ease-of-use for many
+simulation software packages.
 
@@
 literature, and for someone else to be able to reproduce or reimplement some
-demographic scenario, it needs to be described fully and without ambiguity. However,
-it is common for reported demographic models to lack clear or even complete
-descriptions, hindering reproducibility. A human-readable description for
-such models would facilitate their sharing and reimplementation.
+demographic scenario, it needs to be described fully and without ambiguity.
+However, it is common for reported demographic models to lack clear or even
+complete descriptions, hindering reproducibility. A human-readable description
+for such models would facilitate their sharing and reimplementation.
 
-Here, we present \demes, a Python package for specifying complex demographic models
-in a way that is both human readable and can be directly passed to a growing
-number of simulation engines. \Demes\ supports model specification in a high-level
-\YAML\ format \citep{ben2009yaml}
-that is designed to be as easy as possible to implement, read, and
-share. Models are then represented in an unambiguous low-level format within Python
-that checks model validity and can then be used as the input for simulation software.
-The initial release of \demes\ (version 1.0, described here) supports \emph{static}
-demographic descriptions, that is, models with fixed parameters. Future releases
-will support [drawing parameters from distributions, for example for ABC-based
-inference, or describing ``dynamic`` models where parameters may be fit to data,
-e.g. \moments\ or \dadi].
+Here, we present \demes, a Python package for specifying complex demographic
+models in a way that is both human readable and can be directly passed to a
+growing number of simulation engines. \Demes\ supports model specification in a
+high-level \YAML\ format \citep{ben2009yaml} that is designed to be as easy as
+possible to implement, read, and share. Models are then represented in an
+unambiguous low-level format within Python that checks model validity and can
+then be used as the input for simulation software.  The initial release of
+\demes\ (version 0.1, described here) supports \emph{static} demographic
+descriptions, that is, models with fixed parameters. Future releases will
+support [drawing parameters from distributions, for example for ABC-based
+inference, or describing ``dynamic`` models where parameters may be fit to
+data, e.g. \moments\ or \dadi].
 
@@
 \begin{code}
-\captionof{listing}{An IM model.}
 \begin{tcolorbox}
@@
 \end{tcolorbox}
+\captionof{listing}{An IM model.}
 \label{code:im_model}

### 2021-04-30 paper.tex

@@
 
-\title{\demes: a Python package for pain-free specification of complex demography}
-\author[1,*]{Graham Gower}
-\author[2,*]{Jerome Kelleher}
-\author[3,*]{Aaron P. Ragsdale}
-\author[4,*]{Kevin Thornton}
-\affil[1]{}
-\affil[2]{}
-\affil[3]{}
-\affil[4]{}
-\affil[*]{Authors listed alphabetically}
+\title{\demes: a pain-free specification for complex demography}
+\author[*]{Graham Gower}
+\author[*]{Aaron P. Ragsdale}
+\author[ ]{...Others here...}
+\author[**]{Kevin Thornton}
+\author[**]{Jerome Kelleher}
+\affil[*,**]{Denote equal contribution}
 \maketitle
@@
 \abstract{
-come back to title and abstract later
 }
@@
 
-Here, we present \demes, a Python package for specifying complex demographic
-models in a way that is both human readable and can be directly passed to a
-growing number of simulation engines. \Demes\ supports model specification in a
-high-level \YAML\ format \citep{ben2009yaml} that is designed to be as easy as
-possible to implement, read, and share. Models are then represented in an
-unambiguous low-level format within Python that checks model validity and can
-then be used as the input for simulation software.  The initial release of
-\demes\ (version 0.1, described here) supports \emph{static} demographic
-descriptions, that is, models with fixed parameters. Future releases will
-support [drawing parameters from distributions, for example for ABC-based
-inference, or describing ``dynamic`` models where parameters may be fit to
-data, e.g. \moments\ or \dadi].
+Here, we present \demes, a specification of complex demographic models that is
+both human readable and can be directly passed to a growing number of
+simulation software. \Demes\ supports model specification in a high-level
+\YAML\ format \citep{ben2009yaml} that is designed to be as easy as possible to
+implement, read, and share. Models are then represented in an unambiguous
+low-level format within Python that checks model validity. The initial Python
+release of \demes\ (version 0.1, described here) supports \emph{static}
+demographic descriptions, that is, models with fixed parameters. Future
+releases will support [[drawing parameters from distributions, for example for
+ABC-based inference, or describing ``dynamic`` models where parameters may be
+fit to data, e.g. \moments\ or \dadi]].
 
-\section*{Defining demographic models with \demes}
+\section*{The \demes\ specification}
+
+\begin{itemize}
+    \item Demographic models are comprised of one or more interacting
+        populations, along with description, doi, time units and generation
+        times.
+    \item Populations (or demes) are treated as discrete collection of
+        exchangeable individuals following a well-defined set of rules
+    \item These rules include population sizes, times of existence, and other
+        information about the mating system (e.g. selfing or cloning rates)
+    \item Relationships between demes are defined by ancestors/descendents
+        and possible continuous or instantaneous migration between coexisting
+        demes
+\end{itemize}
 
@@
 \begin{itemize}
-\item Short description of YAML language and why it's appropriate here
-\item General overview (1 paragraph) of defining demography in YAML
-\item A generic, commonly used, simple demographic model for illustration (e.g.
-  an IM in Model~\ref{code:im_model})
+    \item YAML allows us to write this information in a human-readable and
+        structured manner (lists, keys/items, etc)
+    \item Implicit values, so that models can be written compactly
+    \item Example of an IM Model~\ref{code:im_model}
 \end{itemize}
@@
 
-\subsection*{Python representation of demography as input for simulation software}
+\subsection*{Unambiguous low-level representation}
 
 \begin{itemize}
-\item The Python API, reading in a YAML, and unambiguous and manipulable representation
+    \item The Python API, which can read in a YAML file or be built using the
+        Builder
+    \item Demographic model validitation
+    \item Unambiguous, redundant information
+    \item Designed to be programatically read by simulation software backends
 \end{itemize}
 
-\subsubsection*{Software support for \demes}
+\section*{Extensions and future directions}
+
+\subsection*{Software support for \demes}
 \aprcomment{at time of publication. TBD...}
 \begin{itemize}
-\item \msprime
-\item \stdpopsim
-\item \fwdpy
-\item \moments
-\item \dadi? should try to get Ryan on board
-\item \slim? I guess that comes with \stdpopsim
+    \item \msprime: coalescent simulation \citep{kelleher2016efficient}
+    \item \stdpopsim: a collection of maintained species and demographic models,
+        which includes a large number of models specified using \demes
+        \citep{adrion2020community}
+    \item \fwdpy: forward-time Wright-Fisher simulation \citep{thornton2014c++}
+    \item \moments: compute expected diversity statistics (SFS, LD) and perform
+        demographic inference using summary statistics
+        \citep{jouganous2017inferring,ragsdale2019models}
+    \item \dadi: compute the expected SFS and perform demographic inference using
+        summary statistics \citep{gutenkunst2009inferring}
 \end{itemize}
 
-\subsection*{Specifying complex demography}
+\subsection*{Other stuff...}
 
-\aprcomment{A few examples of complex demography - e.g. 1) the very common out-of-Africa
-model, 2) Denisovan admixture with a ton of populations and demographic events, 3) a
-non-human model, perhaps with selfing rates and rate changes}
-
-\aprcomment{Perhaps examples should be in the appendix, to keep the paper concise}
-
-\section*{Prospects and future directions}
+\begin{itemize}
+    \item \texttt{demesdraw} (\url{https://github.com/grahamgower/demesdraw.git})
+    \item General framework for inference (have been working on that over in \moments,
+        and perhaps it could be generalized to work with any simulation engine
+    \item Defining distributions of parameter values, perhaps?
+\end{itemize}
 
@@
 
+\aprcomment{Perhaps and example of a many-population model, and API to simulate using
+\msprime, \fwdpy, and \moments?}
+
 \end{document}

### 2021-06-11 paper.tex
Add GADMA and rough pass at the software table.

@@
 \newcommand{\fwdpy}[0]{\texttt{fwdpy11}}
-\newcommand{\slim}[0]{\texttt{slim}}
+\newcommand{\slim}[0]{\texttt{SLiM}}
+\newcommand{\gadma}[0]{\texttt{GADMA}}
 
 \newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
+\newcommand{\jkcomment}[1]{{\textcolor{red}{JK: #1}}}
 
@@
 \author[ ]{...Others here...}
+\author[**]{Jerome Kelleher}
 \author[**]{Kevin Thornton}
-\author[**]{Jerome Kelleher}
 \affil[*,**]{Denote equal contribution}
@@
 
+\begin{table}
+\begin{center}
+\begin{tabular}{llr}
+\msprime & \citep{kelleher2016efficient} & Coalescent simulator \\
+\stdpopsim & \citep{adrion2020community} & Library of standardised simulations \\
+\fwdpy & \citep{thornton2014c++} & Forward-time simulator \\
+\moments & \citep{jouganous2017inferring,ragsdale2019models}
+    & Demographic inference \\
+\dadi & \citep{gutenkunst2009inferring}  & Demographic inference \\
+\gadma & \citep{noskova2020gadma}  & Demographic inference \\
+\end{tabular}
+\end{center}
+\caption{Software supporting Demes as an input or output format.
+\jkcomment{This is a very rough first pass. It's not clear that a table is the
+right format for presenting this information, maybe a description
+list would be better instead?}}
+\end{table}
+
 \section*{Extensions and future directions}

### 2021-06-14 paper.tex
Reworked software as a description list.

--- 2021-04-30 paper.tex
@@
 
-\section*{Extensions and future directions}
+% \begin{table}
+% \begin{center}
+% \begin{tabular}{llr}
+% \msprime & \citep{kelleher2016efficient} & Coalescent simulator \\
+% \stdpopsim & \citep{adrion2020community} & Library of standardised simulations \\
+% \fwdpy & \citep{thornton2014c++} & Forward-time simulator \\
+% \moments & \citep{jouganous2017inferring,ragsdale2019models}
+%     & Demographic inference \\
+% \dadi & \citep{gutenkunst2009inferring}  & Demographic inference \\
+% \gadma & \citep{noskova2020gadma}  & Demographic inference \\
+% \end{tabular}
+% \end{center}
+% \caption{Software supporting Demes as an input or output format.
+% \jkcomment{This is a very rough first pass. It's not clear that a table is the
+% right format for presenting this information, maybe a description
+% list would be better instead?}}
+% \end{table}
 
-\subsection*{Software support for \demes}
-\aprcomment{at time of publication. TBD...}
-\begin{itemize}
-    \item \msprime: coalescent simulation \citep{kelleher2016efficient}
-    \item \stdpopsim: a collection of maintained species and demographic models,
-        which includes a large number of models specified using \demes
-        \citep{adrion2020community}
-    \item \fwdpy: forward-time Wright-Fisher simulation \citep{thornton2014c++}
-    \item \moments: compute expected diversity statistics (SFS, LD) and perform
+\subsection*{Software support for Demes}
+
+\begin{description}
+    \item[demes-python] A Python library for loading, saving, and working with
+        Demes models. Includes conversion utilities for existing formats
+        demographic model descriptions such as
+        \texttt{ms}~\citep{hudson2002generating}.
+    \item[demes-c] A C library for parsing Demes YAML descriptions.
+    \item[demesdraw] A Python library for visualising Demes models
+     (\url{https://github.com/grahamgower/demesdraw.git}).
+\end{description}
+
+\subsection*{Software implementing Demes}
+At the time of writing, the following software implements the Demes
+specification, as a supported input or output format.
+
+\begin{description}
+    \item[msprime:] Coalescent simulation
+        \citep{kelleher2016efficient,kelleher2020coalescent}
+    \item[stdpopsim:] A collection of community empirical data for species
+        and demographic models specified in Demes format~\citep{adrion2020community}.
+    \item[fwdpy11:] forward-time Wright-Fisher simulation \citep{thornton2014c++}
+    \item[moments:] compute expected diversity statistics (SFS, LD) and perform
         demographic inference using summary statistics
-        \citep{jouganous2017inferring,ragsdale2019models}
-    \item \dadi: compute the expected SFS and perform demographic inference using
-        summary statistics \citep{gutenkunst2009inferring}
-\end{itemize}
+        \citep{jouganous2017inferring,ragsdale2019models}.
+    \item[$\partial$a$\partial$i] compute the expected SFS and
+        perform demographic inference using
+        summary statistics \citep{gutenkunst2009inferring}.
+    \item[GADMA] Demographic inference \citep{noskova2020gadma}.
+\end{description}
 
@@
 \begin{itemize}
-    \item \texttt{demesdraw} (\url{https://github.com/grahamgower/demesdraw.git})
     \item General framework for inference (have been working on that over in \moments,

### 2021-06-14 paper.tex
Tidy up table.

--- 2021-04-30 paper.tex
@@
 \usepackage{graphicx}
+\usepackage{booktabs}
 
@@
 
-\section*{Extensions and future directions}
+% Add a bit more space between rows.
+\renewcommand{\arraystretch}{1.5}
+\begin{table}
+\begin{center}
+\begin{tabular}{lp{12cm}}
+\toprule
+\multicolumn{2}{l}{Software infrastructure}\\
+\midrule
+\texttt{demes-python} &
+    A Python library for loading, saving, and working with
+    Demes models. Includes conversion utilities for existing formats
+    demographic model descriptions such as
+    \texttt{ms}~\citep{hudson2002generating}.\\
 
-\subsection*{Software support for \demes}
-\aprcomment{at time of publication. TBD...}
-\begin{itemize}
-    \item \msprime: coalescent simulation \citep{kelleher2016efficient}
-    \item \stdpopsim: a collection of maintained species and demographic models,
-        which includes a large number of models specified using \demes
-        \citep{adrion2020community}
-    \item \fwdpy: forward-time Wright-Fisher simulation \citep{thornton2014c++}
-    \item \moments: compute expected diversity statistics (SFS, LD) and perform
+\texttt{demes-c} &
+    A C library for parsing Demes YAML descriptions.
+    (\url{https://github.com/grahamgower/demes-c.git}). \\
+
+\texttt{demesdraw} &
+    A Python library for visualising Demes models
+    (\url{https://github.com/grahamgower/demesdraw.git}). \\
+
+\midrule
+\multicolumn{2}{l}{Methods using Demes as input/output format}\\
+\midrule
+
+% TOOL AUTHORS: please add a row to this table to summarise what the software
+% does, and how it implements demes.
+
+\dadi &
+    Summary of what dadi does and how it implements Demes~\citep{gutenkunst2009inferring}.
+    \\
+
+\fwdpy &
+    Forward-time Wright-Fisher simulation \citep{thornton2014c++}\\
+
+\gadma &
+    Demographic inference \citep{noskova2020gadma}.\\
+
+\moments &
+    Compute expected diversity statistics (SFS, LD) and perform
     demographic inference using summary statistics
-        \citep{jouganous2017inferring,ragsdale2019models}
-    \item \dadi: compute the expected SFS and perform demographic inference using
-        summary statistics \citep{gutenkunst2009inferring}
-\end{itemize}
+    \citep{jouganous2017inferring,ragsdale2019models}.\\
+
+\msprime &
+    Coalescent simulation of Demes models, including command line
+    interface~\citep{kelleher2016efficient,kelleher2020coalescent}.\\
+
+\stdpopsim &
+    A collection of community empirical data for species
+    and demographic models specified in Demes
+    format~\citep{adrion2020community}.\\
+
+\bottomrule
+\end{tabular}
+\end{center}
+\caption{Software support for Demes. We have included software infrastructure developed
+for working with Demes models (parsing, validation, visualisation, etc)
+as well as all known downstream software that implements the specification,
+at the time of writing.}
+\end{table}
 
@@
 
-\section*{Appendix}
-
-\aprcomment{Perhaps and example of a many-population model, and API to simulate using
-\msprime, \fwdpy, and \moments?}
-
 \end{document}

### 2021-06-15 paper.tex
Apply suggestions from code review
Co-authored-by: Graham Gower <graham.gower@gmail.com>

@@
     A Python library for loading, saving, and working with
-    Demes models. Includes conversion utilities for existing formats
+    Demes models. Includes conversion utilities for existing
     demographic model descriptions such as
@@
 \stdpopsim &
-    A collection of community empirical data for species
+    A community-maintained collection of empirical data for species
     and demographic models specified in Demes

### 2021-06-15 paper.tex
First pass at introduction.

@@
 % local definitions
+\newcommand{\ms}[0]{\texttt{ms}}
 \newcommand{\msprime}[0]{\texttt{msprime}}
@@
 \newcommand{\Demes}[0]{\texttt{Demes}}
-\newcommand{\YAML}[0]{\texttt{YAML}}
 \newcommand{\moments}[0]{\texttt{moments}}
@@
 genome sequences
-\citep{thornton2014c++,kelleher2016efficient,haller2019slim,adrion2020community}
+\citep{thornton2014cpp,staab2015scrm,kelleher2016efficient,haller2019slim}
 or informative summary statistics of diversity
 \citep{gutenkunst2009inferring,kamm2017efficient,jouganous2017inferring} have
-enabled the increasing complexity of genomic studies, with many software
-packages able to handle large sample sizes, many interacting populations, and
+enabled the increasing complexity of genomic studies, with several software
+packages capable of working with large sample sizes, many interacting populations, and
 deviations from random-mating models of panmixia.
+This ability to infer and simulate such complex demographic scenarios, however,
+has highlighted a major shortcoming in community standards. The fragmented
+landscape of different approaches to describing demographic models makes
+it difficult to compare inferences made by different methods,
+and to reliably simulate from previously inferred models.
 
-The ability to simulate complex demographic scenarios, however, does not come
-without its own set of obstacles and frustrations. First, specifying models
-with many populations and parameters can be cumbersome and error-prone (e.g.,
-\citep{ragsdale2020lessons}). It is time-consuming and tedious to implement and
-then verify the correctness of such demographic models. Making matters worse,
-each software package typically has its own application programming interface
-(API), which makes it difficult to translate models between simulation
-software.
+Inference methods for demography usually output a set of
+text files which describe the inferred model,
+and nearly all methods use different file formats.
+% Thus, any comparison
+% between inferences requires these file formats to be parsed, with the user
+% being left with the task of unifying the data models.
+Inference results are typically reported in publications
+via a combination of visual depiction,
+a list of key parameters in tabular form and a discussion within the text.
+Unfortunately there is often ambiguity in these descriptions and
+implementing the precise model inferred for later simulation
+is at best tedious and error prone~\citep{ragsdale2020lessons}
+and occasionally impossible because of missing information.
+The \stdpopsim\ project is a community-maintained catalog of species
+information and demographic models designed to provide standardised
+population genetic simulations~\citep{adrion2020community}.
+The development process for adding demographic models to
+\stdpopsim\ tries to avoid errors in model implementation by requiring two
+independent implementations of each model
+to be equivalent. While this process is
+robust and has uncovered both significant
+errors~\citep{ragsdale2020lessons} and ambiguities in
+published models, it is labour intensive.
+It would be far better if published results
+were reported in an unambiguous standardised format that could be provided
+directly as input to simulators.
 
-Second, many software API are not particularly easy to read, especially when
-there are a large number of demographic events and parameters. This not only
-makes debugging difficult, but it can be difficult to even notice if a mistake
-has been made. A common interface for unambiguously specifying demographic
-models would reduce implementation errors and promote ease-of-use for many
-simulation software packages.
+Simulation is a core tool in population genetics, and a
+large number of different methods have been developed
+over the past three decades~\citep{carvajal2008simulation,liu2008survey,
+arenas2012simulation,yuan2012overview,hoban2012computer}.
+Simulations are based on highly idealised population models
+and one of the key uses of inferred demographic histories
+is to make simulations more realistic.
+Simulation methods take three broad approches to specifying
+the demographic model to simulate,
+using either a command line
+% JK: is there a simulator that just uses a CLI that's not using ms format?
+interface~\citep[e.g.][]{hudson2002generating},
+a custom input file format~\citep[e.g.][]{excoffier2011fastsimcoal,guillaume2006nemo},
+or an Application Programming Interface (API)~\citep[e.g.][]{
+thornton2014cpp,kelleher2016efficient,haller2019slim}.
+Command line interfaces are a concise way of expressing
+demographic models, and the syntax defined by \ms~\citep{hudson2002generating}
+is used by a number of different
+simulators~\citep[e.g.][]{ewing2010msms,staab2015scrm},
+and the POPDemog visualisation method~\citep{zhou2018popdemog}.
+However, this conciseness means that models of even intermediate complexity
+are difficult for humans to understand, making errors highly likely.
+Input parameter file formats for simulators allow the model specification
+to be less terse and (potentially) also allow for documentation in the
+form of comments. However, there is no standard format shared across
+simulators, and the details of the demographic models are not separated
+from other simulation parameters. The APIs use by individual simulators
+to define demographic models are also not appropriate as an interchange
+standard because of the dependency on both the programming language involved,
+and the specific details of the simulator in question.
 
-Finally, simulated or inferred demographic models are regularly shared in the
-literature, and for someone else to be able to reproduce or reimplement some
-demographic scenario, it needs to be described fully and without ambiguity.
-However, it is common for reported demographic models to lack clear or even
-complete descriptions, hindering reproducibility. A human-readable description
-for such models would facilitate their sharing and reimplementation.
+Here we present ``Demes'', a data model and file format
+specification for complex demographic models developed by the PopSim
+Consortium. The Demes data model precisely defines key parameters
+of populations and their relationships over time in an extensible way,
+and provides a way to explicitly encode all information relevant to
+demography while avoiding repetition. This data model is implemented
+in the widely used YAML format~\citep{ben2009yaml}, which provides a
+good balance between human and machine readability.
+The specification precisely
+defines the required behaviour of implementations, ensuring that there
+is no ambiguity of interpretation, and includes both a reference
+implementation and an extensive suite of test examples and their expected
+output. The initial software ecosystem includes high-quality
+Python and C parser implementations, as well as utilities for verification
+and visualisation of Demes models, and has been implemented in
+several popular inference and simulation methods
+(Table~\ref{tab-software}).
+We hope that this
+data model and file format will be widely adopted by the community,
+such that users can expect to simulate directly from
+inferred models with little or no programming effort.
 
-Here, we present \demes, a specification of complex demographic models that is
-both human readable and can be directly passed to a growing number of
-simulation software. \Demes\ supports model specification in a high-level
-\YAML\ format \citep{ben2009yaml} that is designed to be as easy as possible to
-implement, read, and share. Models are then represented in an unambiguous
-low-level format within Python that checks model validity. The initial Python
-release of \demes\ (version 0.1, described here) supports \emph{static}
-demographic descriptions, that is, models with fixed parameters. Future
-releases will support [[drawing parameters from distributions, for example for
-ABC-based inference, or describing ``dynamic`` models where parameters may be
-fit to data, e.g. \moments\ or \dadi]].
-
-\section*{The \demes\ specification}
+\section*{Demes}
 
@@
 
-\subsection*{High-level model specification in \YAML}
+\subsection*{High-level model specification in YAML}
 
@@
 \fwdpy &
-    Forward-time Wright-Fisher simulation \citep{thornton2014c++}\\
+    Forward-time Wright-Fisher simulation \citep{thornton2014cpp}\\
 
@@
 \end{center}
-\caption{Software support for Demes. We have included software infrastructure developed
+\caption{\label{tab-software}
+Software support for Demes. We have included software infrastructure developed
 for working with Demes models (parsing, validation, visualisation, etc)
@@
 
-\subsection*{Other stuff...}
+\section*{Discussion}
+
+This is an unstructured list of things we should mention:
 
 \begin{itemize}
-    \item General framework for inference (have been working on that over in \moments,
-        and perhaps it could be generalized to work with any simulation engine
-    \item Defining distributions of parameter values, perhaps?
+    \item Recent developments in software engineering have favoured
+    static, declarative configuration formats. It feels like a good idea to
+    have a Turing complete configuration format, but it turns out not to be.
+
+    \item Drawing parameters from distributions. Why we are static.
+    % https://github.com/popsim-consortium/demes-paper/issues/3
+
+    \item Why we've avoided turning this into a model of everything, allowing
+    for the specification of genome properties like recombination rates,
+    etc.
+
+    \item
+    % JK: I thought there was another simulation frontend somewhere..
+    Previous efforts such as \texttt{coala}~\citep{staab2016coala} have focused on
+    abstracting the differences between individual simulators by providing
+    a frontend.
+
+    \item
+    % https://github.com/popsim-consortium/demes-paper/issues/11
+    We should mention coasim~\citep{mailund2005coasim} because its model
+    seems quite close to what we have. It's not suitable as a standard though
+    because it uses a programming language.
+
 \end{itemize}

### 2021-06-15 paper.tex
Update author list and title.

@@
 
-\title{\demes: a pain-free specification for complex demography}
-\author[*]{Graham Gower}
-\author[*]{Aaron P. Ragsdale}
-\author[ ]{...Others here...}
-\author[**]{Jerome Kelleher}
-\author[**]{Kevin Thornton}
-\affil[*,**]{Denote equal contribution}
+\title{Demes: a specification for interchanging complex demography}
+% First authors
+\author[1,$\star$]{Graham Gower}
+\author[2,$\star$]{Aaron P. Ragsdale}
+
+% Authors: please add your name and affiliation here, in alphabetical order.
+% Second Authors
+\author[3]{Ryan N. Gutenkunst}
+\author[4]{Matthew Hartfield}
+\author[5]{Ekaterina Noskova}
+\author[3]{Travis J. Struck}
+%Senior Authors
+\author[6,$\dagger$,$\aleph$]{Jerome Kelleher}
+\author[7,$\dagger$]{Kevin R. Thornton}
+
+\affil[1]{Lundbeck GeoGenetics Centre, Globe Institute, University of Copenhagen}
+\affil[2]{FIXME: Aaron's affiliation}
+\affil[3]{Department of Molecular and Cellular Biology, University of Arizona}
+\affil[4]{Institute of Evolutionary Biology, The University of Edinburgh}
+\affil[5]{Computer Technologies Laboratory, ITMO University}
+\affil[6]{Big Data Institute, Li Ka Shing Centre for Health Information and
+    Discovery, University of Oxford}
+\affil[7]{Ecology and Evolutionary Biology, University of California at Irvine}
+
+\affil[$\star$]{Denotes shared first authorship, listed alphabetically}
+\affil[$\dagger$]{Denotes shared senior authorship, listed alphabetically}
+\affil[$\aleph$]{Denotes corresponding authors}
+
 \maketitle
 
-\abstract{
-}
+\abstract{}
 

### 2021-06-15 paper.tex

@@
 using either a command line
-% JK: is there a simulator that just uses a CLI that's not using ms format?
-interface~\citep[e.g.][]{hudson2002generating},
-a custom input file format~\citep[e.g.][]{excoffier2011fastsimcoal,guillaume2006nemo},
-or an Application Programming Interface (API)~\citep[e.g.][]{
-thornton2014cpp,kelleher2016efficient,haller2019slim}.
+interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
+a custom input file
+format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
+or an Application Programming Interface (API) to allow
+models be defined programmatically~\citep[e.g.][]{
+thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
 Command line interfaces are a concise way of expressing
@@
 is used by a number of different
-simulators~\citep[e.g.][]{ewing2010msms,staab2015scrm},
+simulators~\citep[e.g.][]{ewing2010msms,chen2009fast,staab2015scrm},
 and the POPDemog visualisation method~\citep{zhou2018popdemog}.

### 2021-06-16 paper.tex
First draft of abstract

@@
 \affil[$\dagger$]{Denotes shared senior authorship, listed alphabetically}
-\affil[$\aleph$]{Denotes corresponding authors}
+\affil[$\aleph$]{Denotes corresponding author}
 
@@
 
-\abstract{}
+\abstract{
+Understanding the demographic history of populations is a
+key goal in population genetics, and with improving methods
+and data, ever more complex models are being estimated.
+Demographic models of current interest (consisting of a set of discrete populations,
+their sizes and growth rates, and continuous and pulse migrations
+between those populations over a number of epochs) can require
+dozens of parameters to fully describe. There is currently
+no standard format to define such models, significantly
+hampering  progress in the field. In particular, the important
+task of translating the model descriptions in published
+work into input suitable for population genetic simulators is labour intensive
+and error prone.
+We propose the Demes data model and file format,
+built on widely used technologies,
+to alleviate these issues. Demes provides a well-defined and unambigous
+graphical model of populations and their properties that is straightforward to
+implement in software, and a text file format that is designed for
+simplicity and clarity.
+We provide thoroughly tested implementations of Demes parsers in Python
+and C, and showcase initial support in a number of simulators and inference
+methods.
+}
 

### 2021-06-17 paper.tex
Add notes on GUIs

@@
     seems quite close to what we have. It's not suitable as a standard though
-    because it uses a programming language.
+    because it uses a programming language. Besides coasim, there was
+    PopPlanner~\citep{ewing2010msms}
+    (a Java GUI which allows the user to build models
+    graphically and export the corresponding \ms\ command lines)
+    and MODELER4SIMCOAL2~\citep{antao2007modeler4simcoal2}
+    (a webapp that provided a set of common model templates that could be
+    parameterised to output the corresponding simcoal input files.)
+    \jkcomment{Not sure where these fit into the narrative, just
+    summarising here for later reference}.
 

### 2021-06-18 paper.tex
Discussion text.

--- 2021-06-16 paper.tex
@@
 
-\begin{code}
+\begin{figure}
+\begin{minipage}{0.48\textwidth}
 \begin{tcolorbox}
-\inputminted[linenos,numbersep=5pt]{yaml}{models/IM.yaml}
+\inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM.yaml}
 \end{tcolorbox}
+\begin{tcolorbox}
+\includegraphics[width=\textwidth]{fig/IM}
+\end{tcolorbox}
+\end{minipage}\hfill
+\begin{minipage}{0.48\textwidth}
+\begin{tcolorbox}
+\inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM-resolved.yaml}
+\end{tcolorbox}
+\end{minipage}\\
 \captionof{listing}{An IM model.}
 \label{code:im_model}
-\end{code}
+\end{figure}
 
@@
     \item Designed to be programatically read by simulation software backends
+\end{itemize}
+
+\subsection*{Design rationale}
+
+This is an unstructured list of things we should mention in terms
+of why certain decisions were made
+
+\begin{itemize}
+    \item Recent developments in software engineering have favoured
+    static, declarative configuration formats. It feels like a good idea to
+    have a Turing complete configuration format, but it turns out not to be.
+
+    \item Drawing parameters from distributions. Why we are static.
+    % https://github.com/popsim-consortium/demes-paper/issues/3
+
+    \item Why we've avoided turning this into a model of everything, allowing
+    for the specification of genome properties like recombination rates,
+    etc.
+
 \end{itemize}
@@
 
-This is an unstructured list of things we should mention:
+The difficulty of describing complex demographic models
+for population genetic simulations has long been acknowledged
+~\citep[][e.g.]{antao2007modeler4simcoal2}, % jk - any older refs for this?
+and a number of different methods have been proposed to
+make such simulations more accessible and less error prone.
+Several Graphical User Interface (GUI) and visualisation
+methods have been developed, which greatly facilitate
+interpretation~\citep{mailund2005coasim,antao2007modeler4simcoal2,
+ewing2010msms,zhou2018popdemog}. However, these methods
+currently have little traction as they are all either directly coupled
+to an internal simulation method or to the specific syntax
+of a given simulator. Another approach that has been taken to
+simplify running such simulations is to provide an API in a
+scripting language such as Python or R, providing a flexible
+interface either to a built-in simulation
+engine~\citep{thornton2014cpp,kelleher2016efficient,haller2017flexible}
+% jk: are/were there other simulation frontends out there?
+or as a frontend for other simulators~\citep{staab2016coala}.
+While APIs have many advantages and have grown in popularity in
+recent years, the demographic model descriptions are tied
+to the details of both the underlying simulation engines
+and the programming language in question.
 
-\begin{itemize}
-    \item Recent developments in software engineering have favoured
-    static, declarative configuration formats. It feels like a good idea to
-    have a Turing complete configuration format, but it turns out not to be.
+Stable and healthy software ecosystems require standard interchange
+formats, allowing for the development of high-quality and long-lasting
+tools that produce and consume the standard.
+Demographic models are a key part of population genetics research,
+and to date the transfer of inferred models to downstream simulations
+has been ad-hoc, and conversions between the many different ways
+of expressing such models both labour intensive and fraught with errors.
+The proposed Demes standard is an attempt to bridge this gap
+between inference and simulation, and also to provide the foundations
+for a sustainable ecosystem of tools built around this data model.
+Table~\ref{tab-software} shows some initial infrastructure that we have
+built as part of developing Demes, but many other useful tools
+can be envisaged that either produce or consume this format.
 
-    \item Drawing parameters from distributions. Why we are static.
-    % https://github.com/popsim-consortium/demes-paper/issues/3
-
-    \item Why we've avoided turning this into a model of everything, allowing
-    for the specification of genome properties like recombination rates,
-    etc.
-
-    \item
-    % JK: I thought there was another simulation frontend somewhere..
-    Previous efforts such as \texttt{coala}~\citep{staab2016coala} have focused on
-    abstracting the differences between individual simulators by providing
-    a frontend.
-
-    \item
-    % https://github.com/popsim-consortium/demes-paper/issues/11
-    We should mention coasim~\citep{mailund2005coasim} because its model
-    seems quite close to what we have. It's not suitable as a standard though
-    because it uses a programming language.
-
-\end{itemize}
+% Could finish up here by talking about reproducibility?
 

### 2021-06-18 paper.tex
Apply suggestions from code review
Co-authored-by: Graham Gower <graham.gower@gmail.com>

@@
 has been ad-hoc, and conversions between the many different ways
-of expressing such models both labour intensive and fraught with errors.
+of expressing such models is both labour intensive and fraught with errors.
 The proposed Demes standard is an attempt to bridge this gap
@@
 
-% Could finish up here by talking about reproducibility?
+A standard format also serves to improve scientific reproducibility.
+Authors simply include their demographic model as a supplementary file, or upload it to a data repository, so that readers need not transcribe the model themselves.
+This removes the opportunity for error due to model misinterpretation or incorrect translation to a simulatable format.
+In the event that the Demes standard becomes outdated, published Demes files will be convertible to future formats with relative ease because the data model is unambiguously specified.
 

### 2021-06-18 paper.tex
Update section on reproducibilty.

@@
 
+    \item Why we've versioned the specification, and how we've planned
+    for future extensions.
+
 \end{itemize}
@@
 
-A standard format also serves to improve scientific reproducibility.
-Authors simply include their demographic model as a supplementary file, or upload it to a data repository, so that readers need not transcribe the model themselves.
-This removes the opportunity for error due to model misinterpretation or incorrect translation to a simulatable format.
-In the event that the Demes standard becomes outdated, published Demes files will be convertible to future formats with relative ease because the data model is unambiguously specified.
+Reproducibility is a significant problem throughout the
+sciences~\citep{baker20161} and various measures have been
+proposed to increase the likelihood of researchers being
+able to replicate results in the
+literature~\citep{munafo2017manifesto}. The most basic requirement
+for reproducibility is that we must be able to state precisely what
+the result in question is. The lack of standardisation in how
+complex demographic models are communicated today, and the lack of
+precision in the published model descriptions means that it is difficult
+to replicate analyses, or reproduce those models for later simulation.
+Thus, we hope that the Demes standard introduced here will be widely adopted
+by inference methods and the results reported in publications
+routinely included as supplemental material or uploaded to a data
+repository in Demes format.
+
+\jkcomment{sketch of last paragraph}
+Another basic requirement for scientific progress in the inference
+of demographic models is that we should be able to compare the output
+of one method to the output of another, or to compare an inferred
+demography to some simulated ground truth. Demes facilitates this,
+paving the way for things like methods competitions. PopSim
+Consortium FTW!
 

### 2021-06-19 paper.tex
Removed final para sketch

@@
 
-\jkcomment{sketch of last paragraph}
-Another basic requirement for scientific progress in the inference
-of demographic models is that we should be able to compare the output
-of one method to the output of another, or to compare an inferred
-demography to some simulated ground truth. Demes facilitates this,
-paving the way for things like methods competitions. PopSim
-Consortium FTW!
-
 \bibliographystyle{plainnat}

### 2021-06-19 paper.tex
First pass at main body.

@@
 
-\begin{itemize}
-    \item Demographic models are comprised of one or more interacting
-        populations, along with description, doi, time units and generation
-        times.
-    \item Populations (or demes) are treated as discrete collection of
-        exchangeable individuals following a well-defined set of rules
-    \item These rules include population sizes, times of existence, and other
-        information about the mating system (e.g. selfing or cloning rates)
-    \item Relationships between demes are defined by ancestors/descendents
-        and possible continuous or instantaneous migration between coexisting
-        demes
-\end{itemize}
+The Demes specification is a formal data model for describing
+the properties of populations over time,
+along with some metadata and provenance information.
+The data model is based on the ubiquitous JSON~\citep{bray2017javascript}
+standard, and the structure formally defined using
+JSON Schema~\citep{wright2020json}.
+The Demes JSON schema rigorously defines the data model,
+specifying the hierarchical structure and the properties that are allowed in different
+contexts along with their types and permissable ranges. This schema, and its
+accompanying documentation, fully describe the entities in the model and their
+relationships, and the required behaviour of implementations. Since the
+schema is definitive, we will not recapitulate the details
+here, but instead focus on the high level properties of the model and
+the rationale behind key design decisions.
 
-\subsection*{High-level model specification in YAML}
+In Demes, demographic models consist of one or more interacting
+% JK: not sure at all about this, but let's see how it goes for now.
+populations (or ``demes"; to avoid confusion with the name of the
+specification itself we will use the term ``population'' in this
+discussion, in the understanding that the terms are interchangeable
+for our purposes). Populations are defined as a collection of
+exchangable individuals that exist for some period of time,
+following a well-defined set of rules regarding population
+sizes, mating systems, etc. Ancestor/descendent relationships
+between populations can be defined, as well as
+possible continuous or instantaneous migration between coexisting
+populations. The history of each population is defined as a
+series of ``epochs'' during which the population's parameters are
+fixed.
 
-\begin{itemize}
-    \item YAML allows us to write this information in a human-readable and
-        structured manner (lists, keys/items, etc)
-    \item Implicit values, so that models can be written compactly
-    \item Example of an IM Model~\ref{code:im_model}
-\end{itemize}
+The design of Demes is a balance between two partially competing requirements:
+that (a) models should be easy for humans to understand and manipulate; and (b)
+software processing Demes models should be provided with an unambiguous
+representation that is straightforward to process. For efficiency of
+understanding and avoidance of model specification error, we require a data
+representation without redundancy (i.e., repetition of values). But, for the
+simplicity of software working with the Demes model (and the avoidance of
+programming error, or divergence in interpretations of the specification)
+it is preferable to have an explicit representation, in
+which all relevant values are readily available. Thus, we consider
+three entities: the high-level, human readable, model specification;
+the low-level, fully-qualified model specification; and the parser,
+which is responsible for transforming the former into the latter.
 
@@
 \end{minipage}\\
-\captionof{listing}{An IM model.}
-\label{code:im_model}
+\caption{\label{fig-example}
+Example isolation-with-migration Demes model. (A) The compact non-redundant
+representation expressed using YAML. (B) The same model in the fully-qualified
+form. (C) A visual representation of the model using \texttt{demesdraw}.
+The YAML descriptions in (A) and (B) correspond exactly to a JSON description,
+but are much more concise.
+}
 \end{figure}
+
+\subsection*{High-level model specification}
+
+The high-level description of Demes models is focused on the efficiency of
+human understanding and the avoidance of errors. We have adopted the widely
+used YAML format~\citep{ben2009yaml} as the recommended means of interchanging
+Demes models (see Figure~\ref{fig-example} for exampels). YAML is a data
+serialisation language with an emphasis on simplicity, and interoperates well
+with JSON (indeed, YAML 1.2 is a superset of JSON). We chose YAML over JSON
+because although JSON is an excellent format for data interchange, it is
+ill-suited for human understanding or manipulation. We also considered other
+data formats such as TOML, but chose YAML because of its greater popularity and
+software support. Since the Demes data model is defined in JSON Schema,
+however, there is no formal dependency on YAML and implementations may choose
+to use JSON directly if they wish (e.g., for greater efficiency).
+
+Structurally, the high-level Demes specification encourages human understanding
+by avoiding redundancy in the description where possible, and
+by providing a mechanism for specifying default values that
+are inherited hierarchically. Figure~\ref{fig-example} shows an example
+Demes model expressed in both the high and low-level forms (both in
+terms of YAML syntax).
+
+\jkcomment{Finish up this section with a discussion of how we (a)
+use default values defined in JSON schema; (b) define more
+subtle implicit values in the specification itself; and (c) use the
+defaults sections, all explified by Fig 1.}
+
+\subsection*{Parsers}
+The implicit value inference and default value propagation required to
+fully resolve the high-level Demes model description is straightforward
+to describe, but still requires significant effort to implement in
+software. Moreover, there are many constraints on the data model
+(for example, epoch \texttt{end\_time} values should be strictly increasing
+within a deme), which must be enforced.
+It would not be reasonable to require every program that
+takes Demes models as input to implement this logic, as the programming
+effort would be significant (limiting adoption)
+and it is likely that if there were many implementations some would differ
+in their details (harming the software ecosystem). We therefore define
+a standard software entity as part of the specification (the parser),
+which performs this task, and which can be shared by programs that
+support demes as an input. By having relatively few (ideally, one
+per programming language), high-quality Demes parsers available as
+libraries the probability of divergences from the standard is
+greatly reduced.
+
+The Demes specification precisely defines the required behaviour of
+parsers, and we also provide a reference implementation written
+in Python to resolve any potential ambiguities and to provide a
+helpful template for other implementations. We also provide an
+extensive suite suite of examples and the expected outputs. In addition,
+we have high-quality parser implementations in the Python and C
+languages (published under liberal open source licences),
+providing a solid foundation for the software ecosystem.
 
@@
 
-\begin{itemize}
-    \item The Python API, which can read in a YAML file or be built using the
-        Builder
-    \item Demographic model validitation
-    \item Unambiguous, redundant information
-    \item Designed to be programatically read by simulation software backends
-\end{itemize}
+\jkcomment{Not sure if there's much point in this as a separate section any
+more? This could really be said in the previous section.}
+
+Parsers output the low-level description of Demes models,
+which is an explicit representation
+of the model where all parameters have been assigned values, and
+all data constraints have been checked. The output is formally defined
+as JSON, but in practise parsers output an object model that
+is suitable for the particular programming language, and
+that can be used directly by the implementing program.
 

### 2021-06-19 paper.tex
Added scope of spec section

@@
 
-\subsection*{Design rationale}
+\subsection*{Scope of the specification}
+\jkcomment{This section is very rough.}
 
-This is an unstructured list of things we should mention in terms
-of why certain decisions were made
+% https://github.com/popsim-consortium/demes-paper/issues/3
+
+The Demes specification is by design simple and static---we wish to
+unambiguously describe a demographic model with fully specified
+and fixed parameters. This simplicity means that we cannot directly
+specify distributions or estimated confidence intervals
+for those parameters. While it is not difficult to imagine extending
+the specification in ways that would allows us to do this, it would
+greatly increase the complexity of parsers and would likely be of
+little practical use in any case. This is because the parameters of
+demographic models are typically tightly coupled, and cases in which
+distributions for different parameters can be simply described are rare.
+In this situation, the simplest way to describe estimated
+distribution is simply to list a large number samples from
+the posterior. While writing out a large number of Demes models in
+YAML format may seem inefficient, it can in fact be a compact
+way to describe these distributions.
+\jkcomment{TODO: add in a quick example here where we take an IM model
+and simulate a posterior distribution under some reasonable model. Report
+how much space it takes to represent a large number of samples from this
+distribution}
+
+Similarly, one might be interested in running simulations in which
+the demographic model parameters are drawn from a distribution, e.g.,
+in ABC inference~\citep{beaumont2002approximate}. Allowing for
+the many distributions of interest would greatly increase the complexity
+parsers, with a relatively limited benefit. It would be far simpler
+and more powerful to write some code to sample directly from the
+distributions of interest, using, for example the Demes Python
+API.
+\jkcomment{This is garbled, but we get the point? You're probably never
+going to be doing ABC from a completely general Demes model as input.
+There will be some basic model that you have in mind, that you're generating
+distributions of parameters around. So, what's the point in writing
+general machinery to do this sampling, when you could write it yourself in
+10 lines of Python?}
 
 \begin{itemize}
-    \item Recent developments in software engineering have favoured
-    static, declarative configuration formats. It feels like a good idea to
-    have a Turing complete configuration format, but it turns out not to be.
-
-    \item Drawing parameters from distributions. Why we are static.
-    % https://github.com/popsim-consortium/demes-paper/issues/3
-
     \item Why we've avoided turning this into a model of everything, allowing

### 2021-06-21 paper.tex
Updates from review.

@@
 The YAML descriptions in (A) and (B) correspond exactly to a JSON description,
-but are much more concise.
+but are much more human-readable.
 }
@@
 used YAML format~\citep{ben2009yaml} as the recommended means of interchanging
-Demes models (see Figure~\ref{fig-example} for exampels). YAML is a data
+Demes models (see Figure~\ref{fig-example} for examples). YAML is a data
 serialisation language with an emphasis on simplicity, and interoperates well
@@
 ill-suited for human understanding or manipulation. We also considered other
-data formats such as TOML, but chose YAML because of its greater popularity and
-software support. Since the Demes data model is defined in JSON Schema,
+declarative data exchange formats such as TOML,
+but chose YAML because of equivalance with JSON,
+its popularity and good software support.
+Since the Demes data model is defined in JSON Schema,
 however, there is no formal dependency on YAML and implementations may choose
@@
 In this situation, the simplest way to describe estimated
-distribution is simply to list a large number samples from
+distribution is simply to list a large number of samples from
 the posterior. While writing out a large number of Demes models in
@@
 in ABC inference~\citep{beaumont2002approximate}. Allowing for
-the many distributions of interest would greatly increase the complexity
+the many distributions of interest would greatly increase the complexity of
 parsers, with a relatively limited benefit. It would be far simpler
@@
 10 lines of Python?}
+\jkcomment{Key point: if there was a concise way of writing down the joint
+distribution of all the parameters, we wouldn't be doing simulations in the
+first place.}
 

### 2021-06-24 paper.tex

@@
 \usepackage{booktabs}
+\usepackage{amsmath}
 
@@
 software. Moreover, there are many constraints on the data model
-(for example, epoch \texttt{end\_time} values should be strictly increasing
+(for example, epoch \texttt{end\_time} values should be strictly decreasing
 within a deme), which must be enforced.
@@
 helpful template for other implementations. We also provide an
-extensive suite suite of examples and the expected outputs. In addition,
+extensive test suite of examples and the expected outputs. In addition,
 we have high-quality parser implementations in the Python and C
@@
 way to describe these distributions.
-\jkcomment{TODO: add in a quick example here where we take an IM model
-and simulate a posterior distribution under some reasonable model. Report
-how much space it takes to represent a large number of samples from this
-distribution}
+For example, consider a one-population model with piecewise-constant sizes over
+20 epochs which has $\sim$~40 free parameters---the \texttt{start\_size} and
+\texttt{end\_time} values for each epoch. Supposing we sample 50,000 models
+from the posterior distribution, the resulting multi-document YAML file is
+45~MiB.
+This format compresses down to 8.4~MiB when gzipped or 6.2~MiB
+when compressed with LZMA2, which is on par with an equivalent binary
+representation of the free parameters
+($40\times50000\times4~\text{bytes} \approx 7.6~\text{MiB}$).
 

### 2021-06-24 paper.tex
start at pop gen model section

--- 2021-06-21 paper.tex
@@
 
-In Demes, demographic models consist of one or more interacting
-% JK: not sure at all about this, but let's see how it goes for now.
-populations (or ``demes"; to avoid confusion with the name of the
-specification itself we will use the term ``population'' in this
-discussion, in the understanding that the terms are interchangeable
-for our purposes). Populations are defined as a collection of
-exchangable individuals that exist for some period of time,
-following a well-defined set of rules regarding population
-sizes, mating systems, etc. Ancestor/descendent relationships
-between populations can be defined, as well as
-possible continuous or instantaneous migration between coexisting
-populations. The history of each population is defined as a
-series of ``epochs'' during which the population's parameters are
-fixed.
-
 The design of Demes is a balance between two partially competing requirements:
@@
 simplicity of software working with the Demes model (and the avoidance of
-programming error, or divergence in interpretations of the specification)
-it is preferable to have an explicit representation, in
-which all relevant values are readily available. Thus, we consider
-three entities: the high-level, human readable, model specification;
-the low-level, fully-qualified model specification; and the parser,
-which is responsible for transforming the former into the latter.
+programming error, or divergence in interpretations of the specification) it is
+preferable to have an explicit representation, in which all relevant values are
+readily available. Thus, we consider three entities: the high-level, human
+readable, model specification; the low-level, fully-qualified model
+specification; and the parser, which is responsible for transforming the former
+into the latter.
+
+\subsection*{Population genetics model}
+
+In Demes, demographic models consist of one or more interacting populations (or
+``demes"; to avoid confusion with the name of the specification itself we will
+use the term ``population'' in this discussion, in the understanding that the
+terms are interchangeable for our purposes). Populations are defined as a
+collection of exchangable individuals that exist a specified period of time,
+following a well-defined set of rules regarding population sizes, mating
+systems, etc. Ancestor/descendent relationships between populations can be
+defined, as well as continuous or instantaneous migration between coexisting
+populations.
+
+Population and event times are written as units in the past, so that time zero
+corresponds to the final generation or ``now'', and event times in the past are
+values greater than zero with larger values for events that occur in the more
+distant past. While a natural specification for time units is in generations,
+other time units are allowed, such as years. In the case the the time units are
+not given in generations, the generation time must also be specified so that
+times can be converted into generations. In general, as time flows from the
+past to the present, populations, epochs, and migration events should be
+specified in their order of appearance, so that their times are in descending
+order.
+
+Models include one or more populations that have sizes greater than zero for
+the duration of their existences, and population sizes are given as the number
+of individuals. A population must exist for some duration of time greater than
+zero, so that its start time must be larger than its end time. It may have one
+or more ancestors, which are other populations that exist at the population's
+start time. If one ancestor is specified, the first generation is constructed
+by randomly sampling parents from the ancestral population to contribute to
+offspring in the newly generated population. If more than one ancestor is
+specified, the proportions of ancestry from each contributing population must
+be provided, and those proportions must sum to one. In this case, parents are
+chosen randomly from each ancestral population with probability given by those
+proportions. If no ancestors are specified, the population is assumed to have
+start time equal to infinity, as individuals are not assumed to spontaneously
+spring into existence.
+
+Within populations, population sizes and mating systems are specified within
+epochs. Epochs are non-overlapping intervals that cover the entire existence
+interval of the population. Each epoch specifies the population size over that
+interval, which can be a constant value or function defined by start and end
+sizes that must remain positive.  If an epoch has a start time of infinity, the
+population size for that epoch must be constant. Epochs can also specify
+parameters for nonrandom mating, such as selfing or cloning rates, which give
+the probability that offspring are generated from one generation to the next by
+self-fertilization or cloning of an indivual \textbf{need to precisely specify
+once this is determined}.
+
+Finally, individuals in a population may have parents from a different
+population through migrations. These can be defined as continuous migration
+rates over intervals that populations overlap in time or through instantaneous
+(or pulse) migration events at a given time. Continuous migration rates are
+defined as the probability that parents in the ``destination`` population are
+chosen from the ``source`` population.  Migration rates are thus per-generation
+and must be less than one. Furthermore, if more than one source population have
+continuous migration into the same destination population, the sum of those
+migration rates must also be less than one, as rates define probabilities. The
+probability that parents come from the same population is just one minus the
+some of incoming migration rates. On the other hand, pulse migration events
+specify the instantaeous replacement of a given fraction of individuals in a
+destination population by individuals with parents from a source population.
+The fraction must be between zero and one, and if more than one source
+population has a pulse migration event to the same destination population and
+the same time, those replacement events are applied sequentially in the order
+that they are specified in the model.
 

### 2021-06-27 paper.tex
incorporating comments, fixing typos

@@
 terms are interchangeable for our purposes). Populations are defined as a
-collection of exchangable individuals that exist a specified period of time,
-following a well-defined set of rules regarding population sizes, mating
-systems, etc. Ancestor/descendent relationships between populations can be
+collection of exchangable individuals that exist for a specified period of
+time, following a well-defined set of rules regarding population sizes, mating
+systems, etc. Ancestor/descendant relationships between populations can be
 defined, as well as continuous or instantaneous migration between coexisting
@@
 distant past. While a natural specification for time units is in generations,
-other time units are allowed, such as years. In the case the the time units are
-not given in generations, the generation time must also be specified so that
-times can be converted into generations. In general, as time flows from the
-past to the present, populations, epochs, and migration events should be
+other time units are allowed, such as years. In the case that the time units
+are not given in generations, the generation time must also be specified so
+that times can be converted into generations. In general, as time flows from
+the past to the present, populations, epochs, and migration events should be
 specified in their order of appearance, so that their times are in descending
@@
 Within populations, population sizes and mating systems are specified within
-epochs. Epochs are non-overlapping intervals that cover the entire existence
-interval of the population. Each epoch specifies the population size over that
-interval, which can be a constant value or function defined by start and end
-sizes that must remain positive.  If an epoch has a start time of infinity, the
-population size for that epoch must be constant. Epochs can also specify
-parameters for nonrandom mating, such as selfing or cloning rates, which give
-the probability that offspring are generated from one generation to the next by
-self-fertilization or cloning of an indivual \textbf{need to precisely specify
-once this is determined}.
+epochs. Epochs are non-overlapping intervals of time that cover the entire
+existence interval of the population. Each epoch specifies the population size
+over that interval, which can be a constant value or function defined by start
+and end sizes that must remain positive.  If an epoch has a start time of
+infinity, the population size for that epoch must be constant. Epochs can also
+specify parameters for nonrandom mating, such as selfing or cloning rates,
+which give the probability that offspring are generated from one generation to
+the next by self-fertilization or cloning of an individual \textbf{need to
+precisely specify once this is determined}.
 
@@
 population through migrations. These can be defined as continuous migration
-rates over intervals that populations overlap in time or through instantaneous
-(or pulse) migration events at a given time. Continuous migration rates are
-defined as the probability that parents in the ``destination`` population are
-chosen from the ``source`` population.  Migration rates are thus per-generation
-and must be less than one. Furthermore, if more than one source population have
-continuous migration into the same destination population, the sum of those
-migration rates must also be less than one, as rates define probabilities. The
-probability that parents come from the same population is just one minus the
-some of incoming migration rates. On the other hand, pulse migration events
-specify the instantaeous replacement of a given fraction of individuals in a
-destination population by individuals with parents from a source population.
-The fraction must be between zero and one, and if more than one source
-population has a pulse migration event to the same destination population and
-the same time, those replacement events are applied sequentially in the order
-that they are specified in the model.
+rates over time intervals for which populations coexist or through
+instantaneous (or pulse) migration events at a given time. Continuous migration
+rates are defined as the probability that parents in the ``destination''
+population are chosen from the ``source'' population.  Migration rates are thus
+per generation and must be less than one. Furthermore, if more than one source
+population have continuous migration into the same destination population, the
+sum of those migration rates must also be less than one, as rates define
+probabilities. The probability that parents come from the same population is
+just one minus the sum of incoming migration rates. On the other hand, pulse
+migration events specify the instantaeous replacement of a given fraction of
+individuals in a destination population by individuals with parents from a
+source population.  The fraction must be between zero and one, and if more than
+one pulse occurs at the same time, those replacement events are applied
+sequentially in the order that they are specified in the model.
 

### 2021-06-29 paper.tex
Trim out spec-like details from popgen model.

@@
 terms are interchangeable for our purposes). Populations are defined as a
-collection of exchangable individuals that exist for a specified period of
+collection of exchangeable individuals that exist for a specified period of
 time, following a well-defined set of rules regarding population sizes, mating
@@
 distant past. While a natural specification for time units is in generations,
-other time units are allowed, such as years. In the case that the time units
-are not given in generations, the generation time must also be specified so
-that times can be converted into generations. In general, as time flows from
-the past to the present, populations, epochs, and migration events should be
-specified in their order of appearance, so that their times are in descending
-order.
+other time units are allowed, such as years.
+\jkcomment{Why did we choose time to flow in this direction?}
 
-Models include one or more populations that have sizes greater than zero for
-the duration of their existences, and population sizes are given as the number
-of individuals. A population must exist for some duration of time greater than
-zero, so that its start time must be larger than its end time. It may have one
+Population sizes are given as the number of individuals.
+\jkcomment{As opposed to genomes. Why did we do this?}
+Sizes and mating system details are specified for each population within
+epochs. Epochs are contiguous time intervals that define
+the existence interval of the population. Each epoch specifies the population size
+over that interval, which can be a constant value or function defined by start
+and end sizes that must remain positive.
+
+Within a population, individuals are assumed to be exchangeable, but
+there are also parameters for nonrandom mating, such as selfing or cloning rates,
+which give the probability that offspring are generated from one generation to
+the next by self-fertilization or cloning of an individual
+\textbf{need to precisely specify once this is determined}.
+\jkcomment{More detail here. Also, why only selfing and cloning, why not selection
+coefficients and whatnot?}
+
+A population may have one
 or more ancestors, which are other populations that exist at the population's
@@
 chosen randomly from each ancestral population with probability given by those
-proportions. If no ancestors are specified, the population is assumed to have
-start time equal to infinity, as individuals are not assumed to spontaneously
-spring into existence.
-
-Within populations, population sizes and mating systems are specified within
-epochs. Epochs are non-overlapping intervals of time that cover the entire
-existence interval of the population. Each epoch specifies the population size
-over that interval, which can be a constant value or function defined by start
-and end sizes that must remain positive.  If an epoch has a start time of
-infinity, the population size for that epoch must be constant. Epochs can also
-specify parameters for nonrandom mating, such as selfing or cloning rates,
-which give the probability that offspring are generated from one generation to
-the next by self-fertilization or cloning of an individual \textbf{need to
-precisely specify once this is determined}.
+proportions.
 
@@
 rates are defined as the probability that parents in the ``destination''
-population are chosen from the ``source'' population.  Migration rates are thus
-per generation and must be less than one. Furthermore, if more than one source
-population have continuous migration into the same destination population, the
-sum of those migration rates must also be less than one, as rates define
-probabilities. The probability that parents come from the same population is
-just one minus the sum of incoming migration rates. On the other hand, pulse
-migration events specify the instantaeous replacement of a given fraction of
+population are chosen from the ``source'' population.
+On the other hand, pulse
+migration events specify the instantaneous replacement of a given fraction of
 individuals in a destination population by individuals with parents from a
-source population.  The fraction must be between zero and one, and if more than
-one pulse occurs at the same time, those replacement events are applied
-sequentially in the order that they are specified in the model.
+source population.
 

### 2021-07-28 paper.tex
edits and filling in bullets

@@
 \affil[1]{Lundbeck GeoGenetics Centre, Globe Institute, University of Copenhagen}
-\affil[2]{FIXME: Aaron's affiliation}
+\affil[2]{Department of Integrative Biology, University of Wisconsin--Madison}
 \affil[3]{Department of Molecular and Cellular Biology, University of Arizona}
@@
 a list of key parameters in tabular form and a discussion within the text.
-Unfortunately there is often ambiguity in these descriptions and
+Unfortunately there is often ambiguity in these descriptions, and
 implementing the precise model inferred for later simulation
@@
 simulators, and the details of the demographic models are not separated
-from other simulation parameters. The APIs use by individual simulators
+from other simulation parameters. The APIs used by individual simulators
 to define demographic models are also not appropriate as an interchange
@@
 specification for complex demographic models developed by the PopSim
-Consortium. The Demes data model precisely defines key parameters
+Consortium. \aprcomment{do we need ``developed by the PopSim Consortium''?}
+The Demes data model precisely defines key parameters
 of populations and their relationships over time in an extensible way,
@@
 accompanying documentation, fully describe the entities in the model and their
-relationships, and the required behaviour of implementations. Since the
+relationships and the required behaviour of implementations. Since the
 schema is definitive, we will not recapitulate the details
@@
 readily available. Thus, we consider three entities: the high-level, human
-readable, model specification; the low-level, fully-qualified model
+readable model specification; the low-level, fully-qualified model
 specification; and the parser, which is responsible for transforming the former
@@
 In Demes, demographic models consist of one or more interacting populations (or
-``demes"; to avoid confusion with the name of the specification itself we will
+``demes''; to avoid confusion with the name of the specification itself we will
 use the term ``population'' in this discussion, in the understanding that the
@@
 time, following a well-defined set of rules regarding population sizes, mating
-systems, etc. Ancestor/descendant relationships between populations can be
-defined, as well as continuous or instantaneous migration between coexisting
+systems, and so on. Ancestor/descendant relationships between populations can
+be defined, as well as continuous or instantaneous migration between coexisting
 populations.
@@
 values greater than zero with larger values for events that occur in the more
-distant past. While a natural specification for time units is in generations,
-other time units are allowed, such as years.
-\jkcomment{Why did we choose time to flow in this direction?}
+distant past. By having time units increase into the past, we avoid the need to
+choose an arbitrary point in history as ``time zero'', so that instead zero
+refers to ``now'' or the final generation of a simulation. A natural
+specification for time units is in generations, although other time units are
+permitted, such as years, accompanied by the generation time.
 
@@
 \jkcomment{As opposed to genomes. Why did we do this?}
+\aprcomment{With this, we might want to be able to specify ploidy in as
+a graph-level option, which defaults to 2. That would make the reasons
+behind this decision obvious then, I think.}
 Sizes and mating system details are specified for each population within
@@
 
-Within a population, individuals are assumed to be exchangeable, but
-there are also parameters for nonrandom mating, such as selfing or cloning rates,
-which give the probability that offspring are generated from one generation to
-the next by self-fertilization or cloning of an individual
-\textbf{need to precisely specify once this is determined}.
+Within a population, individuals are assumed to be exchangeable, but there are
+also parameters for nonrandom mating, such as selfing or cloning rates, which
+give the probability that offspring are generated from one generation to the
+next by self-fertilization or cloning of an individual. Cloning and selfing
+rates can take any value between zero and one, and if both are specified then
+the selfing rate is the probability an offspring is generated through
+self-fertilization given it was produced sexually (i.e. with probability one
+minus the cloning rate). We omit any details regarding selection, as those would
+fall outside the scope of demographic model specification, but...
+\aprcomment{If selection is placed on top of the demographic model, this implies
+that we are living in a soft selection world. Though sizes could just as easily
+be interpreted as carrying capacities in a hard selection model.}
+
 \jkcomment{More detail here. Also, why only selfing and cloning, why not selection
@@
 \begin{figure}
-\begin{minipage}{0.48\textwidth}
+\begin{minipage}{0.44\textwidth}
 \begin{tcolorbox}
@@
 \end{minipage}\hfill
-\begin{minipage}{0.48\textwidth}
+\begin{minipage}{0.54\textwidth}
 \begin{tcolorbox}
@@
 Structurally, the high-level Demes specification encourages human understanding
-by avoiding redundancy in the description where possible, and
-by providing a mechanism for specifying default values that
-are inherited hierarchically. Figure~\ref{fig-example} shows an example
-Demes model expressed in both the high and low-level forms (both in
-terms of YAML syntax).
+by avoiding redundancy in the description where possible and by providing a
+mechanism for specifying default values that are inherited hierarchically.
+Figure~\ref{fig-example} shows an example Demes model expressed in both the
+high and low-level forms (both in terms of YAML syntax). For values that repeat
+across fields, default may be used to implicity assign a default value to
+fields of the given type.  A default is superceded by an explicitly provided
+value if given. Other values are inherited naturally following the progression
+of time. For example, if an epoch \texttt{start\_size} is not provided, it is
+assumed to be equal to the \texttt{end\_size} of the previous epoch. This also
+means that the first epoch of each population must specify the initial size.
 
-\jkcomment{Finish up this section with a discussion of how we (a)
-use default values defined in JSON schema; (b) define more
-subtle implicit values in the specification itself; and (c) use the
-defaults sections, all explified by Fig 1.}
+\aprcomment{not sure if this is sufficient here}
 
@@
 which performs this task, and which can be shared by programs that
-support demes as an input. By having relatively few (ideally, one
-per programming language), high-quality Demes parsers available as
-libraries the probability of divergences from the standard is
-greatly reduced.
+support demes as an input. By having relatively few high-quality Demes
+parsers available as libraries (ideally, one per programming language),
+the probability of divergences from the standard is greatly reduced.
 
-The Demes specification precisely defines the required behaviour of
-parsers, and we also provide a reference implementation written
-in Python to resolve any potential ambiguities and to provide a
-helpful template for other implementations. We also provide an
-extensive test suite of examples and the expected outputs. In addition,
-we have high-quality parser implementations in the Python and C
-languages (published under liberal open source licences),
+The Demes specification precisely defines the required behaviour of parsers,
+which translate the high-level specifications into unambiguous low-level
+representations where all parameters have been assigned values and data
+constraints have been checked. The output is formally defined as JSON, but in
+practise parsers output an object model that is suitable for the particular
+programming language and that can be used directly by the implementing program.
+We provide a reference implementation written in Python to resolve any
+potential ambiguities and to provide a helpful template for other
+implementations, as well as an extensive test suite of examples and the
+expected outputs. In addition, we have high-quality parser implementations in
+the Python and C languages (published under liberal open source licences),
 providing a solid foundation for the software ecosystem.
 
-\subsection*{Unambiguous low-level representation}
-
-\jkcomment{Not sure if there's much point in this as a separate section any
-more? This could really be said in the previous section.}
-
-Parsers output the low-level description of Demes models,
-which is an explicit representation
-of the model where all parameters have been assigned values, and
-all data constraints have been checked. The output is formally defined
-as JSON, but in practise parsers output an object model that
-is suitable for the particular programming language, and
-that can be used directly by the implementing program.
+\aprcomment{wrapped below section about parsers to here}
 
@@
 
-\begin{itemize}
-    \item Why we've avoided turning this into a model of everything, allowing
-    for the specification of genome properties like recombination rates,
-    etc.
+Demes is meant to interface with a large number of simulation and inference
+methods, so we have restricted to the specification to only describe
+demographic features at the population level. As the underlying discrete
+demographic structure is the common thread connecting many widely used
+population genetic software, many features of underlying genome biology are
+omitted.  Most notably, mutation and recombination rates are left unspecified
+(though they can be encoded as metadata in the data structure). Genome ploidy
+is left ambiguous, and selection and dominance models are absent. Using a Demes
+model, which specifies deterministic population sizes, as the demographic input
+for simulation using software that allows for selection implies that selection
+is \emph{soft}, that is, genome fitnesses are rescaled relative to the mean
+population fitness \citep{christiansen1975hard}. It is possible to instead
+interpret population sizes as carrying capacities, allowing for a model of
+\emph{hard} selection. However, this should not be understood as the default
+interpretation of population sizes, which conventionally specifies a
+deterministic number of individuals.  Finally, as Demes assumes the
+exchangeability of individuals within populations, Demes does not provide a way
+to define geographic spatial structure within populations.
 
-    \item Why we've versioned the specification, and how we've planned
-    for future extensions.
-
-\end{itemize}
+Demes version 1.0 is intended to provide a stable and minimal specification
+that is compatible with a wide range of simulation softwares. Beginning with
+version 1.0, a Demes-specified model is expected to provide the version of
+Demes for which it is valid, and if unprovided, it should be a valid model
+under the default version, 1.0. Future extensions of the Demes specification
+will be released with versions following semantic versioning guidelines [cite
+or url?], and downstream software that incorporate Demes are expected to
+indicate and check Demes version compatibility.
 
@@
 Thus, we hope that the Demes standard introduced here will be widely adopted
-by inference methods and the results reported in publications
-routinely included as supplemental material or uploaded to a data
+by simulation and inference methods and be used for reporting results in
+publications, either as supplemental material or uploaded to a data
 repository in Demes format.

### 2021-07-29 paper.tex
addressing comments

@@
 
-Here we present ``Demes'', a data model and file format
-specification for complex demographic models developed by the PopSim
-Consortium. \aprcomment{do we need ``developed by the PopSim Consortium''?}
-The Demes data model precisely defines key parameters
-of populations and their relationships over time in an extensible way,
-and provides a way to explicitly encode all information relevant to
-demography while avoiding repetition. This data model is implemented
-in the widely used YAML format~\citep{ben2009yaml}, which provides a
-good balance between human and machine readability.
-The specification precisely
-defines the required behaviour of implementations, ensuring that there
-is no ambiguity of interpretation, and includes both a reference
-implementation and an extensive suite of test examples and their expected
-output. The initial software ecosystem includes high-quality
-Python and C parser implementations, as well as utilities for verification
-and visualisation of Demes models, and has been implemented in
-several popular inference and simulation methods
-(Table~\ref{tab-software}).
-We hope that this
-data model and file format will be widely adopted by the community,
-such that users can expect to simulate directly from
-inferred models with little or no programming effort.
+Here we present ``Demes'', a data model and file format specification for
+complex demographic models developed by the PopSim Consortium. The Demes data
+model precisely defines key parameters of populations and their relationships
+over time in an extensible way, and provides a way to explicitly encode all
+information relevant to demography while avoiding repetition. This data model
+is implemented in the widely used YAML format~\citep{ben2009yaml}, which
+provides a good balance between human and machine readability. The
+specification precisely defines the required behaviour of implementations,
+ensuring that there is no ambiguity of interpretation, and includes both a
+reference implementation and an extensive suite of test examples and their
+expected output. The initial software ecosystem includes high-quality Python
+and C parser implementations, as well as utilities for verification and
+visualisation of Demes models, and has been implemented in several popular
+inference and simulation methods (Table~\ref{tab-software}).  We hope that this
+data model and file format will be widely adopted by the community, such that
+users can expect to simulate directly from inferred models with little or no
+programming effort.
 
@@
 high and low-level forms (both in terms of YAML syntax). For values that repeat
-across fields, default may be used to implicity assign a default value to
+across fields, defaults may be used to implicity assign default values to
 fields of the given type.  A default is superceded by an explicitly provided
-value if given. Other values are inherited naturally following the progression
-of time. For example, if an epoch \texttt{start\_size} is not provided, it is
-assumed to be equal to the \texttt{end\_size} of the previous epoch. This also
-means that the first epoch of each population must specify the initial size.
+value if given. Other implicit values are inherited naturally following the
+progression of time. For example, if an epoch \texttt{start\_size} is not
+provided, it is assumed to be equal to the \texttt{end\_size} of the previous
+epoch. This also means that the first epoch of each population must specify the
+initial size.
 
@@
 constraints have been checked. The output is formally defined as JSON, but in
-practise parsers output an object model that is suitable for the particular
+practice parsers output an object model that is suitable for the particular
 programming language and that can be used directly by the implementing program.
@@
 
-Demes is meant to interface with a large number of simulation and inference
+Demes is designed to interface with a large number of simulation and inference
 methods, so we have restricted to the specification to only describe
@@
 demographic structure is the common thread connecting many widely used
-population genetic software, many features of underlying genome biology are
+population genetic methods, many features of underlying genome biology are
 omitted. Most notably, mutation and recombination rates are left unspecified
@@
 interpretation of population sizes, which conventionally specifies a
-deterministic number of individuals.  Finally, as Demes assumes the
-exchangeability of individuals within populations, Demes does not provide a way
-to define geographic spatial structure within populations.
+deterministic number of individuals. Finally, as Demes assumes exchangeability
+of individuals within populations, it not provide a way to define geographic
+spatial structure within populations.
 
 Demes version 1.0 is intended to provide a stable and minimal specification
-that is compatible with a wide range of simulation softwares. Beginning with
-version 1.0, a Demes-specified model is expected to provide the version of
-Demes for which it is valid, and if unprovided, it should be a valid model
-under the default version, 1.0. Future extensions of the Demes specification
-will be released with versions following semantic versioning guidelines [cite
-or url?], and downstream software that incorporate Demes are expected to
-indicate and check Demes version compatibility.
+that is compatible with a wide range of simulators. To ensure consistency
+across implementations, the specification is versioned and we commit to
+following semantic versioning guidelines.
 

### 2021-08-06 paper.tex

@@
 \newcommand{\jkcomment}[1]{{\textcolor{red}{JK: #1}}}
+\newcommand{\ggcomment}[1]{{\textcolor{yellow!60!red}{GG: #1}}}
 
@@
 built on widely used technologies,
-to alleviate these issues. Demes provides a well-defined and unambigous
+to alleviate these issues. Demes provides a well-defined and unambiguous
 graphical model of populations and their properties that is straightforward to
@@
 is to make simulations more realistic.
-Simulation methods take three broad approches to specifying
+Simulation methods take three broad approaches to specifying
 the demographic model to simulate,
@@
 simulators~\citep[e.g.][]{ewing2010msms,chen2009fast,staab2015scrm},
-and the POPDemog visualisation method~\citep{zhou2018popdemog}.
+and the POPDemog visualisation software~\citep{zhou2018popdemog}.
 However, this conciseness means that models of even intermediate complexity
@@
 specifying the hierarchical structure and the properties that are allowed in different
-contexts along with their types and permissable ranges. This schema, and its
+contexts along with their types and permissible ranges. This schema, and its
 accompanying documentation, fully describe the entities in the model and their
@@
 
-In Demes, demographic models consist of one or more interacting populations (or
-``demes''; to avoid confusion with the name of the specification itself we will
-use the term ``population'' in this discussion, in the understanding that the
-terms are interchangeable for our purposes). Populations are defined as a
+In Demes, demographic models consist of one or more interacting populations,
+or ``demes'', by which we mean a grouping of individuals that is convenient
+for modelling \citep{gilmour_demes_1939,gilmour_deme_1955}.
+To avoid confusion with the name of the specification itself we will
+use the term ``population'' in this discussion, with the understanding that the
+terms are interchangeable for our purposes. Populations are defined as a
 collection of exchangeable individuals that exist for a specified period of
@@
 give the probability that offspring are generated from one generation to the
-next by self-fertilization or cloning of an individual. Cloning and selfing
+next by self-fertilisation or cloning of an individual. Cloning and selfing
 rates can take any value between zero and one, and if both are specified then
 the selfing rate is the probability an offspring is generated through
-self-fertilization given it was produced sexually (i.e. with probability one
+self-fertilisation given it was produced sexually (i.e. with probability one
 minus the cloning rate). We omit any details regarding selection, as those would
@@
 declarative data exchange formats such as TOML,
-but chose YAML because of equivalance with JSON,
+but chose YAML because of equivalence with JSON,
 its popularity and good software support.
@@
 high and low-level forms (both in terms of YAML syntax). For values that repeat
-across fields, defaults may be used to implicity assign default values to
-fields of the given type.  A default is superceded by an explicitly provided
+across fields, defaults may be used to implicitly assign default values to
+fields of the given type.  A default is superseded by an explicitly provided
 value if given. Other implicit values are inherited naturally following the
@@
 and it is likely that if there were many implementations some would differ
-in their details (harming the software ecosystem). We therefore define
+in their details (harming the software ecosystem).
+% GG: this is currently the case for programs accepting ms command lines. E.g.
+%  - MaCS interprets -es proportion as 1-proportion compared to ms.
+%  - SCRM uses different options for -ema command.
+We therefore define
 a standard software entity as part of the specification (the parser),
@@
 \subsection*{Scope of the specification}
+\subsubsection*{Static models, not parameterised models}
 \jkcomment{This section is very rough.}
@@
 
-The Demes specification is by design simple and static---we wish to
-unambiguously describe a demographic model with fully specified
-and fixed parameters. This simplicity means that we cannot directly
-specify distributions or estimated confidence intervals
+The Demes specification is static by design---we wish to
+unambiguously describe a demographic model with a concrete set
+of parameters. This simplicity means that we cannot directly
+specify parameter distributions or estimated confidence intervals
 for those parameters. While it is not difficult to imagine extending
-the specification in ways that would allows us to do this, it would
-greatly increase the complexity of parsers and would likely be of
-little practical use in any case. This is because the parameters of
+the specification in ways that would allow this, it's not immediately
+clear that the benefits are worth the increased parser complexity.
+
+\ggcomment{
+Possible reasons to advocate for non-static features are as follows.
+\begin{enumerate}
+\item Specifying prior distributions of parameters for an ABC.
+\item Specifying upper and lower bounds of parameters and inter-parameter
+      constraints for input to an optimisation-based inference.
+\item Specifying joint distribution, or confidence intervals, for the output of
+      an inference procedure.
+\item Simulating from some arbitrary joint distribution that was inferred
+      previously. E.g. using an ABC posterior distribution of models to
+      train a neural network.
+\item Extending someone else's static model to reinfer parameters,
+      or add new parameters.
+\end{enumerate}
+}
+
+The parameters of
 demographic models are typically tightly coupled, and cases in which
 distributions for different parameters can be simply described are rare.
-In this situation, the simplest way to describe estimated
-distribution is simply to list a large number of samples from
+In this situation, the simplest way to describe an estimated
+distribution is to list a large number of samples from
 the posterior. While writing out a large number of Demes models in
@@
 the demographic model parameters are drawn from a distribution, e.g.,
-in ABC inference~\citep{beaumont2002approximate}. Allowing for
-the many distributions of interest would greatly increase the complexity of
-parsers, with a relatively limited benefit. It would be far simpler
-and more powerful to write some code to sample directly from the
-distributions of interest, using, for example the Demes Python
-API.
-\jkcomment{This is garbled, but we get the point? You're probably never
-going to be doing ABC from a completely general Demes model as input.
-There will be some basic model that you have in mind, that you're generating
-distributions of parameters around. So, what's the point in writing
-general machinery to do this sampling, when you could write it yourself in
-10 lines of Python?}
-\jkcomment{Key point: if there was a concise way of writing down the joint
-distribution of all the parameters, we wouldn't be doing simulations in the
-first place.}
+in ABC inference~\citep{beaumont2002approximate}.
+Other inference procedures based on optimising a loss function
+\citep{gutenkunst2009inferring,kamm2017efficient,jouganous2017inferring}
+(\ggcomment{more refs? MCMC? neural networks?})
+need users to specify parameter bounds,
+and possibly non-linear constraints between parameters.
+Indeed, the choice of how to parameterise a model could be important for
+some inference methods (e.g. absolute times versus relative times between events).
 
+Implementing the many distributions of interest, and supporting a general
+way to describe a model's free parameters, would greatly increase the
+complexity of parsers, with relatively limited benefit to most users.
+It's unlikely that Demes could be made sufficiently
+flexible without considerable effort to implement many features of
+general-purpose programming languages, such as variables, arithmetic,
+and flow control.
+Such use cases are therefore better served by writing model-generating
+functions in an existing programming language, for example
+using the Demes Python API.
+To modify an existing static Demes model, perhaps to infer new parameters,
+this would unfortunately require reimplementing the model.
+However, there exist many templating solutions for YAML and JSON that are
+specifically designed for extending static data in arbitrarily complex
+(\ggcomment{refs: YTT, Jsonnet, CUE, Dhall}),
+offering intriguing possibilities.
+
+\subsubsection*{Population-level features, not genome features}
 Demes is designed to interface with a large number of simulation and inference
@@
 deterministic number of individuals. Finally, as Demes assumes exchangeability
-of individuals within populations, it not provide a way to define geographic
+of individuals within populations, it doesn't provide a way to define geographic
 spatial structure within populations.
@@
     demographic model descriptions such as
-    \texttt{ms}~\citep{hudson2002generating}.\\
+    \texttt{ms}~\citep{hudson2002generating}
+    (\url{https://github.com/popsim-consortium/demes-python}).\\
 
@@
     A C library for parsing Demes YAML descriptions.
-    (\url{https://github.com/grahamgower/demes-c.git}). \\
+    (\url{https://github.com/grahamgower/demes-c}). \\
 
@@
     A Python library for visualising Demes models
-    (\url{https://github.com/grahamgower/demesdraw.git}). \\
+    (\url{https://github.com/grahamgower/demesdraw}). \\
 
@@
 % jk: are/were there other simulation frontends out there?
-or as a frontend for other simulators~\citep{staab2016coala}.
+or as a front end for other simulators~\citep{staab2016coala}.
 While APIs have many advantages and have grown in popularity in
@@
 built as part of developing Demes, but many other useful tools
-can be envisaged that either produce or consume this format.
+can be envisaged that consume, transform, or produce this format.
 

### 2021-08-20 paper.tex
Tweak

@@
 \newcommand{\ggcomment}[1]{{\textcolor{yellow!60!red}{GG: #1}}}
+\newcommand{\krtcomment}[1]{{\textcolor{purple}{KRT: #1}}}
 
@@
 genome sequences
-\citep{thornton2014cpp,staab2015scrm,kelleher2016efficient,haller2019slim}
+\citep{thornton2014cpp,thornton2019-nu,staab2015scrm,baumdicker2021-iu,kelleher2016efficient,haller2019slim}
 or informative summary statistics of diversity
@@
 models be defined programmatically~\citep[e.g.][]{
-thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
+thornton2014cpp,thornton2019-nu,baumdicker2021-iu,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
 Command line interfaces are a concise way of expressing
@@
 complex demographic models developed by the PopSim Consortium. The Demes data
-model precisely defines key parameters of populations and their relationships
-over time in an extensible way, and provides a way to explicitly encode all
+model precisely defines the sizes of populations and the relationships
+between populations in terms of migrations over time,
+and provides a way to explicitly encode the
 information relevant to demography while avoiding repetition. This data model
@@
 use the term ``population'' in this discussion, with the understanding that the
-terms are interchangeable for our purposes. Populations are defined as a
-collection of exchangeable individuals that exist for a specified period of
-time, following a well-defined set of rules regarding population sizes, mating
-systems, and so on. Ancestor/descendant relationships between populations can
-be defined, as well as continuous or instantaneous migration between coexisting
-populations.
+terms are interchangeable for our purposes.
+A population is defined as some collection of individuals that exists for
+some period of time, and has a well-defined size (i.e., number of individuals)
+during that time period. Individuals can move between populations
+either through ancestor-descendant relationships between the populations,
+or through processes of continuous or pulse migration.
+No other properties of the populations are specified in the model:
+we are concerned only with defining the populations, their sizes, and the
+movement of individuals between those populations.
 
@@
 
-Population sizes are given as the number of individuals.
-\jkcomment{As opposed to genomes. Why did we do this?}
-\aprcomment{With this, we might want to be able to specify ploidy in as
-a graph-level option, which defaults to 2. That would make the reasons
-behind this decision obvious then, I think.}
+Population sizes are given as the number of individuals, and details
+such as ploidy levels are considered external to the model.
+We focus on the number of individuals as opposed to the number of genomes
 Sizes and mating system details are specified for each population within
-epochs. Epochs are contiguous time intervals that define
+epochs.
+Epochs are contiguous time intervals that define
 the existence interval of the population. Each epoch specifies the population size
@@
 self-fertilisation given it was produced sexually (i.e. with probability one
-minus the cloning rate). We omit any details regarding selection, as those would
-fall outside the scope of demographic model specification, but...
+minus the cloning rate). We omit any details regarding selection.
+Thus, the history of population size changes does not depend on mean fitnesses within
+a population nor on differences in mean fitness between populations.
+This indepedence of population size and
+selection is often called ``soft" selection \citep{christiansen1975hard}.
+
+List of thoughts to turn into a paragraph:
+
+\begin{enumerate}
+    \item Selection parameters may be arbitrarily complex.  Different fitness effects in different demes.
+          Variation over time.  Etc..
+          The specification would become unwieldy and perhaps limiting.
+    \item By allowing selfing and cloning, we allow many "standard" models to be considered with "no or little"
+          extra effort.
+    \item However, by parameterising selfing and cloning as we have, we are assuming that these properties of
+          populations at given times can be specified independently from the genetics.
+          In other words, mutations that cause selfing to "come and go" are not considered.
+          However, consuming software is free to ignore these parameters and model the genetics within
+          the framework of a soft selection model.
+    \item We don't view these as limitations. Rather, this is what most people are doing most of the time.
+          Just now we can finally try to do this part w/fewer errors.
+\end{enumerate}
+
 \aprcomment{If selection is placed on top of the demographic model, this implies
@@
 be interpreted as carrying capacities in a hard selection model.}
+\krtcomment{We are firmly in the soft selection world here.  The connection to hard selection
+requires some other strong assumptions that will sometimes be met, and sometimes not.
+For example, if the mean fitnesses are approximately equal across demes, then a hard selection
+model could be similar to a soft selection model.  Relevant literature to dig up: Felsenstein's review,
+which will lead to other stuff.}
 
@@
 coefficients and whatnot?}
+\krtcomment{Selection is a can of worms: scaling of diploid genotype fitnesses will differ between tools.
+The possibility of "GxE" ("different s for same genotype in different demes") can lead to quadratic complexity.}
 
@@
 \end{figure}
+
+This model is, by design, very limited. We do not specify any details of
+individuals or of the processes that happen \emph{within} populations,
+because such details are necessarily complicated, vary greatly by
+application, and any attempt to encompass such a broad range of biological
+processes is doomed to failure. An ``individual'' (i.e., a discrete
+organism) is an abstraction that should have a well defined interpretation
+in most cases.
+This narrow focus on individuals and their groupings into populations
+should ensure that the Demes standard is stable, and not subject to
+continual development as more and more elaborations of basic processes
+are added. We could never hope to fully encompass all the details
+of every possible simulation (including ploidy,
+genome sizes, recombination, mutation and gene conversion rates,
+selection coefficients, and so on) and so we clearly demarcate the
+% FIXME Not quite true here, we're including selfing and cloning rate
+role of the specification by including \emph{none} of them.
+Such details can be specified by existing mechanisms used by
+simulators. If they vary according to the demography (for example,
+we have different recombination maps across populations), then
+we can unambiguously define these by referring to the population
+identifiers and times defined by the Demes model. Rather than attempting
+to embed this information \emph{in} the demographic model, we
+can define it externally and refer \emph{to} the demography.
 
@@
 clear that the benefits are worth the increased parser complexity.
+
+\krtcomment{Aaron has done a bit of this non-static stuff in moments.  It may be useful to look over there?}
 
@@
 \fwdpy &
-    Forward-time Wright-Fisher simulation \citep{thornton2014cpp}\\
+    Forward-time Wright-Fisher simulation \citep{thornton2014cpp,thornton2019-nu}\\
 
@@
     Coalescent simulation of Demes models, including command line
-    interface~\citep{kelleher2016efficient,kelleher2020coalescent}.\\
+    interface~\citep{kelleher2016efficient,kelleher2020coalescent,baumdicker2021-iu}.\\
 
@@
 interface either to a built-in simulation
-engine~\citep{thornton2014cpp,kelleher2016efficient,haller2017flexible}
+engine~\citep{thornton2014cpp,thornton2019-nu,baumdicker2021-iu,kelleher2016efficient,haller2017flexible}
 % jk: are/were there other simulation frontends out there?

### 2022-01-18 drawfig.py
Update drawfig.py to new demesdraw API
Progress blocked by grahamgower/demesdraw#83

@@ -12,8 +12,8 @@ def size_max(graph):
 graph = demes.load("models/IM.yaml")
 w = 0.8 * size_max(graph)
 positions = dict(A=0, X=-w, Y=w)
-ax = demesdraw.utils.get_axes(aspect=1)
-ax = demesdraw.tubes(graph, ax=ax, positions=positions, seed=1, inf_ratio=0.4)
+fig, ax = demesdraw.utils.get_fig_axes(aspect=1)
+ax = demesdraw.tubes(graph) #, ax=ax, positions=positions, seed=1, inf_ratio=0.4)
 ax.figure.savefig(
     "fig/IM.pdf",
     # Save with a transparent background.

### 2022-01-18 paper.tex
Minor from KRT and some citation updates.

--- 2021-08-06 paper.tex
@@
 behind this decision obvious then, I think.}
+\krtcomment{The number of genomes can be a non-obvious function of the number of individuals, too.
+Polymorphism for chromosome number, etc., are all real complications.}
 Sizes and mating system details are specified for each population within

### 2022-01-19 drawfig.py
Minor from KRT and some citation updates. (#31)

@@ -13,7 +13,7 @@ graph = demes.load("models/IM.yaml")
 w = 0.8 * size_max(graph)
 positions = dict(A=0, X=-w, Y=w)
 fig, ax = demesdraw.utils.get_fig_axes(aspect=1)
-ax = demesdraw.tubes(graph) #, ax=ax, positions=positions, seed=1, inf_ratio=0.4)
+ax = demesdraw.tubes(graph, ax=ax, positions=positions, seed=1, inf_ratio=0.4)
 ax.figure.savefig(
     "fig/IM.pdf",
     # Save with a transparent background.

### 2022-01-27 paper.tex
Various edits (#32)

@@
 Population sizes are given as the number of individuals.
+We focus on the number of individuals as opposed to the number of genomes
+in order to not limit other software (need term -- "consumers"?) by placing potentially arbitrary
+restrictions on how the numbers of chromosomes, etc., vary amongst individuals.
 \jkcomment{As opposed to genomes. Why did we do this?}
@@
 Polymorphism for chromosome number, etc., are all real complications.}
+\krtcomment{We support limited aspects of the mating system--more on this below.}
 Sizes and mating system details are specified for each population within
@@
 self-fertilisation given it was produced sexually (i.e. with probability one
-minus the cloning rate). We omit any details regarding selection, as those would
-fall outside the scope of demographic model specification, but...
+minus the cloning rate). We omit any details regarding selection.
+Thus, the history of population size changes does not depend on mean fitnesses within
+a population nor on differences in mean fitness between populations.
+This indepedence of population size and selection is often called ``soft" selection \citep{christiansen1975hard}.
+
+List of thoughts to turn into a paragraph:
+
+\begin{enumerate}
+    \item Selection parameters may be arbitrarily complex.  Different fitness effects in different demes.
+          Variation over time.  Etc..
+          The specification would become unwieldy and perhaps limiting.
+    \item By allowing selfing and cloning, we allow many "standard" models to be considered with "no or little"
+          extra effort.
+    \item However, by parameterising selfing and cloning as we have, we are assuming that these properties of
+          populations at given times can be specified independently from the genetics.
+          In other words, mutations that cause selfing to "come and go" are not considered.
+          However, consuming software is free to ignore these parameters and model the genetics within
+          the framework of a soft selection model.
+    \item We don't view these as limitations. Rather, this is what most people are doing most of the time.
+          Just now we can finally try to do this part w/fewer errors.
+\end{enumerate}
+
 \aprcomment{If selection is placed on top of the demographic model, this implies

### 2022-02-22 paper.tex
Controversial text addition

--- 2021-08-20 paper.tex
@@
 
-Population sizes are given as the number of individuals, and details
-such as ploidy levels are considered external to the model.
+Population sizes are given as the number of individuals.
 We focus on the number of individuals as opposed to the number of genomes
+in order to not limit other software (need term -- "consumers"?) by placing potentially arbitrary
+restrictions on how the numbers of chromosomes, etc., vary amongst individuals.
+\jkcomment{As opposed to genomes. Why did we do this?}
+\aprcomment{With this, we might want to be able to specify ploidy in as
+a graph-level option, which defaults to 2. That would make the reasons
+behind this decision obvious then, I think.}
+\krtcomment{The number of genomes can be a non-obvious function of the number of individuals, too.
+Polymorphism for chromosome number, etc., are all real complications.}
+\krtcomment{We support limited aspects of the mating system--more on this below.}
 Sizes and mating system details are specified for each population within
-epochs.
+epochs. Epochs are contiguous time intervals that define
+
 Epochs are contiguous time intervals that define
@@
 a population nor on differences in mean fitness between populations.
-This indepedence of population size and
-selection is often called ``soft" selection \citep{christiansen1975hard}.
+This indepedence of population size and selection is often called ``soft" selection \citep{christiansen1975hard}.
 
@@
 organism) is an abstraction that should have a well defined interpretation
-in most cases.
+in most cases, despite the endless creativity of biological reality.
 This narrow focus on individuals and their groupings into populations
@@
 are added. We could never hope to fully encompass all the details
-of every possible simulation (including ploidy,
+of every possible simulation (including mating system, ploidy,
 genome sizes, recombination, mutation and gene conversion rates,
 selection coefficients, and so on) and so we clearly demarcate the
-% FIXME Not quite true here, we're including selfing and cloning rate
 role of the specification by including \emph{none} of them.
@@
 identifiers and times defined by the Demes model. Rather than attempting
-to embed this information \emph{in} the demographic model, we
+to \emph{embed} this information into the demographic model, we
 can define it externally and refer \emph{to} the demography.

### 2022-03-07 paper.tex
Add text re: the assumed model being Wright-Fisher (#33)

@@
 
-Population sizes are given as the number of individuals.
+Population sizes are given as the number of individuals, and details
+such as ploidy levels are considered external to the model.
 We focus on the number of individuals as opposed to the number of genomes
-in order to not limit other software (need term -- "consumers"?) by placing potentially arbitrary
-restrictions on how the numbers of chromosomes, etc., vary amongst individuals.
-\jkcomment{As opposed to genomes. Why did we do this?}
-\aprcomment{With this, we might want to be able to specify ploidy in as
-a graph-level option, which defaults to 2. That would make the reasons
-behind this decision obvious then, I think.}
-\krtcomment{The number of genomes can be a non-obvious function of the number of individuals, too.
-Polymorphism for chromosome number, etc., are all real complications.}
-\krtcomment{We support limited aspects of the mating system--more on this below.}
 Sizes and mating system details are specified for each population within
-epochs. Epochs are contiguous time intervals that define
-
+epochs.
 Epochs are contiguous time intervals that define
@@
 and end sizes that must remain positive.
+
+\krtcomment{The paragraph(s) below subsume(s)/replace(s) some of what appears in later pargraphs.
+	I'm not making an attempt yet to revise that later content--I think other authors should
+	weigh in.}
+
+\krtcomment{Aaron -- I have perhaps over-simplified the description of marginal allele frequency dynamics?}
+
+Within a population, we implicitly assume that allele frequency dynamics can
+be described by the Wright-Fisher model.  Briefly, generations are non-overlapping (all parents
+reproduce and die simultaneously) and the expected frequency (at birth) of the $i^{th}$ allele currently at frequency $p_i$
+one generation in the next generation is $p_iw_i/\bar{w}$, where $w_i$ and $\bar{w}$ are the marginal and
+mean fitnesses, respectively, properly weighted according to ancestry proportions. In this framework,
+a forward-time simulation of finite populations is equivalent to multinomial sampling of allele frequencies each
+generation (\citet[][pp 29-31]{burger2000-ul}, \citet[][pp 179-181]{crowkimura1970})
+and a backwards-time (coalescent) simulation follows the approximations described in \cite[][chapter 3]{wakeley2008-hd}.
+Further, this model assumes "soft" selection \citep{christiansen1975hard}, meaning that the dynamics of population sizes
+changes are \textit{independent} of the details of individual fitnesses.
+In the \texttt{Discussion}, we discuss alternative models compatible with this framework and outline the expectations for tools implementing
+them.
+A plethora of forwards and backwards time simulators currently implement this model (REFS).
+Each has their own unique method of specifying the details of selection and demography.
+The goal of \Demes\ is to provide a common method for the latter task.
+
+\krtcomment{We need to revisit the paragraph(s) below.  We do not assume individuals are exchangeable, because we do not
+	require that all individuals have the same fitness. See, e.g., p 60 of Wakeley for why fitness == exchangeability
+	and why that is required for the standard coalescent.}
 
@@
 a population nor on differences in mean fitness between populations.
-This indepedence of population size and selection is often called ``soft" selection \citep{christiansen1975hard}.
+This indepedence of population size and
+selection is often called ``soft" selection \citep{christiansen1975hard}.
 
@@
 organism) is an abstraction that should have a well defined interpretation
-in most cases, despite the endless creativity of biological reality.
+in most cases.
 This narrow focus on individuals and their groupings into populations
@@
 are added. We could never hope to fully encompass all the details
-of every possible simulation (including mating system, ploidy,
+of every possible simulation (including ploidy,
 genome sizes, recombination, mutation and gene conversion rates,
 selection coefficients, and so on) and so we clearly demarcate the
+% FIXME Not quite true here, we're including selfing and cloning rate
 role of the specification by including \emph{none} of them.
@@
 identifiers and times defined by the Demes model. Rather than attempting
-to \emph{embed} this information into the demographic model, we
+to embed this information \emph{in} the demographic model, we
 can define it externally and refer \emph{to} the demography.
@@
 
+\krtcomment{Here, we should say something about other models that people may wish to implement within the demes
+framework.}
+
 \bibliographystyle{plainnat}

### 2022-03-15 paper.tex
begin the overhaul

@@
 
+\usepackage{filecontents,catchfile}
+\begin{filecontents*}{software-table.tex}
+\end{filecontents*}
+
 \begin{document}
@@
 
-\abstract{
+\begin{abstract}
 Understanding the demographic history of populations is a
@@
 and data, ever more complex models are being estimated.
-Demographic models of current interest (consisting of a set of discrete populations,
+Demographic models of current interest
+typically consist of a set of discrete populations,
 their sizes and growth rates, and continuous and pulse migrations
-between those populations over a number of epochs) can require
+between those populations over a number of epochs, which can require
 dozens of parameters to fully describe. There is currently
@@
 methods.
-}
+\end{abstract}
 
@@
 geographically diverse species and populations has allowed us to infer complex
-demography and study life history at fine scales. An integral component to such
-population genetics studies is simulation. Software to either simulate whole
-genome sequences
-\citep{thornton2014cpp,thornton2019-nu,staab2015scrm,baumdicker2021-iu,kelleher2016efficient,haller2019slim}
-or informative summary statistics of diversity
-\citep{gutenkunst2009inferring,kamm2017efficient,jouganous2017inferring} have
-enabled the increasing complexity of genomic studies, with several software
-packages capable of working with large sample sizes, many interacting populations, and
-deviations from random-mating models of panmixia.
+demography and study life history at fine scales.
+An integral component to such population genetics studies is simulation.
+Software to either simulate whole genome
+sequences~\citep{thornton2014cpp,thornton2019-nu,staab2015scrm,baumdicker2021-iu,kelleher2016efficient,haller2019slim}
+or informative summary statistics of
+diversity~\citep{gutenkunst2009inferring,kamm2017efficient,jouganous2017inferring}
+have enabled the increasing complexity of genomic studies, with several software
+packages capable handling large sample sizes, many interacting populations, and
+deviations from panmictic random-mating assumptions.
 This ability to infer and simulate such complex demographic scenarios, however,
-has highlighted a major shortcoming in community standards. The fragmented
-landscape of different approaches to describing demographic models makes
-it difficult to compare inferences made by different methods,
+has highlighted a major shortcoming in community standards.
+The fragmented landscape of different approaches to describing demographic
+models makes it difficult to compare inferences made by different methods
 and to reliably simulate from previously inferred models.
@@
 and nearly all methods use different file formats.
-% Thus, any comparison
-% between inferences requires these file formats to be parsed, with the user
-% being left with the task of unifying the data models.
 Inference results are typically reported in publications
@@
 a list of key parameters in tabular form and a discussion within the text.
-Unfortunately there is often ambiguity in these descriptions, and
+Unfortunately these descriptions are often ambiguous, and
 implementing the precise model inferred for later simulation
@@
 and occasionally impossible because of missing information.
-The \stdpopsim\ project is a community-maintained catalog of species
-information and demographic models designed to provide standardised
-population genetic simulations~\citep{adrion2020community}.
-The development process for adding demographic models to
-\stdpopsim\ tries to avoid errors in model implementation by requiring two
-independent implementations of each model
-to be equivalent. While this process is
-robust and has uncovered both significant
-errors~\citep{ragsdale2020lessons} and ambiguities in
-published models, it is labour intensive.
-It would be far better if published results
-were reported in an unambiguous standardised format that could be provided
+%\aprcomment{This feels tangential - maybe discussion}
+%The \stdpopsim\ project is a community-maintained catalog of species
+%information and demographic models designed to provide standardised
+%population genetic simulations~\citep{adrion2020community}.
+%The development process for adding demographic models to
+%\stdpopsim\ tries to avoid errors in model implementation by requiring two
+%independent implementations of each model
+%to be equivalent. While this process is
+%robust and has uncovered both significant
+%errors~\citep{ragsdale2020lessons} and ambiguities in
+%published models, it is labour intensive.
+Instead, it would be far better if published results were reported
+in an unambiguous standardised format that could be provided
 directly as input to simulators.
@@
 Simulation is a core tool in population genetics, and a
-large number of different methods have been developed
-over the past three decades~\citep{carvajal2008simulation,liu2008survey,
+many methods have been developed over the past three
+decades~\citep{carvajal2008simulation,liu2008survey,
 arenas2012simulation,yuan2012overview,hoban2012computer}.
-Simulations are based on highly idealised population models
+Simulations are based on highly idealised population models,
 and one of the key uses of inferred demographic histories
@@
 Input parameter file formats for simulators allow the model specification
-to be less terse and (potentially) also allow for documentation in the
-form of comments. However, there is no standard format shared across
-simulators, and the details of the demographic models are not separated
-from other simulation parameters. The APIs used by individual simulators
-to define demographic models are also not appropriate as an interchange
-standard because of the dependency on both the programming language involved,
+to be less terse and allow for documentation in the form of comments.
+However, there is no standard format shared across simulators,
+and the details of the demographic models are not separated
+from other simulation parameters.
+The APIs used by individual simulators to define demographic models are also
+not appropriate as an interchange standard because of the dependency on both
+the programming language involved,
 and the specific details of the simulator in question.
@@
 Here we present ``Demes'', a data model and file format specification for
-complex demographic models developed by the PopSim Consortium. The Demes data
-model precisely defines the sizes of populations and the relationships
-between populations in terms of migrations over time,
-and provides a way to explicitly encode the
-information relevant to demography while avoiding repetition. This data model
-is implemented in the widely used YAML format~\citep{ben2009yaml}, which
-provides a good balance between human and machine readability. The
-specification precisely defines the required behaviour of implementations,
+complex demographic models developed by the
+PopSim Consortium~\citep{adrion2020community}.
+The Demes data model precisely defines the sizes and relationships
+of populations, and it provides a way to explicitly encode the
+information relevant to demography while avoiding repetition.
+This data model is implemented in the widely used YAML
+format~\citep{ben2009yaml}, which provides a good balance between human
+and machine readability.
+The specification precisely defines the required behaviour of implementations,
 ensuring that there is no ambiguity of interpretation, and includes both a
 reference implementation and an extensive suite of test examples and their
-expected output. The initial software ecosystem includes high-quality Python
+expected output.
+The initial software ecosystem includes high-quality Python
 and C parser implementations, as well as utilities for verification and
 visualisation of Demes models, and has been implemented in several popular
-inference and simulation methods (Table~\ref{tab-software}).  We hope that this
-data model and file format will be widely adopted by the community, such that
-users can expect to simulate directly from inferred models with little or no
-programming effort.
+inference and simulation methods (Table~\ref{tab-software}).
+We hope that this data model and file format will be widely adopted
+by the community, such that users can expect to simulate directly
+from inferred models with little or no programming effort.
 
 \section*{Demes}
+
+The design of Demes is a balance between two partially competing requirements:
+that (a) models should be easy for humans to understand and manipulate;
+and (b) software processing Demes models should be provided with an unambiguous
+representation that is straightforward to process.
+For efficiency of understanding and avoidance of model specification error,
+we require a data representation without redundancy (i.e., repetition of values).
+However, for the simplicity of software working with the Demes model
+(and the avoidance of programming error, or divergence in
+interpretations of the specification) it is preferable to have an explicit
+representation, in which all relevant values are readily available.
+Thus, Demes is composed of three entities: the high-level, human-readable
+model specification; the low-level, fully-qualified model
+specification; and the parser, which is responsible for transforming the former
+into the latter.
+
+Below we provide a brief overview of the population genetics models
+that Demes supports and the components of the Demes infrastructure.
+We then highlight an example usage, including visualization and simulation
+of a multi-population demographic model.
+Detailed descriptions of the population genetic model, parsers, scope, and
+additional examples are given in the Appendix.
+
+\CatchFileDef{\softwaretable}{software-table.tex}{}
+\renewcommand{\arraystretch}{1.5}
+\begin{table}
+\begin{center}
+\begin{tabular}{lp{12cm}}
+\softwaretable
+\end{tabular}
+\end{center}
+\caption{\label{tab-software}
+Software support for Demes. We have included software infrastructure developed
+for working with Demes models (parsing, validation, visualisation, etc)
+as well as all known downstream software that implements the specification,
+at the time of writing.}
+\end{table}
+
+\subsection*{Population genetic model}
+1 paragraph
+
+\subsection*{High level specification}
+1 paragraph
+
+\subsection*{Parsers}
+and languages. 1 paragraph
+
+\subsection*{The scope of Demes}
+
+\begin{itemize}
+    \item Static, instead of parameterized models (1 paragraph)
+    \item Population level instead of genomic features (1 paragraph)
+\end{itemize}
+
+\section*{Illustration: simulation using Demes}
+
+\begin{itemize}
+    \item Multi-population model with migrations and/or pulses
+    \item Demesdraw
+    \item msprime simulation
+    \item moments simulation
+    \item Compare SFS in each population
+    \item The input YAML file should go in the Appendix
+\end{itemize}
+
+\section*{Discussion}
+
+The difficulty of describing complex demographic models
+for population genetic simulations has long been acknowledged
+~\citep[][e.g.]{antao2007modeler4simcoal2}, % jk - any older refs for this?
+and a number of different methods have been proposed to
+make such simulations more accessible and less error prone.
+Several Graphical User Interface (GUI) and visualisation
+methods have been developed, which greatly facilitate
+interpretation~\citep{mailund2005coasim,antao2007modeler4simcoal2,
+ewing2010msms,zhou2018popdemog}. However, these methods
+currently have little traction as they are all either directly coupled
+to an internal simulation method or to the specific syntax
+of a given simulator. Another approach that has been taken to
+simplify running such simulations is to provide an API in a
+scripting language such as Python or R, providing a flexible
+interface either to a built-in simulation
+engine~\citep{thornton2014cpp,thornton2019-nu,baumdicker2021-iu,kelleher2016efficient,haller2017flexible}
+% jk: are/were there other simulation frontends out there?
+or as a front end for other simulators~\citep{staab2016coala}.
+While APIs have many advantages and have grown in popularity in
+recent years, the demographic model descriptions are tied
+to the details of both the underlying simulation engines
+and the programming language in question.
+
+Stable and healthy software ecosystems require standard interchange
+formats, allowing for the development of high-quality and long-lasting
+tools that produce and consume the standard.
+Demographic models are a key part of population genetics research,
+and to date the transfer of inferred models to downstream simulations
+has been ad-hoc, and conversions between the many different ways
+of expressing such models is both labour intensive and fraught with errors.
+The proposed Demes standard is an attempt to bridge this gap
+between inference and simulation, and also to provide the foundations
+for a sustainable ecosystem of tools built around this data model.
+Table~\ref{tab-software} shows some initial infrastructure that we have
+built as part of developing Demes, but many other useful tools
+can be envisaged that consume, transform, or produce this format.
+
+Reproducibility is a significant problem throughout the
+sciences~\citep{baker20161} and various measures have been
+proposed to increase the likelihood of researchers being
+able to replicate results in the
+literature~\citep{munafo2017manifesto}. The most basic requirement
+for reproducibility is that we must be able to state precisely what
+the result in question is. The lack of standardisation in how
+complex demographic models are communicated today, and the lack of
+precision in the published model descriptions means that it is difficult
+to replicate analyses, or reproduce those models for later simulation.
+Thus, we hope that the Demes standard introduced here will be widely adopted
+by simulation and inference methods and be used for reporting results in
+publications, either as supplemental material or uploaded to a data
+repository in Demes format.
+
+\krtcomment{Here, we should say something about other models that people may wish to implement within the demes
+framework.}
+
+\bibliographystyle{genetics}
+\bibliography{paper}
+
+\renewcommand{\thefigure}{A\arabic{figure}}
+\renewcommand{\thetable}{A\arabic{table}}
+\renewcommand{\theequation}{A\arabic{equation}}
+\renewcommand{\thesection}{A\arabic{section}}
+\setcounter{figure}{0}
+\setcounter{table}{0}
+\setcounter{equation}{0}
+
+\section*{Appendix}
 
@@
 The data model is based on the ubiquitous JSON~\citep{bray2017javascript}
-standard, and the structure formally defined using
+standard, and the structure is formally defined using
 JSON Schema~\citep{wright2020json}.
@@
 specifying the hierarchical structure and the properties that are allowed in different
-contexts along with their types and permissible ranges. This schema, and its
-accompanying documentation, fully describe the entities in the model and their
-relationships and the required behaviour of implementations. Since the
-schema is definitive, we will not recapitulate the details
+contexts along with their types and permissible ranges.
+This schema, and its accompanying documentation,
+fully describe the entities in the model and their
+relationships and the required behaviour of implementations.
+Since the schema is definitive, we will not recapitulate the details
 here, but instead focus on the high level properties of the model and
@@
 
-The design of Demes is a balance between two partially competing requirements:
-that (a) models should be easy for humans to understand and manipulate; and (b)
-software processing Demes models should be provided with an unambiguous
-representation that is straightforward to process. For efficiency of
-understanding and avoidance of model specification error, we require a data
-representation without redundancy (i.e., repetition of values). But, for the
-simplicity of software working with the Demes model (and the avoidance of
-programming error, or divergence in interpretations of the specification) it is
-preferable to have an explicit representation, in which all relevant values are
-readily available. Thus, we consider three entities: the high-level, human
-readable model specification; the low-level, fully-qualified model
-specification; and the parser, which is responsible for transforming the former
-into the latter.
+Below, we detail
+\begin{itemize}
+    \item The population genetics model in Section~\ref{sec:appendix-pop-gen-model}
+    \item The high-level model specification in Section~\ref{sec:appendix-spec}
+    \item The desired behaviour of Demes-format parsers in Section~\ref{sec:appendix-parsers}
+    \item And the intended scope of Demes in Section~\ref{sec:appendix-scope}
+\end{itemize}
 
-\subsection*{Population genetics model}
+\section{Population genetics model details}\label{sec:appendix-pop-gen-model}
 
@@
 
-Population sizes are given as the number of individuals, and details
+Population sizes are given as numbers of individuals, and details
 such as ploidy levels are considered external to the model.
-We focus on the number of individuals as opposed to the number of genomes
+We therefore focus on the number of individuals as opposed to the number of genomes.
 Sizes and mating system details are specified for each population within
@@
 
-\subsection*{High-level model specification}
+\section{High-level model specification}\label{sec:appendix-spec}
 
@@
 
-\subsection*{Parsers}
+\section{Parsers}\label{sec:appendix-parsers}
+
 The implicit value inference and default value propagation required to
@@
 
-\subsection*{Scope of the specification}
-\subsubsection*{Static models, not parameterised models}
+\section{Scope of the specification}
+\label{sec:appendix-scope}
+
+\subsection{Static models, not parameterised models}
+\label{sec:appendix-static}
+
 \jkcomment{This section is very rough.}
@@
 
-\subsubsection*{Population-level features, not genome features}
+\subsection{Population-level features, not genome features}
+\label{sec:appendix-features}
+
 Demes is designed to interface with a large number of simulation and inference
@@
 following semantic versioning guidelines.
-
-% Add a bit more space between rows.
-\renewcommand{\arraystretch}{1.5}
-\begin{table}
-\begin{center}
-\begin{tabular}{lp{12cm}}
-\toprule
-\multicolumn{2}{l}{Software infrastructure}\\
-\midrule
-\texttt{demes-python} &
-    A Python library for loading, saving, and working with
-    Demes models. Includes conversion utilities for existing
-    demographic model descriptions such as
-    \texttt{ms}~\citep{hudson2002generating}
-    (\url{https://github.com/popsim-consortium/demes-python}).\\
-
-\texttt{demes-c} &
-    A C library for parsing Demes YAML descriptions.
-    (\url{https://github.com/grahamgower/demes-c}). \\
-
-\texttt{demesdraw} &
-    A Python library for visualising Demes models
-    (\url{https://github.com/grahamgower/demesdraw}). \\
-
-\midrule
-\multicolumn{2}{l}{Methods using Demes as input/output format}\\
-\midrule
-
-% TOOL AUTHORS: please add a row to this table to summarise what the software
-% does, and how it implements demes.
-
-\dadi &
-    Summary of what dadi does and how it implements Demes~\citep{gutenkunst2009inferring}.
-    \\
-
-\fwdpy &
-    Forward-time Wright-Fisher simulation \citep{thornton2014cpp,thornton2019-nu}\\
-
-\gadma &
-    Demographic inference \citep{noskova2020gadma}.\\
-
-\moments &
-    Compute expected diversity statistics (SFS, LD) and perform
-    demographic inference using summary statistics
-    \citep{jouganous2017inferring,ragsdale2019models}.\\
-
-\msprime &
-    Coalescent simulation of Demes models, including command line
-    interface~\citep{kelleher2016efficient,kelleher2020coalescent,baumdicker2021-iu}.\\
-
-\stdpopsim &
-    A community-maintained collection of empirical data for species
-    and demographic models specified in Demes
-    format~\citep{adrion2020community}.\\
-
-\bottomrule
-\end{tabular}
-\end{center}
-\caption{\label{tab-software}
-Software support for Demes. We have included software infrastructure developed
-for working with Demes models (parsing, validation, visualisation, etc)
-as well as all known downstream software that implements the specification,
-at the time of writing.}
-\end{table}
-
-\section*{Discussion}
-
-The difficulty of describing complex demographic models
-for population genetic simulations has long been acknowledged
-~\citep[][e.g.]{antao2007modeler4simcoal2}, % jk - any older refs for this?
-and a number of different methods have been proposed to
-make such simulations more accessible and less error prone.
-Several Graphical User Interface (GUI) and visualisation
-methods have been developed, which greatly facilitate
-interpretation~\citep{mailund2005coasim,antao2007modeler4simcoal2,
-ewing2010msms,zhou2018popdemog}. However, these methods
-currently have little traction as they are all either directly coupled
-to an internal simulation method or to the specific syntax
-of a given simulator. Another approach that has been taken to
-simplify running such simulations is to provide an API in a
-scripting language such as Python or R, providing a flexible
-interface either to a built-in simulation
-engine~\citep{thornton2014cpp,thornton2019-nu,baumdicker2021-iu,kelleher2016efficient,haller2017flexible}
-% jk: are/were there other simulation frontends out there?
-or as a front end for other simulators~\citep{staab2016coala}.
-While APIs have many advantages and have grown in popularity in
-recent years, the demographic model descriptions are tied
-to the details of both the underlying simulation engines
-and the programming language in question.
-
-Stable and healthy software ecosystems require standard interchange
-formats, allowing for the development of high-quality and long-lasting
-tools that produce and consume the standard.
-Demographic models are a key part of population genetics research,
-and to date the transfer of inferred models to downstream simulations
-has been ad-hoc, and conversions between the many different ways
-of expressing such models is both labour intensive and fraught with errors.
-The proposed Demes standard is an attempt to bridge this gap
-between inference and simulation, and also to provide the foundations
-for a sustainable ecosystem of tools built around this data model.
-Table~\ref{tab-software} shows some initial infrastructure that we have
-built as part of developing Demes, but many other useful tools
-can be envisaged that consume, transform, or produce this format.
-
-Reproducibility is a significant problem throughout the
-sciences~\citep{baker20161} and various measures have been
-proposed to increase the likelihood of researchers being
-able to replicate results in the
-literature~\citep{munafo2017manifesto}. The most basic requirement
-for reproducibility is that we must be able to state precisely what
-the result in question is. The lack of standardisation in how
-complex demographic models are communicated today, and the lack of
-precision in the published model descriptions means that it is difficult
-to replicate analyses, or reproduce those models for later simulation.
-Thus, we hope that the Demes standard introduced here will be widely adopted
-by simulation and inference methods and be used for reporting results in
-publications, either as supplemental material or uploaded to a data
-repository in Demes format.
-
-\krtcomment{Here, we should say something about other models that people may wish to implement within the demes
-framework.}
-
-\bibliographystyle{plainnat}
-\bibliography{paper}
-
 \end{document}

### 2022-03-15 paper.tex
additional changes and showcase figure

@@
 
-\usepackage{filecontents,catchfile}
+\usepackage{catchfile}
 \begin{filecontents*}{software-table.tex}
@@
 visualisation of Demes models, and has been implemented in several popular
-inference and simulation methods (Table~\ref{tab-software}).
+inference and simulation methods (Table~\ref{tab:software}).
 We hope that this data model and file format will be widely adopted
@@
 of a multi-population demographic model.
-Detailed descriptions of the population genetic model, parsers, scope, and
+Detailed descriptions of the population genetics model, parsers, scope, and
 additional examples are given in the Appendix.
@@
     \end{center}
-\caption{\label{tab-software}
-Software support for Demes. We have included software infrastructure developed
-for working with Demes models (parsing, validation, visualisation, etc)
-as well as all known downstream software that implements the specification,
-at the time of writing.}
+    \caption{
+        \label{tab:software}
+        \textbf{Software support for Demes.}
+        We have included software infrastructure developed
+        for working with Demes models (parsing, validation,
+        visualisation, etc) as well as all known downstream 
+        software that implements the specification,
+        at the time of writing.
+    }
 \end{table}
 
-\subsection*{Population genetic model}
-1 paragraph
+\aprcomment{I want to keep the subsections that follow short and punchy.
+Refer readers to the Appendix for details.}
 
-\subsection*{High level specification}
-1 paragraph
+\subsection*{Population genetics model}
+
+Demographic models consist of one or more populations (or ``demes'') defined
+by their size histories and the time intervals of their existence.
+Individuals can move between populations based on their ancestor-descendant
+relationships (i.e., the topology) or by continuous or discrete migration events.
+The deomgraphic model itself omits information about genome biology, such as
+genome size and architecture, ploidy, and mutation and recombination rates.
+It also omits any information about selection operating within populations.
+Within populations, allele-frequency dynamics are implicitly assumed to
+follow the Wright-Fisher model, so that generations are discrete and
+non-overlapping and population sizes are independent of mean fitness
+(i.e., ``soft'' selection~\citep{christiansen1975hard}).
+While these restrictions preclude some population-genetic scenarios of
+interest, such as explicit spatial structure or
+\aprcomment{stochastic carrying-capacity models, or something like that},
+many current forwards- and backwards-in-time simulation software
+make these same assumptions, so that Demes is relevant to a range of
+widely-used simulators.
+In Appendix~\ref{sec:appendix-pop-gen-model}, we expand on these concepts
+and provide more details on the population genetics models supported by Demes.
+
+\subsection*{High-level specification}
+
+The high-level description of Demes models is focused on efficient human
+understanding and avoiding errors.
+We have adopted the widely used YAML format~\citep{ben2009yaml} as the
+primary interface for writing and interchangin demographic models
+(see Appendix~\ref{sec:appendix-spec} for details about our choice of
+YAML formatting for demographic models).
+Demographic models written in YAML format provide information about global
+features of the model (such as time units and generation times),
+populations (as ``demes'') and their existence intervals (as ``epochs''),
+and gene flow between populations
+(as continuous ``migrations'' or instantaneous ``pulse'' events)
+(e.g., Figures~\ref{fig:showcase}~and~\ref{fig:IM}).
+The high-level Demes specification encourages human understanding by avoiding
+redundancy in the description where possible and by providing a mechanism for
+specifying default values that are inherited hierarchically.
 
 \subsection*{Parsers}
-and languages. 1 paragraph
+
+While the high-level specification is designed for human readability and
+conciseness, the underlying data model suitable for software implementation
+is redundant and exhaustive. 
+Translation from the high-level specification to the low-level data model
+requires resolving heierarchically-defined default values and verifying
+relationships between populations and the validity of specified parameter
+values.
+Because this translation and validation requires significant
+programming effort, we define a standard software entity as part of the
+specification to perform this task (the parser),
+which is intended to be shared by programs that support Demes as an input.
+By having relatively few high-quality Demes parsers available as libraries
+(ideally one per programming language), we ensure consistency across simulation
+software in their implementation of Demes-specified models.
+We provide a reference implementation written in Python to resolve any
+potential ambiguities and to provide a helpful template for other
+implementations, as well as an extensive test suite of examples and the
+expected outputs.
+In addition, we have high-quality parser implementations in the Python and
+C languages (published under liberal open source licences), providing a
+solid foundation for the software ecosystem (Table~\ref{tab:software}).
+
+\aprcomment{redundancies with Appendix - that could be expanded and reworked a bit}
 
@@
 
-\begin{itemize}
-    \item Static, instead of parameterized models (1 paragraph)
-    \item Population level instead of genomic features (1 paragraph)
-\end{itemize}
+The Demes specification is static by design -- we wish to unambiguously
+describe a demographic model with a concrete set of parameters.
+This simplicity means that we cannot directly specify parameter distributions
+or estimated confidence intervals for those parameters.
+While it is not difficult to imagine extending the specification in ways that
+would allow this (Appendix~\ref{sec:appendix-scope}), it is not immediately
+clear that the benefits are worth the increased parser complexity.
+
+Demes is also designed to interface with a large number of simulation and
+inference methods, so we have restricted the specification to only
+describe demographic features at the population level.
+The underlying discrete demographic structure is the common thread connecting
+many widely-used population genetic methods, so we have omitted many
+features of the underlying genome biology (as described above, and in
+Appendix~\ref{sec:appendix-scope}).
 
@@
 \end{itemize}
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics{fig/showcase}
+    \caption{
+        \textbf{Illustration and simulation using Demes.}
+        \aprcomment{Extended snippet and Tennessen YAML in the Appendix.}
+    }
+    \label{fig:showcase}
+\end{figure}
 
@@
 for a sustainable ecosystem of tools built around this data model.
-Table~\ref{tab-software} shows some initial infrastructure that we have
+Table~\ref{tab:software} shows some initial infrastructure that we have
 built as part of developing Demes, but many other useful tools
@@
     \end{minipage}\\
-\caption{\label{fig-example}
+    \caption{
+        \label{fig:IM}
         Example isolation-with-migration Demes model. (A) The compact non-redundant
@@
 \aprcomment{not sure if this is sufficient here}
+
+\aprcomment{Need some information about defaults}
 

### 2022-03-15 fig/showcase.py

@@ -1,0 +1,129 @@
+"""
+
+Requires msprime~1.1, moments, demesdraw
+"""
+import msprime
+import moments
+import demes
+import demesdraw
+import numpy as np
+
+import matplotlib.pylab as plt
+import matplotlib.image as mpimg
+
+# set font sizes
+import matplotlib
+
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=6)
+matplotlib.rc("ytick", labelsize=6)
+matplotlib.rc("axes", labelsize=7)
+matplotlib.rc("axes", titlesize=7)
+matplotlib.rc("legend", fontsize=6)
+
+
+def plot_model(ax, graph):
+    demesdraw.tubes(
+        graph,
+        ax=ax,
+        num_lines_per_migration=3,
+        log_time=True,
+        positions={"ancestral": 0, "AFR": -300000, "EUR": 300000},
+    )
+    ax.set_ylim(top=1e6)
+
+
+def plot_sfs(ax, fs_moments, fs_msprime, ylabel=True):
+    ax.plot(fs_moments, "o", lw=0.5, ms=3, fillstyle="none", label="moments")
+    ax.plot(fs_msprime, ".--", ms=2, lw=0.5, label="msprime")
+    ax.set_xlabel("Derived allele count")
+    if ylabel:
+        ax.set_ylabel("Number observed")
+    ax.set_yscale("log")
+    ax.legend()
+    ax.set_xticks(np.arange(0, fs_moments.sample_sizes[0] + 1, 5))
+
+
+if __name__ == "__main__":
+    # load the graph
+    graph = demes.load("../models/tennessen.yml")
+
+    # set up the figure
+    fig = plt.figure(54321, figsize=(6.5, 3.0))
+    fig.clf()
+    grid = (2, 10)
+
+    # draw the graph using demesdraw
+    ax1 = plt.subplot2grid(grid, (0, 0), colspan=4, rowspan=2)
+    plot_model(ax1, graph)
+
+    # simulate data using msprime
+    demog = msprime.Demography.from_demes(graph)
+    L = 1e7
+    r = 2e-8
+    # grr.. 1 sample is actually 2 samples
+    samples = {"AFR": 20, "EUR": 20}
+
+    seed = 12345
+    ts = msprime.sim_ancestry(
+        {p: s // 2 for p, s in samples.items()},
+        demography=demog,
[…]

### 2022-03-22 paper.tex
Update paper.tex
Co-authored-by: Jerome Kelleher <jk@well.ox.ac.uk>

@@
 We have adopted the widely used YAML format~\citep{ben2009yaml} as the
-primary interface for writing and interchangin demographic models
+primary interface for writing and interchanging demographic models
 (see Appendix~\ref{sec:appendix-spec} for details about our choice of

### 2022-03-22 paper.tex
comments from jerome

@@
 It also omits any information about selection operating within populations.
-Within populations, allele-frequency dynamics are implicitly assumed to
+Within populations, allele-frequency dynamics are assumed to
 follow the Wright-Fisher model, so that generations are discrete and
@@
 While these restrictions preclude some population-genetic scenarios of
-interest, such as explicit spatial structure or
+interest, such as continuous spatial structure
+\aprcomment{citations here, and revisit} or
 \aprcomment{stochastic carrying-capacity models, or something like that},
@@
 widely-used simulators.
+\jkcomment{I guess these are also the basic assumptions also shared by
+inference methods, which is important. Mabye another sentence here making this
+explicit, and listing a few different types of inference methods and giving
+citations? This is critical - we want to establish that Demes is the
+"middleware" between the inference methods and the simulators, and captures the
+basic assumptions made by them all.}
 In Appendix~\ref{sec:appendix-pop-gen-model}, we expand on these concepts

### 2022-03-22 software-table.tex

\toprule
\multicolumn{2}{l}{Software infrastructure}\\
\midrule
\texttt{demes-python} &
    A Python library for loading, saving, and working with
    Demes models. Includes conversion utilities for existing
    demographic model descriptions such as
    \texttt{ms}~\citep{hudson2002generating}
    (\url{https://github.com/popsim-consortium/demes-python}).\\

\texttt{demes-c} &
    A C library for parsing Demes YAML descriptions.
    (\url{https://github.com/grahamgower/demes-c}). \\

\texttt{demesdraw} &
    A Python library for visualising Demes models
    (\url{https://github.com/grahamgower/demesdraw}). \\

\midrule
\multicolumn{2}{l}{Methods using Demes as input/output format}\\
\midrule

% TOOL AUTHORS: please add a row to this table to summarise what the software
% does, and how it implements demes.

\dadi &
    Summary of what dadi does and how it implements Demes~\citep{gutenkunst2009inferring}.
    \\

\fwdpy &
    Forward-time Wright-Fisher simulation \citep{thornton2014cpp,thornton2019-nu}\\

\gadma &
    Demographic inference \citep{noskova2020gadma}.\\

\moments &
    Compute expected diversity statistics (SFS, LD) and perform
    demographic inference using summary statistics
    \citep{jouganous2017inferring,ragsdale2019models}.\\

\msprime &
    Coalescent simulation of Demes models, including command line
    interface~\citep{kelleher2016efficient,kelleher2020coalescent,baumdicker2021-iu}.\\

\stdpopsim &
    A community-maintained collection of empirical data for species
    and demographic models specified in Demes
    format~\citep{adrion2020community}.\\

\bottomrule

### 2022-03-23 fig/showcase.py
update illustration

@@ -26,7 +26,7 @@ def plot_model(ax, graph):
     demesdraw.tubes(
         graph,
         ax=ax,
-        num_lines_per_migration=3,
+        num_lines_per_migration=2,
         log_time=True,
         positions={"ancestral": 0, "AFR": -300000, "EUR": 300000},
     )
@@ -51,22 +51,26 @@ if __name__ == "__main__":
     # set up the figure
     fig = plt.figure(54321, figsize=(6.5, 3.0))
     fig.clf()
-    grid = (2, 10)
+    grid = (5, 10)
 
     # draw the graph using demesdraw
-    ax1 = plt.subplot2grid(grid, (0, 0), colspan=4, rowspan=2)
+    ax1 = plt.subplot2grid(grid, (0, 0), colspan=4, rowspan=5)
     plot_model(ax1, graph)
 
     # simulate data using msprime
     demog = msprime.Demography.from_demes(graph)
     L = 1e7
     r = 2e-8
-    # grr.. 1 sample is actually 2 samples
+    #
     samples = {"AFR": 20, "EUR": 20}
+    msprime_samples = [
+        msprime.SampleSet(n, ploidy=1, population=p) for p, n in samples.items()
+    ]
 
-    seed = 12345
+    seed = 1234567
     ts = msprime.sim_ancestry(
-        {p: s // 2 for p, s in samples.items()},
+        msprime_samples,
+        # {p: s // 2 for p, s in samples.items()},
         demography=demog,
         sequence_length=L,
         recombination_rate=r,
@@ -99,16 +103,17 @@ if __name__ == "__main__":
     moments_eur = fs.marginalize([0])
 
     # plot AFR SFS comparison
-    ax2 = plt.subplot2grid(grid, (0, 4), colspan=3)
+    ax2 = plt.subplot2grid(grid, (0, 4), colspan=3, rowspan=2)
     plot_sfs(ax2, moments_afr, msprime_afr)
     ax2.set_title("AFR frequency spectrum")
 
     # plot EUR SFS comparison
-    ax3 = plt.subplot2grid(grid, (0, 7), colspan=3)
+    ax3 = plt.subplot2grid(grid, (0, 7), colspan=3, rowspan=2)
     plot_sfs(ax3, moments_eur, msprime_eur, ylabel=False)
     ax3.set_title("EUR frequency spectrum")
[…]

### 2022-04-04 paper.tex

@@
 \newcommand{\demes}[0]{\texttt{demes}}
+\newcommand{\demesdraw}[0]{\texttt{demesdraw}}
 \newcommand{\Demes}[0]{\texttt{Demes}}
@@
 
-\aprcomment{I want to keep the subsections that follow short and punchy.
-Refer readers to the Appendix for details.}
-
 \subsection*{Population genetics model}
@@
 
-\section*{Illustration: simulation using Demes}
+\section*{Application: simulation using Demes}
+
+\aprcomment{Need a paragraph or two here, extolling the virtues of simulation with demes}
 
@@
 
-\begin{figure}[ht!]
+\begin{figure}[tb!]
     \centering
@@
         \textbf{Illustration and simulation using Demes.}
-        \aprcomment{Extended snippet and Tennessen YAML in the Appendix.}
+        (A)
+        Using an inferred demographic model from \citet{tennessen2012evolution}
+        specified as a YAML in Demes format (Figure~\ref{fig:tennessen}), we
+        use \demesdraw\ to visualize the demographic model.
+        We then use \msprime\ to simulate genomic data for 20 genome copies
+        sampled from the two contemporary populations, and we use \moments\
+        to compute the expected joint site-frequency spectrum for the same
+        sample sizes (Figure~\ref{fig:tennessen-simulation}).
+        (B, C) We compare the single-population SFS in each population, showing
+        agreement between the simulation methods.
+        (D) Code snippets of the interactions between \demes\ and the simulation
+        software. An extended script to compute the SFS shown in (B) and (C) is
+        given in Figure~\ref{fig:tennessen-simulation}.
+        \aprcomment{There may be a better model to use here -- the massive
+        exponential growth really makes the rest of the history hard to see.}
     }
@@
 
-\begin{figure}
-    \begin{minipage}{0.44\textwidth}
-        \begin{tcolorbox}
-            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM.yaml}
-        \end{tcolorbox}
-        \begin{tcolorbox}
-            \includegraphics[width=\textwidth]{fig/IM}
-        \end{tcolorbox}
-    \end{minipage}\hfill
-    \begin{minipage}{0.54\textwidth}
-        \begin{tcolorbox}
-            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM-resolved.yaml}
-        \end{tcolorbox}
-    \end{minipage}\\
-    \caption{
-        \label{fig:IM}
-        Example isolation-with-migration Demes model. (A) The compact non-redundant
-        representation expressed using YAML. (B) The same model in the fully-qualified
-        form. (C) A visual representation of the model using \texttt{demesdraw}.
-        The YAML descriptions in (A) and (B) correspond exactly to a JSON description,
-        but are much more human-readable.
-    }
-\end{figure}
-
 This model is, by design, very limited. We do not specify any details of
@@
 following semantic versioning guidelines.
+
+\begin{figure}[h!]
+    \begin{minipage}{0.44\textwidth}
+        \begin{tcolorbox}
+            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM.yaml}
+        \end{tcolorbox}
+        \begin{tcolorbox}
+            \includegraphics[width=\textwidth]{fig/IM}
+        \end{tcolorbox}
+    \end{minipage}\hfill
+    \begin{minipage}{0.54\textwidth}
+        \begin{tcolorbox}
+            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM-resolved.yaml}
+        \end{tcolorbox}
+    \end{minipage}\\
+    \caption{
+        \label{fig:IM}
+        Example isolation-with-migration Demes model. (A) The compact non-redundant
+        representation expressed using YAML. (B) The same model in the fully-qualified
+        form. (C) A visual representation of the model using \texttt{demesdraw}.
+        The YAML descriptions in (A) and (B) correspond exactly to a JSON description,
+        but are much more human-readable.
+    }
+\end{figure}
+
+\begin{figure}[h!]
+    \begin{tcolorbox}
+        \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/tennessen.yml}
+    \end{tcolorbox}
+    \caption{
+        \textbf{The Tennessen et al. two-population demographic model in Demes format.}
+    }
+    \label{fig:tennessen}
+\end{figure}
+
+\begin{figure}[h!]
+    \begin{tcolorbox}
+        \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{python}{models/tennessen-simulation.py}
+    \end{tcolorbox}
+    \caption{
+        \textbf{Simulation of SFS for the Tennessen model.}
+        We first load the demographic model using \demes\ (as \texttt{graph}),
+        which can then be used by \msprime\ to create the demographic model used in
+        \texttt{msprime.sim\_ancestry()}. The same loaded graph can also be
+        passed to \moments\ to computed the expected joint SFS.
+        To compare the SFS in Figure~\ref{fig:showcase}, we \emph{marginalize} the
+        joint SFS to obtain the single-population SFS for both AFR and EUR populations.
+    }
+    \label{fig:tennessen-simulation}
+\end{figure}
+
 \end{document}

### 2022-04-04 paper.tex
fill in illustration text

@@
 \newcommand{\gadma}[0]{\texttt{GADMA}}
+\newcommand{\tskit}[0]{\texttt{tskit}}
 
@@
 
-\aprcomment{Need a paragraph or two here, extolling the virtues of simulation with demes}
-
-\begin{itemize}
-    \item Multi-population model with migrations and/or pulses
-    \item Demesdraw
-    \item msprime simulation
-    \item moments simulation
-    \item Compare SFS in each population
-    \item The input YAML file should go in the Appendix
-\end{itemize}
-
 \begin{figure}[tb!]
@@
 \end{figure}
+
+Here, we highlight the interface between Demes and other software, including
+simulation and model illustration tools. Demes allows us to specify a
+demographic model a single time, which can be used as the input for a growing
+number of simulation software (Table~\ref{tab:software}). We implemented the
+human two-population demographic model from \citet{tennessen2012evolution}
+inferred from European and African-American sequencing data. This model (given
+in Demes format in Figure~\ref{fig:tennessen}) is parameterized by an ancestral
+population with an ancient growth, divergence into ``AFR'' and ``EUR'' that
+each have multiple-epoch size histories, and multiple epochs of continuous
+migration between the two branches (illustrated using \demesdraw\ in
+Figure~\ref{fig:showcase}A). The large final sizes (\(\approx 500,000\)
+individuals each) are one to three orders of magnitude larger than ancestral
+population sizes, reflecting the recent explosive population size increase in
+human populations.
+
+We used this model to simulate 20 haploid genome copies from each EUR and AFR
+at time zero (i.e., present day) to obtain the joint site-frequency spectrum
+(SFS), a summary of observed allele frequencies widely used in evolutionary
+inference
+\citep{bustamante2001directional,gutenkunst2009inferring,tennessen2012evolution,jouganous2017inferring,kamm2017efficient,kim2017inference}.
+We used \msprime\ \citep{baumdicker2021-iu} to simulate a large recombining
+region with the mutation rate assumed in \citet{tennessen2012evolution} and
+computed the observed SFS using \tskit\ \citep{ralph2020efficiently}. We
+compared that simulated SFS to its expectation computed using \moments\
+\citep{jouganous2017inferring} (Figure~\ref{fig:showcase}B,C). For both
+\msprime\ and \moments\ simulations, the same demographic model was used as
+input, avoiding the need to specify the complex demographic model multiple
+times using the unique API of each software.
 
@@
     \caption{
-        \textbf{The Tennessen et al. two-population demographic model in Demes format.}
+        \textbf{The \citet{tennessen2012evolution} two-population demographic model in Demes format.}
     }
@@
     \begin{tcolorbox}
-        \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{python}{models/tennessen-simulation.py}
+        % NOTE: line highlights may need adjusting after changing this code
+        \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt,highlightlines={9,19,52}]{python}{models/tennessen-simulation.py}
     \end{tcolorbox}
@@
         joint SFS to obtain the single-population SFS for both AFR and EUR populations.
+        Lines interfacing \demes\ and other software are highlighted.
     }

### 2022-04-05 paper.tex
Expanded on selfing, clonal rates

--- 2022-04-04 paper.tex
@@
 \newcommand{\krtcomment}[1]{{\textcolor{purple}{KRT: #1}}}
+\newcommand{\mhcomment}[1]{{\textcolor{cyan}{MH: #1}}}
 
@@
 is at best tedious and error prone~\citep{ragsdale2020lessons}
-and occasionally impossible because of missing information.
+and occasionally impossible because of missing information. \mhcomment{Maybe add a short, concrete example to convince readers of the problem?}
 %\aprcomment{This feels tangential - maybe discussion}
@@
 by the community, such that users can expect to simulate directly
-from inferred models with little or no programming effort.
+from inferred models with little programming effort.
 
@@
 	and why that is required for the standard coalescent.}
+\mhcomment{The `exchangeability' assumption also breaks down under different reproductive modes, since the sampling of the alleles affects coalescent times.
+For example, in a highly selfing species the coalescent time for two different alleles within an individual is much lower than two alleles from different individuals.
+I've tried to rewrite the below paragraph to update and clarify the reproduction parameters in line with the discussions on Github.
+Finally, 'rates' imply events occur as Poisson-like processes over continuous time, whereas the values used here instead define the \emph{fraction} of reproductions of each type.
+Shall we rename to selfing/clonal `fractions', or would that mess up the rest of the schema too much?}
 
-Within a population, individuals are assumed to be exchangeable, but there are
-also parameters for nonrandom mating, such as selfing or cloning rates, which
-give the probability that offspring are generated from one generation to the
-next by self-fertilisation or cloning of an individual. Cloning and selfing
-rates can take any value between zero and one, and if both are specified then
-the selfing rate is the probability an offspring is generated through
-self-fertilisation given it was produced sexually (i.e. with probability one
-minus the cloning rate). We omit any details regarding selection.
+Each population has a selfing rate and cloning rate assigned to it, where each denotes the probability that offspring are generated from one generation to the next by either self-fertilisation or cloning of an individual.
+More specifically, denote the clonal rate by $\sigma$ and the selfing rate by $S$.
+In each population a proportion of offspring $\sigma$ reproduce clonally, while $1-\sigma$ reproduce sexually.
+Within the sexually--reproducing offspring, $S$ were born via self--fertilisation while the rest were outcrossed.
+Hence $S$ and $\sigma$ can take any value between zero and one, but do not have to sum to one.
+See \citet{hartfield2016facsexcoal} for more details of the mathematical properties of these rates in a coalescent context.
+\mhcomment{If requested I can quickly draw up a schematic diagram to illustrate these partitions.} Note that even if $S$, $\sigma$ are set to zero then some residual selfing can occur due to sampling in the Wright--Fisher model.
+
+%Within a population, individuals are assumed to be exchangeable, but there are
+%also parameters for nonrandom mating, such as selfing or cloning rates, which
+%give the probability that offspring are generated from one generation to the
+%next by self-fertilisation or cloning of an individual. Cloning and selfing
+%rates can take any value between zero and one, and if both are specified then
+%the selfing rate is the probability an offspring is generated through
+%self-fertilisation given it was produced sexually (i.e. with probability one
+%minus the cloning rate).
+
+We omit any details regarding selection.
 Thus, the history of population size changes does not depend on mean fitnesses within
@@
 
-\begin{figure}[h!]
-    \begin{minipage}{0.44\textwidth}
-        \begin{tcolorbox}
-            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM.yaml}
-        \end{tcolorbox}
-        \begin{tcolorbox}
-            \includegraphics[width=\textwidth]{fig/IM}
-        \end{tcolorbox}
-    \end{minipage}\hfill
-    \begin{minipage}{0.54\textwidth}
-        \begin{tcolorbox}
-            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM-resolved.yaml}
-        \end{tcolorbox}
-    \end{minipage}\\
-    \caption{
-        \label{fig:IM}
-        Example isolation-with-migration Demes model. (A) The compact non-redundant
-        representation expressed using YAML. (B) The same model in the fully-qualified
-        form. (C) A visual representation of the model using \texttt{demesdraw}.
-        The YAML descriptions in (A) and (B) correspond exactly to a JSON description,
-        but are much more human-readable.
-    }
-\end{figure}
+%\begin{figure}[h!]
+%    \begin{minipage}{0.44\textwidth}
+%        \begin{tcolorbox}
+%            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM.yaml}
+%        \end{tcolorbox}
+%        \begin{tcolorbox}
+%            \includegraphics[width=\textwidth]{fig/IM}
+%        \end{tcolorbox}
+%    \end{minipage}\hfill
+%    \begin{minipage}{0.54\textwidth}
+%        \begin{tcolorbox}
+%            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM-resolved.yaml}
+%        \end{tcolorbox}
+%    \end{minipage}\\
+%    \caption{
+%        \label{fig:IM}
+%        Example isolation-with-migration Demes model. (A) The compact non-redundant
+%        representation expressed using YAML. (B) The same model in the fully-qualified
+%        form. (C) A visual representation of the model using \texttt{demesdraw}.
+%        The YAML descriptions in (A) and (B) correspond exactly to a JSON description,
+%        but are much more human-readable.
+%    }
+%\end{figure}
 

### 2022-04-05 paper.tex
uncommented IM figure

@@
 
-%\begin{figure}[h!]
-%    \begin{minipage}{0.44\textwidth}
-%        \begin{tcolorbox}
-%            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM.yaml}
-%        \end{tcolorbox}
-%        \begin{tcolorbox}
-%            \includegraphics[width=\textwidth]{fig/IM}
-%        \end{tcolorbox}
-%    \end{minipage}\hfill
-%    \begin{minipage}{0.54\textwidth}
-%        \begin{tcolorbox}
-%            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM-resolved.yaml}
-%        \end{tcolorbox}
-%    \end{minipage}\\
-%    \caption{
-%        \label{fig:IM}
-%        Example isolation-with-migration Demes model. (A) The compact non-redundant
-%        representation expressed using YAML. (B) The same model in the fully-qualified
-%        form. (C) A visual representation of the model using \texttt{demesdraw}.
-%        The YAML descriptions in (A) and (B) correspond exactly to a JSON description,
-%        but are much more human-readable.
-%    }
-%\end{figure}
+\begin{figure}[h!]
+    \begin{minipage}{0.44\textwidth}
+        \begin{tcolorbox}
+            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM.yaml}
+        \end{tcolorbox}
+        \begin{tcolorbox}
+            \includegraphics[width=\textwidth]{fig/IM}
+        \end{tcolorbox}
+    \end{minipage}\hfill
+    \begin{minipage}{0.54\textwidth}
+        \begin{tcolorbox}
+            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM-resolved.yaml}
+        \end{tcolorbox}
+    \end{minipage}\\
+    \caption{
+        \label{fig:IM}
+        Example isolation-with-migration Demes model. (A) The compact non-redundant
+        representation expressed using YAML. (B) The same model in the fully-qualified
+        form. (C) A visual representation of the model using \texttt{demesdraw}.
+        The YAML descriptions in (A) and (B) correspond exactly to a JSON description,
+        but are much more human-readable.
+    }
+\end{figure}
 

### 2022-04-05 paper.tex
Added line breaks

@@
 is at best tedious and error prone~\citep{ragsdale2020lessons}
-and occasionally impossible because of missing information. \mhcomment{Maybe add a short, concrete example to convince readers of the problem?}
+and occasionally impossible because of missing information. 
+\mhcomment{Maybe add a short, concrete example to convince readers of the problem?}
 %\aprcomment{This feels tangential - maybe discussion}
@@
 	and why that is required for the standard coalescent.}
-\mhcomment{The `exchangeability' assumption also breaks down under different reproductive modes, since the sampling of the alleles affects coalescent times.
-For example, in a highly selfing species the coalescent time for two different alleles within an individual is much lower than two alleles from different individuals.
-I've tried to rewrite the below paragraph to update and clarify the reproduction parameters in line with the discussions on Github.
-Finally, 'rates' imply events occur as Poisson-like processes over continuous time, whereas the values used here instead define the \emph{fraction} of reproductions of each type.
-Shall we rename to selfing/clonal `fractions', or would that mess up the rest of the schema too much?}
+\mhcomment{The `exchangeability' assumption also breaks down under 
+	different reproductive modes, since the sampling of the alleles affects coalescent times. 
+	For example, in a highly selfing species the coalescent time for two different alleles 
+	within an individual is much lower than two alleles from different individuals. I've tried 
+	to rewrite the below paragraph to update and clarify the reproduction parameters in 
+	line with the discussions on Github. Finally, 'rates' imply events occur as Poisson-like 
+	processes over continuous time, whereas the values used here instead define the 
+	\emph{fraction} of reproductions of each type. Shall we rename to selfing/clonal `fractions',
+	 or would that mess up the rest of the schema too much?}
 
-Each population has a selfing rate and cloning rate assigned to it, where each denotes the probability that offspring are generated from one generation to the next by either self-fertilisation or cloning of an individual.
-More specifically, denote the clonal rate by $\sigma$ and the selfing rate by $S$.
-In each population a proportion of offspring $\sigma$ reproduce clonally, while $1-\sigma$ reproduce sexually.
-Within the sexually--reproducing offspring, $S$ were born via self--fertilisation while the rest were outcrossed.
-Hence $S$ and $\sigma$ can take any value between zero and one, but do not have to sum to one.
-See \citet{hartfield2016facsexcoal} for more details of the mathematical properties of these rates in a coalescent context.
-\mhcomment{If requested I can quickly draw up a schematic diagram to illustrate these partitions.} Note that even if $S$, $\sigma$ are set to zero then some residual selfing can occur due to sampling in the Wright--Fisher model.
-
-%Within a population, individuals are assumed to be exchangeable, but there are
-%also parameters for nonrandom mating, such as selfing or cloning rates, which
-%give the probability that offspring are generated from one generation to the
-%next by self-fertilisation or cloning of an individual. Cloning and selfing
-%rates can take any value between zero and one, and if both are specified then
-%the selfing rate is the probability an offspring is generated through
-%self-fertilisation given it was produced sexually (i.e. with probability one
-%minus the cloning rate).
+Each population has a selfing rate and cloning rate assigned to it, where each denotes 
+the probability that offspring are generated from one generation to the next by either 
+self-fertilisation or cloning of an individual. More specifically, denote the clonal rate by 
+$\sigma$ and the selfing rate by $S$. In each population a proportion of offspring 
+$\sigma$ reproduce clonally, while $1-\sigma$ reproduce sexually. Within the 
+sexually--reproducing offspring, $S$ were born via self--fertilisation while the rest were 
+outcrossed. Hence $S$ and $\sigma$ can take any value between zero and one, but 
+do not have to sum to one. See \citet{hartfield2016facsexcoal} for more details of the 
+mathematical properties of these rates in a coalescent context. \mhcomment{If requested 
+I can quickly draw up a schematic diagram to illustrate these partitions.} Note that even if 
+$S$, $\sigma$ are set to zero then some residual selfing can occur due to sampling in 
+the Wright--Fisher model.
 

### 2022-04-05 paper.tex
Update paper.tex
Co-authored-by: Jerome Kelleher <jk@well.ox.ac.uk>

@@
 \newcommand{\krtcomment}[1]{{\textcolor{purple}{KRT: #1}}}
-\newcommand{\mhcomment}[1]{{\textcolor{cyan}{MH: #1}}}
 
@@
 and occasionally impossible because of missing information.
-\mhcomment{Maybe add a short, concrete example to convince readers of the problem?}
 %\aprcomment{This feels tangential - maybe discussion}
@@
 by the community, such that users can expect to simulate directly
-from inferred models with little programming effort.
+from inferred models with little or no programming effort.
 
@@
 \end{figure}
+
+Here, we highlight the interface between Demes and other software, including
+simulation and model illustration tools. Demes allows us to specify a
+demographic model which can be used as the input for a growing
+number of simulation software (Table~\ref{tab:software}). We implemented the
+human two-population demographic model from \citet{tennessen2012evolution}
+inferred from European and African-American sequencing data. This model (given
+in Demes format in Figure~\ref{fig:tennessen}) is parameterized by an ancestral
+population with an ancient growth, divergence into ``AFR'' and ``EUR'' that
+each have multiple-epoch size histories, and multiple epochs of continuous
+migration between the two branches (illustrated using \demesdraw\ in
+Figure~\ref{fig:showcase}A). The large final sizes (\(\approx 500,000\)
+individuals each) are one to three orders of magnitude larger than ancestral
+population sizes, reflecting the recent explosive population size increase in
+human populations.
+
+We used this model to simulate 20 haploid genome copies from each EUR and AFR
+at time zero (i.e., present day) to obtain the joint site-frequency spectrum
+(SFS), a summary of observed allele frequencies widely used in evolutionary
+inference
+\citep{bustamante2001directional,gutenkunst2009inferring,tennessen2012evolution,jouganous2017inferring,kamm2017efficient,kim2017inference}.
+We used \msprime\ \citep{baumdicker2021-iu} to simulate a large recombining
+region with the mutation rate assumed in \citet{tennessen2012evolution} and
+computed the observed SFS using \tskit\ \citep{ralph2020efficiently}. We
+compared that simulated SFS to its expectation computed using \moments\
+\citep{jouganous2017inferring} (Figure~\ref{fig:showcase}B,C). For both
+\msprime\ and \moments\ simulations, the same demographic model was used as
+input, avoiding the need to specify the complex demographic model multiple
+times using the unique API of each software.
 
@@
 	and why that is required for the standard coalescent.}
-\mhcomment{The `exchangeability' assumption also breaks down under 
-	different reproductive modes, since the sampling of the alleles affects coalescent times. 
-	For example, in a highly selfing species the coalescent time for two different alleles 
-	within an individual is much lower than two alleles from different individuals. I've tried 
-	to rewrite the below paragraph to update and clarify the reproduction parameters in 
-	line with the discussions on Github. Finally, 'rates' imply events occur as Poisson-like 
-	processes over continuous time, whereas the values used here instead define the 
-	\emph{fraction} of reproductions of each type. Shall we rename to selfing/clonal `fractions',
-	 or would that mess up the rest of the schema too much?}
 
-Each population has a selfing rate and cloning rate assigned to it, where each denotes 
-the probability that offspring are generated from one generation to the next by either 
-self-fertilisation or cloning of an individual. More specifically, denote the clonal rate by 
-$\sigma$ and the selfing rate by $S$. In each population a proportion of offspring 
-$\sigma$ reproduce clonally, while $1-\sigma$ reproduce sexually. Within the 
-sexually--reproducing offspring, $S$ were born via self--fertilisation while the rest were 
-outcrossed. Hence $S$ and $\sigma$ can take any value between zero and one, but 
-do not have to sum to one. See \citet{hartfield2016facsexcoal} for more details of the 
-mathematical properties of these rates in a coalescent context. \mhcomment{If requested 
-I can quickly draw up a schematic diagram to illustrate these partitions.} Note that even if 
-$S$, $\sigma$ are set to zero then some residual selfing can occur due to sampling in 
-the Wright--Fisher model.
-
-We omit any details regarding selection.
+Within a population, individuals are assumed to be exchangeable, but there are
+also parameters for nonrandom mating, such as selfing or cloning rates, which
+give the probability that offspring are generated from one generation to the
+next by self-fertilisation or cloning of an individual. Cloning and selfing
+rates can take any value between zero and one, and if both are specified then
+the selfing rate is the probability an offspring is generated through
+self-fertilisation given it was produced sexually (i.e. with probability one
+minus the cloning rate). We omit any details regarding selection.
 Thus, the history of population size changes does not depend on mean fitnesses within

### 2022-04-06 paper.tex
Update paper.tex
Co-authored-by: Jerome Kelleher <jk@well.ox.ac.uk>

@@
 demographic model which can be used as the input for a growing
-number of simulation software (Table~\ref{tab:software}). We implemented the
+number of simulation packages (Table~\ref{tab:software}). We implemented the
 human two-population demographic model from \citet{tennessen2012evolution}
-inferred from European and African-American sequencing data. This model (given
+inferred from European and African-American sequencing data. This model (shown
 in Demes format in Figure~\ref{fig:tennessen}) is parameterized by an ancestral
@@
 \citep{bustamante2001directional,gutenkunst2009inferring,tennessen2012evolution,jouganous2017inferring,kamm2017efficient,kim2017inference}.
-We used \msprime\ \citep{baumdicker2021-iu} to simulate a large recombining
-region with the mutation rate assumed in \citet{tennessen2012evolution} and
-computed the observed SFS using \tskit\ \citep{ralph2020efficiently}. We
-compared that simulated SFS to its expectation computed using \moments\
-\citep{jouganous2017inferring} (Figure~\ref{fig:showcase}B,C). For both
-\msprime\ and \moments\ simulations, the same demographic model was used as
-input, avoiding the need to specify the complex demographic model multiple
-times using the unique API of each software.
+The Demes model (Figure~\ref{fig:showcase}A~and~\ref{fig:tennessen}) was
+provided as the input demography to \msprime\ \citep{baumdicker2021-iu} to
+simulate a large recombining region with the mutation rate assumed in
+\citet{tennessen2012evolution}, and we computed the observed SFS using \tskit\
+\citep{ralph2020efficiently}. Using the same Demes model as input to \moments\
+\citep{jouganous2017inferring}, we computed the expectation of the joint SFS
+and compared to the \msprime\ simulated data (Figure~\ref{fig:showcase}B,C).
+To reiterate, for both \msprime\ and \moments\ simulations, the same
+demographic model was used as input, avoiding the need to specify the complex
+demographic model multiple times using the unique API of each software.
 

### 2022-04-07 paper.tex
Updating ms, resolving bib conflicts

--- 2022-04-05 paper.tex
@@
 outcrossed. Hence $S$ and $\sigma$ can take any value between zero and one, but 
-do not have to sum to one. See \citet{hartfield2016facsexcoal} for more details of the 
+can sum to more than one. See \citet{hartfield2016facsexcoal} for more details of the 
 mathematical properties of these rates in a coalescent context. \mhcomment{If requested 
-I can quickly draw up a schematic diagram to illustrate these partitions.} Note that even if 
-$S$, $\sigma$ are set to zero then some residual selfing can occur due to sampling in 
-the Wright--Fisher model.
+I can quickly draw up a schematic diagram to illustrate these partitions.}
 
@@
     \item By allowing selfing and cloning, we allow many "standard" models to be considered with "no or little"
-          extra effort.
+          extra effort. \mhcomment{Plus something on how programs deal with residual selfing.}
     \item However, by parameterising selfing and cloning as we have, we are assuming that these properties of

### 2022-04-07 paper.tex

--- 2022-04-06 paper.tex
@@
 and occasionally impossible because of missing information. 
+\mhcomment{Maybe add a short, concrete example to convince readers of the problem?}
 %\aprcomment{This feels tangential - maybe discussion}
@@
 	and why that is required for the standard coalescent.}
+\mhcomment{The `exchangeability' assumption also breaks down under 
+	different reproductive modes, since the sampling of the alleles affects coalescent times. 
+	For example, in a highly selfing species the coalescent time for two different alleles 
+	within an individual is much lower than two alleles from different individuals. I've tried 
+	to rewrite the below paragraph to update and clarify the reproduction parameters in 
+	line with the discussions on Github. Finally, 'rates' imply events occur as Poisson-like 
+	processes over continuous time, whereas the values used here instead define the 
+	\emph{fraction} of reproductions of each type. Shall we rename to selfing/clonal `fractions',
+	 or would that mess up the rest of the schema too much?}
 
-Within a population, individuals are assumed to be exchangeable, but there are
-also parameters for nonrandom mating, such as selfing or cloning rates, which
-give the probability that offspring are generated from one generation to the
-next by self-fertilisation or cloning of an individual. Cloning and selfing
-rates can take any value between zero and one, and if both are specified then
-the selfing rate is the probability an offspring is generated through
-self-fertilisation given it was produced sexually (i.e. with probability one
-minus the cloning rate). We omit any details regarding selection.
+Each population has a selfing rate and cloning rate assigned to it, where each denotes 
+the probability that offspring are generated from one generation to the next by either 
+self-fertilisation or cloning of an individual. More specifically, denote the clonal rate by 
+$\sigma$ and the selfing rate by $S$. In each population a proportion of offspring 
+$\sigma$ reproduce clonally, while $1-\sigma$ reproduce sexually. Within the 
+sexually--reproducing offspring, $S$ were born via self--fertilisation while the rest were 
+outcrossed. Hence $S$ and $\sigma$ can take any value between zero and one, but 
+can sum to more than one. See \citet{hartfield2016facsexcoal} for more details of the 
+mathematical properties of these rates in a coalescent context. \mhcomment{If requested 
+I can quickly draw up a schematic diagram to illustrate these partitions.}
+
+We omit any details regarding selection.
 Thus, the history of population size changes does not depend on mean fitnesses within

### 2022-04-08 paper.tex
Clean up the pop gen model section

--- 2022-04-06 paper.tex
@@
 relationships and the required behaviour of implementations.
-Since the schema is definitive, we will not recapitulate the details
+Since the schema is definitive,
+\aprcomment{what exactly do we mean by ``definitive''?}
+we will not recapitulate the details
 here, but instead focus on the high level properties of the model and
@@
 
-Below, we detail
-\begin{itemize}
-    \item The population genetics model in Section~\ref{sec:appendix-pop-gen-model}
-    \item The high-level model specification in Section~\ref{sec:appendix-spec}
-    \item The desired behaviour of Demes-format parsers in Section~\ref{sec:appendix-parsers}
-    \item And the intended scope of Demes in Section~\ref{sec:appendix-scope}
-\end{itemize}
+Below we detaail the population genetics model in 
+Appendix~\ref{sec:appendix-pop-gen-model},
+the high-level, human-readable model specification in
+Appendix~\ref{sec:appendix-spec}, demographic model parsers and their
+intended behaviour in Appendix~\ref{sec:appendix-parsers}, and the
+intended scope of Demes in Appendix~\ref{sec:appendix-scope}.
+The extended figures in Appendix~\ref{sec:appendix-figures} provides
+additional examples and illustrations of Demes usage.
 
-\section{Population genetics model details}\label{sec:appendix-pop-gen-model}
+\section{Population genetics model details}
+\label{sec:appendix-pop-gen-model}
 
 In Demes, demographic models consist of one or more interacting populations,
-or ``demes'', by which we mean a grouping of individuals that is convenient
-for modelling \citep{gilmour_demes_1939,gilmour_deme_1955}.
+or ``demes'', understood to be a collection of individuals that can be
+conveniently modelled using a defined set of rules and parameters
+\citep{gilmour_demes_1939,gilmour_deme_1955}.
 To avoid confusion with the name of the specification itself we will
 use the term ``population'' in this discussion, with the understanding that the
-terms are interchangeable for our purposes.
+terms are interchangeable.
 A population is defined as some collection of individuals that exists for
@@
 during that time period. Individuals can move between populations
-either through ancestor-descendant relationships between the populations,
-or through processes of continuous or pulse migration.
-No other properties of the populations are specified in the model:
-we are concerned only with defining the populations, their sizes, and the
+either according to their ancestor-descendant relationships
+or through processes of continuous or pulse migrations.
+Few other properties of the populations are specified in the model:
+we are concerned primarily with defining the populations, their sizes, and the
 movement of individuals between those populations.
+\aprcomment{previous sentence changed to allow for selfing/cloning}
 
@@
 corresponds to the final generation or ``now'', and event times in the past are
-values greater than zero with larger values for events that occur in the more
+values greater than zero with larger values corresponding to times in the more
 distant past. By having time units increase into the past, we avoid the need to
-choose an arbitrary point in history as ``time zero'', so that instead zero
-refers to ``now'' or the final generation of a simulation. A natural
+choose an arbitrary point in history as ``time zero''. A natural
 specification for time units is in generations, although other time units are
@@
 such as ploidy levels are considered external to the model.
-We therefore focus on the number of individuals as opposed to the number of genomes.
+We therefore focus on the number of individuals as opposed
+to the number of genome copies.
 Sizes and mating system details are specified for each population within
@@
 
-\krtcomment{The paragraph(s) below subsume(s)/replace(s) some of what appears in later pargraphs.
-	I'm not making an attempt yet to revise that later content--I think other authors should
-	weigh in.}
+%\krtcomment{The paragraph(s) below subsume(s)/replace(s) some of what appears in later pargraphs.
+%	I'm not making an attempt yet to revise that later content--I think other authors should
+%	weigh in.}
 
-\krtcomment{Aaron -- I have perhaps over-simplified the description of marginal allele frequency dynamics?}
+%\krtcomment{Aaron -- I have perhaps over-simplified the description of marginal allele frequency dynamics?}
 
 Within a population, we implicitly assume that allele frequency dynamics can
-be described by the Wright-Fisher model.  Briefly, generations are non-overlapping (all parents
-reproduce and die simultaneously) and the expected frequency (at birth) of the $i^{th}$ allele currently at frequency $p_i$
-one generation in the next generation is $p_iw_i/\bar{w}$, where $w_i$ and $\bar{w}$ are the marginal and
-mean fitnesses, respectively, properly weighted according to ancestry proportions. In this framework,
-a forward-time simulation of finite populations is equivalent to multinomial sampling of allele frequencies each
-generation (\citet[][pp 29-31]{burger2000-ul}, \citet[][pp 179-181]{crowkimura1970})
-and a backwards-time (coalescent) simulation follows the approximations described in \cite[][chapter 3]{wakeley2008-hd}.
-Further, this model assumes "soft" selection \citep{christiansen1975hard}, meaning that the dynamics of population sizes
+be described by the Wright-Fisher model.
+Briefly, generations are non-overlapping (all parents
+reproduce and die simultaneously) and the expected frequency (at birth)
+of the $i^{th}$ allele currently at frequency $p_i$
+one generation in the next generation is $p_iw_i/\bar{w}$,
+where $w_i$ and $\bar{w}$ are the marginal and
+mean fitnesses, respectively, properly weighted according to ancestry proportions.
+In this framework,
+a forward-time simulation of finite populations is equivalent
+to multinomial sampling of allele frequencies each
+generation (\citet[][pp 29-31]{burger2000-ul}, \citet[][pp 179-181]{crowkimura1970}),
+and a backwards-time (coalescent) simulation follows the approximations
+described in \cite[][chapter 3]{wakeley2008-hd}.
+Further, this model assumes "soft" selection \citep{christiansen1975hard},
+meaning that the dynamics of population sizes
 changes are \textit{independent} of the details of individual fitnesses.
-In the \texttt{Discussion}, we discuss alternative models compatible with this framework and outline the expectations for tools implementing
-them.
-A plethora of forwards and backwards time simulators currently implement this model (REFS).
+% In the \texttt{Discussion}, we discuss alternative models compatible
+% with this framework and outline the expectations for tools implementing
+% them.
+\aprcomment{As such, this model excludes scenarios such as ``hard selection,''
+in which population sizes are dependent on a population's mean fitness, or
+stochastic fluctuations in population size, such as defining ``carrying capacities''.}
+Many forwards and backwards time simulators currently implement this model
+\citep[e.g.,][\aprcomment{and more\dots?}]{hudson2002generating,gutenkunst2009inferring,excoffier2011fastsimcoal,jouganous2017inferring,haller2019slim,thornton2019-nu}.
 Each has their own unique method of specifying the details of selection and demography.
@@
 
-\krtcomment{We need to revisit the paragraph(s) below.  We do not assume individuals are exchangeable, because we do not
-	require that all individuals have the same fitness. See, e.g., p 60 of Wakeley for why fitness == exchangeability
-	and why that is required for the standard coalescent.}
+% \krtcomment{
+%     We need to revisit the paragraph(s) below.
+%     We do not assume individuals are exchangeable, because we do not
+% 	require that all individuals have the same fitness.
+%     See, e.g., p 60 of Wakeley for why fitness == exchangeability
+% 	and why that is required for the standard coalescent.
+% }
+% \mhcomment{The `exchangeability' assumption also breaks down under 
+% 	different reproductive modes, since the sampling of the alleles
+%     affects coalescent times. 
+% 	For example, in a highly selfing species the coalescent time for two different alleles 
+% 	within an individual is much lower than two alleles from different individuals.
+%     I've triedto rewrite the below paragraph to update and clarify
+%     the reproduction parameters in line with the discussions on Github.
+%     Finally, 'rates' imply events occur as Poisson-like 
+% 	processes over continuous time, whereas the values used here instead define the 
+% 	\emph{fraction} of reproductions of each type.
+%     Shall we rename to selfing/clonal `fractions',
+% 	or would that mess up the rest of the schema too much?
+% }
 
-Within a population, individuals are assumed to be exchangeable, but there are
-also parameters for nonrandom mating, such as selfing or cloning rates, which
-give the probability that offspring are generated from one generation to the
-next by self-fertilisation or cloning of an individual. Cloning and selfing
-rates can take any value between zero and one, and if both are specified then
-the selfing rate is the probability an offspring is generated through
-self-fertilisation given it was produced sexually (i.e. with probability one
-minus the cloning rate). We omit any details regarding selection.
-Thus, the history of population size changes does not depend on mean fitnesses within
-a population nor on differences in mean fitness between populations.
-This indepedence of population size and
-selection is often called ``soft" selection \citep{christiansen1975hard}.
+Each population has an assigned selfing rate and cloning rate, where each defines
+the probability that offspring are generated from one generation to the next by either
+self-fertilisation or cloning of an individual.
+More specifically, for a given epoch within a population denote the clonal rate by
+$\sigma$ and the selfing rate by $S$.
+$S$ and $\sigma$ can take any value between zero and one and
+can sum to more than one.
+Each generation a proportion of offspring
+$\sigma$ are expected to be generated through clonal reproduction,
+while $1-\sigma$ are expected to arise through sexual reproduction.
+Within the sexually-reproduced offspring,
+$S$ are born via self-fertilisation while the rest
+have parents drawn at random from the previous generation.
+Depending on the simulation software, this random drawing of parent may occur
+either with or without replacement. When drawing occurs with replacement, a small
+amount of ``residual'' selfing is expected, so that the realized selfing probability
+is $(1-\sigma)(S + (1-S)/N)$ instead of $(1-\sigma)S$ (so that even with $\sigma=0$
+and $S=0$, selfing may still occur with probability $1/N$).
+Simulation software that allow variable rates of selfing are expected to clearly
+document the expected behaviour.
 
-List of thoughts to turn into a paragraph:
+By allowing the definition of selfing and cloning probabilities, we allow many 
+standard models to be defined.
+However, by parameterising selfing and cloning as we have, we assume that these
+properties of populations can be specified independently from the genetics.
+In other words, mutations that cause selfing probabilities to fluctuate within
+an epoch are not considered.
+\aprcomment{
+    Do we still want to say something along these lines?
+    ``However, consuming software is free to ignore these parameters
+    and model the genetics within
+    the framework of a soft selection model.
+    We don't view these as limitations. Rather, this is what most people
+    are doing most of the time.
+    Just now we can finally try to do this part w/fewer errors.''
+}
+More details of the mathematical properties of selfing and cloning rates
+in a coalescent context can be found in \citet{hartfield2016facsexcoal}.
 
-\begin{enumerate}
-    \item Selection parameters may be arbitrarily complex.  Different fitness effects in different demes.
-          Variation over time.  Etc..
-          The specification would become unwieldy and perhaps limiting.
-    \item By allowing selfing and cloning, we allow many "standard" models to be considered with "no or little"
-          extra effort.
-    \item However, by parameterising selfing and cloning as we have, we are assuming that these properties of
-          populations at given times can be specified independently from the genetics.
-          In other words, mutations that cause selfing to "come and go" are not considered.
-          However, consuming software is free to ignore these parameters and model the genetics within
-          the framework of a soft selection model.
-    \item We don't view these as limitations. Rather, this is what most people are doing most of the time.
-          Just now we can finally try to do this part w/fewer errors.
-\end{enumerate}
-
-\aprcomment{If selection is placed on top of the demographic model, this implies
-that we are living in a soft selection world. Though sizes could just as easily
-be interpreted as carrying capacities in a hard selection model.}
-\krtcomment{We are firmly in the soft selection world here.  The connection to hard selection
-requires some other strong assumptions that will sometimes be met, and sometimes not.
-For example, if the mean fitnesses are approximately equal across demes, then a hard selection
-model could be similar to a soft selection model.  Relevant literature to dig up: Felsenstein's review,
-which will lead to other stuff.}
-
-\jkcomment{More detail here. Also, why only selfing and cloning, why not selection
-coefficients and whatnot?}
-\krtcomment{Selection is a can of worms: scaling of diploid genotype fitnesses will differ between tools.
-The possibility of "GxE" ("different s for same genotype in different demes") can lead to quadratic complexity.}
+\mhcomment{If requested
+I can quickly draw up a schematic diagram to illustrate these partitions.}
 
@@
 
-This model is, by design, very limited. We do not specify any details of
+This model is, by design, very limited. We do not specify any further details of
 individuals or of the processes that happen \emph{within} populations,
@@
 This narrow focus on individuals and their groupings into populations
-should ensure that the Demes standard is stable, and not subject to
+should ensure that the \Demes\ standard is stable, and not subject to
 continual development as more and more elaborations of basic processes
@@
 
+\section{Extended data and figures}
+\label{sec:appendix-figures}
+
 \begin{figure}[h!]

### 2022-04-08 paper.tex
changes from kevin's comments

@@
 citations? This is critical - we want to establish that Demes is the
-"middleware" between the inference methods and the simulators, and captures the
+``middleware'' between the inference methods and the simulators, and captures the
 basic assumptions made by them all.}
@@
 either according to their ancestor-descendant relationships
-or through processes of continuous or pulse migrations.
+or through processes involving migrations.
 Few other properties of the populations are specified in the model:
@@
 described in \cite[][chapter 3]{wakeley2008-hd}.
-Further, this model assumes "soft" selection \citep{christiansen1975hard},
+Further, this model assumes ``soft'' selection \citep{christiansen1975hard},
 meaning that the dynamics of population sizes
@@
     Just now we can finally try to do this part w/fewer errors.''
+}
+\krtcomment{
+    I think we want a section in the Discussion outlining some obvious things
+    that people can do w/demes that are sitting "just outside" the strict
+    specification: hard selection, etc..
 }

### 2022-04-08 paper.tex
add hudson and tajima refs

@@
 and a backwards-time (coalescent) simulation follows the approximations
-described in \cite[][chapter 3]{wakeley2008-hd}.
+described in \citet{tajima1983evolutionary,hudson1983testing}
+and \citet[][chapter 3]{wakeley2008-hd}.
 Further, this model assumes ``soft'' selection \citep{christiansen1975hard},

### 2022-04-09 paper.tex
add subsections
address graham's comments

@@
 Since the schema is definitive,
-\aprcomment{what exactly do we mean by ``definitive''?}
 we will not recapitulate the details
@@
 
-Below we detaail the population genetics model in 
+Below we detail the population genetics model in 
 Appendix~\ref{sec:appendix-pop-gen-model},
@@
 intended scope of Demes in Appendix~\ref{sec:appendix-scope}.
-The extended figures in Appendix~\ref{sec:appendix-figures} provides
+The extended figures in Appendix~\ref{sec:appendix-figures} provide
 additional examples and illustrations of Demes usage.
@@
 
+\subsection{Time units}
+
 Population and event times are written as units in the past, so that time zero
@@
 permitted, such as years, accompanied by the generation time.
+
+\subsection{Sizes and epochs}
 
@@
 
+\subsection{Population dynamics}
+
 Within a population, we implicitly assume that allele frequency dynamics can
@@
 Briefly, generations are non-overlapping (all parents
-reproduce and die simultaneously) and the expected frequency (at birth)
-of the $i^{th}$ allele currently at frequency $p_i$
-one generation in the next generation is $p_iw_i/\bar{w}$,
+reproduce and die simultaneously), and for allele $i$ currently at frequency $p_i$,
+its frequency in the next generation (at birth) is expected to be $p_iw_i/\bar{w}$,
 where $w_i$ and $\bar{w}$ are the marginal and
@@
 % }
+
+\subsection{Selfing and cloning}
 
@@
 I can quickly draw up a schematic diagram to illustrate these partitions.}
+
+\subsection{Relationships between populations}
 

### 2022-04-11 paper.tex
implement suggestions and cleanup

@@
 
+\krtcomment{
+    I think we want a section in the Discussion outlining some obvious things
+    that people can do w/demes that are sitting "just outside" the strict
+    specification: hard selection, etc..
+}
+
 \bibliographystyle{genetics}
@@
 movement of individuals between those populations.
-\aprcomment{previous sentence changed to allow for selfing/cloning}
 
@@
 
-Within a population, we implicitly assume that allele frequency dynamics can
+Within a population, we assume that allele frequency dynamics can
 be described by the Wright-Fisher model.
@@
 changes are \textit{independent} of the details of individual fitnesses.
-% In the \texttt{Discussion}, we discuss alternative models compatible
-% with this framework and outline the expectations for tools implementing
-% them.
-\aprcomment{As such, this model excludes scenarios such as ``hard selection,''
+As such, this model excludes scenarios such as ``hard selection,''
 in which population sizes are dependent on a population's mean fitness, or
-stochastic fluctuations in population size, such as defining ``carrying capacities''.}
+stochastic fluctuations in population size, such as interpreting
+population sizes as ``carrying capacities''.
 Many forwards and backwards time simulators currently implement this model
-\citep[e.g.,][\aprcomment{and more\dots?}]{hudson2002generating,gutenkunst2009inferring,excoffier2011fastsimcoal,jouganous2017inferring,haller2019slim,thornton2019-nu}.
+\citep[e.g.,][]{hudson2002generating,gutenkunst2009inferring,excoffier2011fastsimcoal,kelleher2016efficient,jouganous2017inferring,haller2019slim,thornton2019-nu}.
 Each has their own unique method of specifying the details of selection and demography.
@@
 is $(1-\sigma)(S + (1-S)/N)$ instead of $(1-\sigma)S$ (so that even with $\sigma=0$
-and $S=0$, selfing may still occur with probability $1/N$).
+and $S=0$, selfing may still occur with probability $1/N$), although
+this effect is negligible in large populations \citep{nordborg1997coalescent}
 Simulation software that allow variable rates of selfing are expected to clearly
@@
 an epoch are not considered.
-\aprcomment{
-    Do we still want to say something along these lines?
-    ``However, consuming software is free to ignore these parameters
-    and model the genetics within
-    the framework of a soft selection model.
-    We don't view these as limitations. Rather, this is what most people
-    are doing most of the time.
-    Just now we can finally try to do this part w/fewer errors.''
-}
-\krtcomment{
-    I think we want a section in the Discussion outlining some obvious things
-    that people can do w/demes that are sitting "just outside" the strict
-    specification: hard selection, etc..
-}
 More details of the mathematical properties of selfing and cloning rates

### 2022-04-12 paper.tex

@@
 
+\aprcomment{Note: Need to standardize use of ``demes'' vs ``Demes'' vs \Demes.}
 \section*{Introduction}
@@
 selection coefficients, and so on) and so we clearly demarcate the
-% FIXME Not quite true here, we're including selfing and cloning rate
 role of the specification by including \emph{none} of them.
@@
 used YAML format~\citep{ben2009yaml} as the recommended means of interchanging
-Demes models (see Figure~\ref{fig-example} for examples). YAML is a data
-serialisation language with an emphasis on simplicity, and interoperates well
+Demes models (e.g., Figures~\ref{fig:IM}~and~\ref{fig:tennessen}). YAML is a data
+serialisation language with an emphasis on simplicity and which interoperates well
 with JSON (indeed, YAML 1.2 is a superset of JSON). We chose YAML over JSON
@@
 mechanism for specifying default values that are inherited hierarchically.
-Figure~\ref{fig-example} shows an example Demes model expressed in both the
+Figure~\ref{fig:IM} shows an example Demes model expressed in both the
 high and low-level forms (both in terms of YAML syntax). For values that repeat
@@
 initial size.
-
-\aprcomment{not sure if this is sufficient here}
-
-\aprcomment{Need some information about defaults}
 
@@
 a standard software entity as part of the specification (the parser),
-which performs this task, and which can be shared by programs that
+which performs this task and can be shared by programs that
 support demes as an input. By having relatively few high-quality Demes
@@
 which translate the high-level specifications into unambiguous low-level
-representations where all parameters have been assigned values and data
+representations where values have been assigned to all parameters and data
 constraints have been checked. The output is formally defined as JSON, but in
@@
 providing a solid foundation for the software ecosystem.
-
-\aprcomment{wrapped below section about parsers to here}
 

### 2022-04-13 paper.tex
edits to the scope section in appendix

@@
 stochastic fluctuations in population size, such as interpreting
-population sizes as ``carrying capacities''.
+population sizes as carrying capacities.
 Many forwards and backwards time simulators currently implement this model
@@
 
-\jkcomment{This section is very rough.}
-
 % https://github.com/popsim-consortium/demes-paper/issues/3
 
-The Demes specification is static by design---we wish to
-unambiguously describe a demographic model with a concrete set
+The Demes specification is static by design--we wish to
+unambiguously describe a demographic model defined by a concrete set
 of parameters. This simplicity means that we cannot directly
@@
 for those parameters. While it is not difficult to imagine extending
-the specification in ways that would allow this, it's not immediately
-clear that the benefits are worth the increased parser complexity.
+the specification in ways that would allow this, this would greatly
+increase parser complexity and require limiting choices to be made.
+Instead, we have opted to restrict Demes to static model descriptions,
+although allowing parameterized distributions or parameter ranges 
+is an direction of possible future work.
 
-\krtcomment{Aaron has done a bit of this non-static stuff in moments.  It may be useful to look over there?}
+%\krtcomment{
+%    Aaron has done a bit of this non-static stuff in moments.
+%    It may be useful to look over there?
+%}
 
-\ggcomment{
-Possible reasons to advocate for non-static features are as follows.
-\begin{enumerate}
-\item Specifying prior distributions of parameters for an ABC.
-\item Specifying upper and lower bounds of parameters and inter-parameter
-      constraints for input to an optimisation-based inference.
-\item Specifying joint distribution, or confidence intervals, for the output of
-      an inference procedure.
-\item Simulating from some arbitrary joint distribution that was inferred
-      previously. E.g. using an ABC posterior distribution of models to
-      train a neural network.
-\item Extending someone else's static model to reinfer parameters,
-      or add new parameters.
-\end{enumerate}
-}
+%\ggcomment{
+%Possible reasons to advocate for non-static features are as follows.
+%\begin{enumerate}
+%\item Specifying prior distributions of parameters for an ABC.
+%\item Specifying upper and lower bounds of parameters and inter-parameter
+%      constraints for input to an optimisation-based inference.
+%\item Specifying joint distribution, or confidence intervals, for the output of
+%      an inference procedure.
+%\item Simulating from some arbitrary joint distribution that was inferred
+%      previously. E.g. using an ABC posterior distribution of models to
+%      train a neural network.
+%\item Extending someone else's static model to reinfer parameters,
+%      or add new parameters.
+%\end{enumerate}
+%}
 
@@
 For example, consider a one-population model with piecewise-constant sizes over
-20 epochs which has $\sim$~40 free parameters---the \texttt{start\_size} and
-\texttt{end\_time} values for each epoch. Supposing we sample 50,000 models
+20 epochs which has $\sim$~40 free parameters: the \texttt{start\_size} and
+\texttt{end\_time} values for each epoch. If we sample 50,000 models
 from the posterior distribution, the resulting multi-document YAML file is
@@
 Other inference procedures based on optimising a loss function
-\citep{gutenkunst2009inferring,kamm2017efficient,jouganous2017inferring}
-(\ggcomment{more refs? MCMC? neural networks?})
+\citep{gutenkunst2009inferring,kamm2017efficient,jouganous2017inferring,ragsdale2019models,excoffier2021fastsimcoal2}
+%(\ggcomment{more refs? MCMC? neural networks?})
 need users to specify parameter bounds,
-and possibly non-linear constraints between parameters.
+and possibly non-linear or conditional constraints between parameters.
 Indeed, the choice of how to parameterise a model could be important for
@@
 
-Implementing the many distributions of interest, and supporting a general
-way to describe a model's free parameters, would greatly increase the
+Implementing the many distributions of interest and supporting a general
+way to describe a model's free parameters would greatly increase the
 complexity of parsers, with relatively limited benefit to most users.
@@
 functions in an existing programming language, for example
-using the Demes Python API.
-To modify an existing static Demes model, perhaps to infer new parameters,
-this would unfortunately require reimplementing the model.
-However, there exist many templating solutions for YAML and JSON that are
-specifically designed for extending static data in arbitrarily complex
-(\ggcomment{refs: YTT, Jsonnet, CUE, Dhall}),
-offering intriguing possibilities.
+using the Demes Python API (e.g., as implemented in \moments\
+\citep{jouganous2017inferring,ragsdale2019models}).
+%To modify an existing static Demes model, perhaps to infer new parameters,
+%this would unfortunately require reimplementing the model.
+As an intriguing possibility for developments in this direction,
+there exist many templating solutions for YAML and JSON that are
+specifically designed for extending static data in arbitrarily complex ways
+(e.g., \texttt{YTT}, \texttt{Jsonnet}, \texttt{CUE}, and \texttt{Dhall}).
+%\citep{ytt,jsonnet,cue,dhall}.
+%(\ggcomment{refs: YTT, Jsonnet, CUE, Dhall}).
 
@@
 omitted. Most notably, mutation and recombination rates are left unspecified
-(though they can be encoded as metadata in the data structure). Genome ploidy
-is left ambiguous, and selection and dominance models are absent. Using a Demes
-model, which specifies deterministic population sizes, as the demographic input
-for simulation using software that allows for selection implies that selection
-is \emph{soft}, that is, genome fitnesses are rescaled relative to the mean
-population fitness \citep{christiansen1975hard}. It is possible to instead
-interpret population sizes as carrying capacities, allowing for a model of
-\emph{hard} selection. However, this should not be understood as the default
-interpretation of population sizes, which conventionally specifies a
-deterministic number of individuals. Finally, as Demes assumes exchangeability
-of individuals within populations, it doesn't provide a way to define geographic
-spatial structure within populations.
+(though they can be encoded as metadata in the data structure), as are
+sequence length or annotations or genomic features.
+Genome ploidy is left ambiguous, and selection and dominance models are absent. 
+We also omit information about the sampling of individuals, as we view
+the demographic history as independent of the sampling of individuals.
+It is therefore left up to simulation and inference tools how to define
+genomic parameters and sampling strategies, as each software will have its
+own set of constraints and relevant input parameters.
 
+In conclusion,
 Demes version 1.0 is intended to provide a stable and minimal specification

### 2022-04-19 paper.tex
small fixup

@@
 (though they can be encoded as metadata in the data structure), as are
-sequence length or annotations or genomic features.
-Genome ploidy is left ambiguous, and selection and dominance models are absent. 
+sequence length, annotations and other genomic features, and individual ploidy.
+Selection and dominance models are absent, as discussed in
+Sec.~\ref{sec:appendix-pop-gen-model}.
 We also omit information about the sampling of individuals, as we view

### 2022-04-20 paper.tex
address comments and edits throughout

@@
 
-\aprcomment{Note: Need to standardize use of ``demes'' vs ``Demes'' vs \Demes.}
 \section*{Introduction}
@@
 Software to either simulate whole genome
-sequences~\citep{thornton2014cpp,thornton2019-nu,staab2015scrm,baumdicker2021-iu,kelleher2016efficient,haller2019slim}
+sequences~\citep{thornton2014cpp,thornton2019-nu,staab2015scrm,
+baumdicker2021-iu,kelleher2016efficient,haller2019slim}
 or informative summary statistics of
@@
 and occasionally impossible because of missing information. 
-\mhcomment{Maybe add a short, concrete example to convince readers of the problem?}
-%\aprcomment{This feels tangential - maybe discussion}
-%The \stdpopsim\ project is a community-maintained catalog of species
-%information and demographic models designed to provide standardised
-%population genetic simulations~\citep{adrion2020community}.
-%The development process for adding demographic models to
-%\stdpopsim\ tries to avoid errors in model implementation by requiring two
-%independent implementations of each model
-%to be equivalent. While this process is
-%robust and has uncovered both significant
-%errors~\citep{ragsdale2020lessons} and ambiguities in
-%published models, it is labour intensive.
+% \mhcomment{Maybe add a short, concrete example to convince readers of the problem?}
+% APR: I think this is ok here.
 Instead, it would be far better if published results were reported
@@
 using either a command line
-interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
+interface~\citep[e.g.,][]{hudson2002generating,kern2016discoal},
 a custom input file
-format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
+format~\citep[e.g.,][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
 or an Application Programming Interface (API) to allow
-models be defined programmatically~\citep[e.g.][]{
-thornton2014cpp,thornton2019-nu,baumdicker2021-iu,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
+models be defined programmatically~\citep[e.g.,][]{
+thornton2014cpp,thornton2019-nu,baumdicker2021-iu,kelleher2016efficient,
+becheler2019quetzal,haller2019slim}.
 Command line interfaces are a concise way of expressing
@@
 is used by a number of different
-simulators~\citep[e.g.][]{ewing2010msms,chen2009fast,staab2015scrm},
+simulators~\citep[e.g.,][]{ewing2010msms,chen2009fast,staab2015scrm},
 and the POPDemog visualisation software~\citep{zhou2018popdemog}.
@@
 not appropriate as an interchange standard because of the dependency on both
-the programming language involved,
+the programming language involved
 and the specific details of the simulator in question.
@@
         We have included software infrastructure developed
-        for working with Demes models (parsing, validation,
-        visualisation, etc) as well as all known downstream 
-        software that implements the specification,
+        for working with Demes models (such as parsing, validation,
+        and visualisation) as well as downstream 
+        software that implement the specification,
         at the time of writing.
@@
 relationships (i.e., the topology) or by continuous or discrete migration events.
-The deomgraphic model itself omits information about genome biology, such as
+The demographic model itself omits information about genome biology, such as
 genome size and architecture, ploidy, and mutation and recombination rates.
@@
 interest, such as continuous spatial structure
-\aprcomment{citations here, and revisit} or
-\aprcomment{stochastic carrying-capacity models, or something like that},
+\citep{wright1943isolation,barton2002neutral,ringbauer2017inferring,battey2020space}
+or stochastic population size dynamics
+\citep{nunney1993assessing,orr2014population},
 many current forwards- and backwards-in-time simulation software
@@
 widely-used simulators.
-\jkcomment{I guess these are also the basic assumptions also shared by
-inference methods, which is important. Mabye another sentence here making this
-explicit, and listing a few different types of inference methods and giving
-citations? This is critical - we want to establish that Demes is the
-``middleware'' between the inference methods and the simulators, and captures the
-basic assumptions made by them all.}
+The basic assumptions of discrete populations connected by instantaneous or continuous
+migrations are also shared by many inference methods
+\citep[e.g.,][]{gutenkunst2009inferring,gravel2012population,kamm2017efficient,
+jouganous2017inferring,ragsdale2019models,excoffier2021fastsimcoal2}.
+Demes therefore serves as ``middleware'' between inference methods and simulation
+software, capturing common assumptions shared across implemented software.
 In Appendix~\ref{sec:appendix-pop-gen-model}, we expand on these concepts
@@
 
-\aprcomment{redundancies with Appendix - that could be expanded and reworked a bit}
-
 \subsection*{The scope of Demes}
@@
 clear that the benefits are worth the increased parser complexity.
+
+\aprcomment{I think this belongs here, just one or two sentences:}
+
+\krtcomment{Here, we should say something about other models that people may wish to implement within the demes
+framework.}
+
+\krtcomment{
+    I think we want a section in the Discussion outlining some obvious things
+    that people can do w/demes that are sitting "just outside" the strict
+    specification: hard selection, etc..
+}
 
@@
         specified as a YAML in Demes format (Figure~\ref{fig:tennessen}), we
-        use \demesdraw\ to visualize the demographic model.
-        We then use \msprime\ to simulate genomic data for 20 genome copies
-        sampled from the two contemporary populations, and we use \moments\
+        used \demesdraw\ to visualize the demographic model (note the recent
+            exponential growth resulting in present-day population sizes that
+        greatly exceed those in the past).
+        We then used \msprime\ to simulate genomic data for 20 genome copies
+        sampled from the two contemporary populations, and we used \moments\
         to compute the expected joint site-frequency spectrum for the same
         sample sizes (Figure~\ref{fig:tennessen-simulation}).
-        (B, C) We compare the single-population SFS in each population, showing
+        (B, C) We compared the single-population SFS in each population, showing
         agreement between the simulation methods.
@@
         given in Figure~\ref{fig:tennessen-simulation}.
-        \aprcomment{There may be a better model to use here -- the massive
-        exponential growth really makes the rest of the history hard to see.}
     }
@@
 inference
-\citep{bustamante2001directional,gutenkunst2009inferring,tennessen2012evolution,jouganous2017inferring,kamm2017efficient,kim2017inference}.
-The Demes model (Figure~\ref{fig:showcase}A~and~\ref{fig:tennessen}) was
+\citep{bustamante2001directional,gutenkunst2009inferring,tennessen2012evolution,
+jouganous2017inferring,kamm2017efficient,kim2017inference}.
+The Demes model (Figures~\ref{fig:showcase}A~and~\ref{fig:tennessen}) was
 provided as the input demography to \msprime\ \citep{baumdicker2021-iu} to
-simulate a large recombining region with the mutation rate assumed in
+simulate a large recombining region under the mutation rate assumed in
 \citet{tennessen2012evolution}, and we computed the observed SFS using \tskit\
@@
 for population genetic simulations has long been acknowledged
-~\citep[][e.g.]{antao2007modeler4simcoal2}, % jk - any older refs for this?
+~\citep[e.g.,][]{antao2007modeler4simcoal2}, % jk - any older refs for this?
 and a number of different methods have been proposed to
@@
 interface either to a built-in simulation
-engine~\citep{thornton2014cpp,thornton2019-nu,baumdicker2021-iu,kelleher2016efficient,haller2017flexible}
+engine~\citep{thornton2014cpp,thornton2019-nu,baumdicker2021-iu,
+kelleher2016efficient,haller2017flexible}
 % jk: are/were there other simulation frontends out there?
@@
 Reproducibility is a significant problem throughout the
-sciences~\citep{baker20161} and various measures have been
+sciences~\citep{baker20161}, and various measures have been
 proposed to increase the likelihood of researchers being
@@
 repository in Demes format.
-
-\krtcomment{Here, we should say something about other models that people may wish to implement within the demes
-framework.}
-
-\krtcomment{
-    I think we want a section in the Discussion outlining some obvious things
-    that people can do w/demes that are sitting "just outside" the strict
-    specification: hard selection, etc..
-}
 
@@
 which performs this task and can be shared by programs that
-support demes as an input. By having relatively few high-quality Demes
+support Demes as an input. By having relatively few high-quality Demes
 parsers available as libraries (ideally, one per programming language),

### 2022-04-21 paper.tex
add barton et al 2010
misc edits

@@
 
-\title{Demes: a specification for interchanging complex demography}
+\title{Demes: a standard format for demographic models}
 % First authors
@@
 key goal in population genetics, and with improving methods
-and data, ever more complex models are being estimated.
+and data, ever more complex models are being constructed.
 Demographic models of current interest
@@
 to alleviate these issues. Demes provides a well-defined and unambiguous
-graphical model of populations and their properties that is straightforward to
+model of populations and their properties that is straightforward to
 implement in software, and a text file format that is designed for
@@
 methods.
+
+\ggcomment{We don't showcase Demes output from any inference software.}
 \end{abstract}
@@
 using either a command line
-interface~\citep[e.g.,][]{hudson2002generating,kern2016discoal},
+interface~\citep[e.g.,][]{hudson2002generating,hernandez_flexible_2008,kern2016discoal},
 a custom input file
@@
 or an Application Programming Interface (API) to allow
-models be defined programmatically~\citep[e.g.,][]{
+models to be defined programmatically~\citep[e.g.,][]{
 thornton2014cpp,thornton2019-nu,baumdicker2021-iu,kelleher2016efficient,
-becheler2019quetzal,haller2019slim}.
+becheler2019quetzal,haller2019slim,hernandez_sfs_code_2015}.
 Command line interfaces are a concise way of expressing
@@
 interest, such as continuous spatial structure
-\citep{wright1943isolation,barton2002neutral,ringbauer2017inferring,battey2020space}
+\citep{wright1943isolation,barton2002neutral,barton2010new,
+ringbauer2017inferring,battey2020space}
 or stochastic population size dynamics
@@
 
+\ggcomment{TODO: deduplicate this section and Appendix~\ref{sec:appendix-parsers}}
+
 While the high-level specification is designed for human readability and
@@
 Translation from the high-level specification to the low-level data model
-requires resolving heierarchically-defined default values and verifying
+requires resolving hierarchically-defined default values and verifying
 relationships between populations and the validity of specified parameter
@@
 
-The Demes specification is static by design -- we wish to unambiguously
+\ggcomment{TODO: deduplicate this section and Appendix~\ref{sec:appendix-scope}}
+
+The Demes specification is static by design---we wish to unambiguously
 describe a demographic model with a concrete set of parameters.
@@
         Using an inferred demographic model from \citet{tennessen2012evolution}
-        specified as a YAML in Demes format (Figure~\ref{fig:tennessen}), we
+        specified as a YAML file in Demes format (Figure~\ref{fig:tennessen}), we
         used \demesdraw\ to visualize the demographic model (note the recent
@@
         agreement between the simulation methods.
-        (D) Code snippets of the interactions between \demes\ and the simulation
+        (D) Python code snippets of the interactions between \demes\ and the simulation
         software. An extended script to compute the SFS shown in (B) and (C) is
@@
 interpretation~\citep{mailund2005coasim,antao2007modeler4simcoal2,
-ewing2010msms,zhou2018popdemog}. However, these methods
+parreira_spams_2009,ewing2010msms,parobek_skelesim_2017,zhou2018popdemog}.
+However, these methods
 currently have little traction as they are all either directly coupled
@@
 engine~\citep{thornton2014cpp,thornton2019-nu,baumdicker2021-iu,
-kelleher2016efficient,haller2017flexible}
+hernandez_sfs_code_2015,kelleher2016efficient,haller2017flexible}
 % jk: are/were there other simulation frontends out there?
@@
 Sizes and mating system details are specified for each population within
-epochs.
+population-specific epochs.
 Epochs are contiguous time intervals that define
@@
 Demes is designed to interface with a large number of simulation and inference
-methods, so we have restricted to the specification to only describe
+methods, so we have restricted the specification to only describe
 demographic features at the population level. As the underlying discrete
@@
     \begin{tcolorbox}
-        \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/tennessen.yml}
+        \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/tennessen.yaml}
     \end{tcolorbox}
@@
         \texttt{msprime.sim\_ancestry()}. The same loaded graph can also be
-        passed to \moments\ to computed the expected joint SFS.
+        passed to \moments\ to compute the expected joint SFS.
         To compare the SFS in Figure~\ref{fig:showcase}, we \emph{marginalize} the

### 2022-04-21 software-table.tex
misc edits

@@
     A Python library for loading, saving, and working with
-    Demes models. Includes conversion utilities for existing
-    demographic model descriptions such as
+    Demes models. Includes support for converting to and from
     \texttt{ms}~\citep{hudson2002generating}
@@
     Summary of what dadi does and how it implements Demes~\citep{gutenkunst2009inferring}.
+    \ggcomment{what is the status of dadi/Demes support?}
     \\
@@
 \fwdpy &
-    Forward-time Wright-Fisher simulation \citep{thornton2014cpp,thornton2019-nu}\\
+    Forward-time Wright-Fisher simulation \citep{thornton2014cpp,thornton2019-nu}.\\
 
@@
     and demographic models specified in Demes
-    format~\citep{adrion2020community}.\\
+    format~\citep{adrion2020community}.
+    \ggcomment{stdpopsim doesn't yet have Demes support}
+    \\
 

### 2022-04-21 paper.tex
changes from review

@@
 key goal in population genetics, and with improving methods
-and data, ever more complex models are being constructed.
+and data, ever more complex models are being proposed and tested.
 Demographic models of current interest
@@
 models to be defined programmatically~\citep[e.g.,][]{
-thornton2014cpp,thornton2019-nu,baumdicker2021-iu,kelleher2016efficient,
-becheler2019quetzal,haller2019slim,hernandez_sfs_code_2015}.
+thornton2014cpp,hernandez_sfs_code_2015,kelleher2016efficient,
+becheler2019quetzal,haller2019slim,thornton2019-nu,baumdicker2021-iu}.
 Command line interfaces are a concise way of expressing
@@
 interface either to a built-in simulation
-engine~\citep{thornton2014cpp,thornton2019-nu,baumdicker2021-iu,
-hernandez_sfs_code_2015,kelleher2016efficient,haller2017flexible}
+engine~\citep{thornton2014cpp,hernandez_sfs_code_2015,kelleher2016efficient,
+haller2017flexible,thornton2019-nu,baumdicker2021-iu}
 % jk: are/were there other simulation frontends out there?

### 2022-04-21 fig/minted-snippet.tex
snippet tex file

\documentclass{article}
\usepackage{fullpage}
\usepackage{minted}
\usepackage{tcolorbox}
\usepackage{etoolbox}
\BeforeBeginEnvironment{minted}{\begin{tcolorbox}}%
\AfterEndEnvironment{minted}{\end{tcolorbox}}%

\begin{document}

\begin{minted}[numbersep=5pt]{python}
import demes, msprime, moments
# load the demographic model and specify sample sizes
model = demes.load("tennessen.yml")
samples = {"AFR": 20, "EUR": 20}
# import model as msprime demography and simulate ancestry
demography = msprime.Demography.from_demes(model)
ts = msprime.sim_ancestry(
    [msprime.SampleSet(n, ploidy=1, population=p) for p, n in samples.items()],
    demography=demography)
# compute the expected joint SFS using moments
sfs = moments.Spectrum.from_demes(model, samples=samples)
\end{minted}

### 2022-04-22 fig/minted-snippet.tex

@@
 # load the demographic model and specify sample sizes
-model = demes.load("tennessen.yml")
+model = demes.load("tennessen.yaml")
 samples = {"AFR": 20, "EUR": 20}

### 2022-04-26 cover-letter/cover-letter.tex
add cover letter draft

\documentclass[11pt]{article}
\usepackage[round]{natbib}

\usepackage{fullpage}
\usepackage{url}

\usepackage{parskip}
\setlength{\parindent}{0em}
\setlength{\parskip}{2ex}
\pagenumbering{gobble}

\usepackage[hang,flushmargin]{footmisc}
\usepackage[document]{ragged2e}

\begin{document}

Editors-in-Chief\hfill\today\\
\emph{Genetics}

Dear Editors:

We are pleased to submit our manuscript “Demes: a standard format for
demographic models” for your consideration for publication in \emph{Genetics}
as a Communications article.

The estimation and simulation of demographic models is central in population
genetics and genomics studies. Realistic demographic models are crucial for
downstream inferences of selection and genome biology, while understanding
species histories is itself a primary goal of the field. The amount of genomic
data and sophistication of inference and simulation tools now allow for the
estimation of increasingly complex demographic models, often involving dozens
of parameters with many interacting components.

However, this growing model complexity has revealed major hurdles to community
goals of replicability, robustness, and communication of
results, e.g.\footnote{Ragsdale, Aaron P., et al. ``Lessons learned from bugs in
models of human history.'' \emph{The American Journal of Human Genetics}
107.4 (2020): 583-588.}. These shortcomings have two primary causes: (1)
simulation and inference tools each have their own format for describing
demographic models, so that errors are common when translating input between
software, and (2) there are no agreed-upon guidelines for reporting demographic
models, so that published results are typically difficult to reconstruct and
are often presented with missing information.

Here, we present a solution to each of these issues. Demes defines a precise
and user-friendly format for demographic model specification, and we include a
suite of tools in multiple widely used programming languages for defining,
manipulating, and validating demographic models. Already, Demes has been
incorporated into popular simulation and inference tools, with growing support
among population genetics simulation developers.

From a user’s perspective, Demes-supported tools greatly reduce programming
burden and the possibility for errors and provide a concise and readable format
for sharing results. In addition to describing Demes, we showcase the
simplicity of specifying a multi-population model, which is used to illustrate
the model, simulate sampled genomes using \emph{msprime}, and compare to
expected patterns of diversity using \emph{moments}. Crucially, this requires
only a single specification of the demographic model, avoiding the need to
implement the model multiple time in different input formats.

We foresee Demes gaining wide use in the field, as it improves replicability of
published demographic inference results, allows interchange between a growing
number of simulation software and programming languages, and reduces potential
for user error and frustration of complex model implementation. Demes provides
a link between multiple popular population genetics simulation and inference
tools, including many that were published in \emph{Genetics} that maintain a
healthy community of users and developers. This project and paper should
therefore be of interest to readers of \emph{Genetics}.

Sincerely,\\
XXX, on behalf of all authors

\end{document}

### 2022-04-27 paper.tex
some edits and de-duplication

@@
 \ggcomment{We don't showcase Demes output from any inference software.}
+\aprcomment{You could say that ``showcase'' also refers to the table that
+includes demes support in inference software, not just the illustration
+in the main text.}
 \end{abstract}
@@
 
-\ggcomment{TODO: deduplicate this section and Appendix~\ref{sec:appendix-parsers}}
-
 While the high-level specification is designed for human readability and
@@
 which is intended to be shared by programs that support Demes as an input.
-By having relatively few high-quality Demes parsers available as libraries
-(ideally one per programming language), we ensure consistency across simulation
+By maintaining high-quality Demes parsers available as libraries,
+we ensure consistency across simulation
 software in their implementation of Demes-specified models.
-We provide a reference implementation written in Python to resolve any
-potential ambiguities and to provide a helpful template for other
-implementations, as well as an extensive test suite of examples and the
-expected outputs.
-In addition, we have high-quality parser implementations in the Python and
-C languages (published under liberal open source licences), providing a
-solid foundation for the software ecosystem (Table~\ref{tab:software}).
+In addition to a reference implementation written in Python,
+we provide high-quality parser implementations in the Python and
+C languages (Table~\ref{tab:software}, and discussed in
+Appendix~\ref{sec:appendix-parsers}).
 
 \subsection*{The scope of Demes}
-
-\ggcomment{TODO: deduplicate this section and Appendix~\ref{sec:appendix-scope}}
 
@@
 clear that the benefits are worth the increased parser complexity.
-
-\aprcomment{I think this belongs here, just one or two sentences:}
-
-\krtcomment{Here, we should say something about other models that people may wish to implement within the demes
-framework.}
-
-\krtcomment{
-    I think we want a section in the Discussion outlining some obvious things
-    that people can do w/demes that are sitting "just outside" the strict
-    specification: hard selection, etc..
-}
+It is important to note, however, that Demes may be used in applications that
+include additional population genetic processes outside of what is explicitly
+modelled in the specification, such as interpreting population sizes as
+carrying capacities, implementations of \emph{hard} selection, or layering more
+complicated mating or spatial structure. The Demes specification therefore
+provides a basic model that can be elaborated on where necessary.
 
@@
 many widely-used population genetic methods, so we have omitted many
-features of the underlying genome biology (as described above, and in
-Appendix~\ref{sec:appendix-scope}).
+features of the underlying genome biology (Appendix~\ref{sec:appendix-scope}).
 
@@
 
-%\krtcomment{The paragraph(s) below subsume(s)/replace(s) some of what appears in later pargraphs.
-%	I'm not making an attempt yet to revise that later content--I think other authors should
-%	weigh in.}
-
-%\krtcomment{Aaron -- I have perhaps over-simplified the description of marginal allele frequency dynamics?}
-
 \subsection{Population dynamics}
@@
 The goal of \Demes\ is to provide a common method for the latter task.
-
-% \krtcomment{
-%     We need to revisit the paragraph(s) below.
-%     We do not assume individuals are exchangeable, because we do not
-% 	require that all individuals have the same fitness.
-%     See, e.g., p 60 of Wakeley for why fitness == exchangeability
-% 	and why that is required for the standard coalescent.
-% }
-% \mhcomment{The `exchangeability' assumption also breaks down under 
-% 	different reproductive modes, since the sampling of the alleles
-%     affects coalescent times. 
-% 	For example, in a highly selfing species the coalescent time for two different alleles 
-% 	within an individual is much lower than two alleles from different individuals.
-%     I've triedto rewrite the below paragraph to update and clarify
-%     the reproduction parameters in line with the discussions on Github.
-%     Finally, 'rates' imply events occur as Poisson-like 
-% 	processes over continuous time, whereas the values used here instead define the 
-% 	\emph{fraction} of reproductions of each type.
-%     Shall we rename to selfing/clonal `fractions',
-% 	or would that mess up the rest of the schema too much?
-% }
 
@@
 More details of the mathematical properties of selfing and cloning rates
-in a coalescent context can be found in \citet{hartfield2016facsexcoal}.
+in a coalescent context can be found in
+\citet{nordborg1997coalescent,hartfield2016facsexcoal}.
 
-\mhcomment{If requested
-I can quickly draw up a schematic diagram to illustrate these partitions.}
+%\mhcomment{If requested
+%I can quickly draw up a schematic diagram to illustrate these partitions.}
 
@@
 
-% https://github.com/popsim-consortium/demes-paper/issues/3
-
-The Demes specification is static by design--we wish to
-unambiguously describe a demographic model defined by a concrete set
-of parameters. This simplicity means that we cannot directly
-specify parameter distributions or estimated confidence intervals
-for those parameters. While it is not difficult to imagine extending
-the specification in ways that would allow this, this would greatly
-increase parser complexity and require limiting choices to be made.
-Instead, we have opted to restrict Demes to static model descriptions,
-although allowing parameterized distributions or parameter ranges 
-is an direction of possible future work.
-
-%\krtcomment{
-%    Aaron has done a bit of this non-static stuff in moments.
-%    It may be useful to look over there?
-%}
-
-%\ggcomment{
-%Possible reasons to advocate for non-static features are as follows.
-%\begin{enumerate}
-%\item Specifying prior distributions of parameters for an ABC.
-%\item Specifying upper and lower bounds of parameters and inter-parameter
-%      constraints for input to an optimisation-based inference.
-%\item Specifying joint distribution, or confidence intervals, for the output of
-%      an inference procedure.
-%\item Simulating from some arbitrary joint distribution that was inferred
-%      previously. E.g. using an ABC posterior distribution of models to
-%      train a neural network.
-%\item Extending someone else's static model to reinfer parameters,
-%      or add new parameters.
-%\end{enumerate}
-%}
+The Demes specification is designed to describe demographic
+models defined by a fixed set of model parameters.
+As described in the main text, it does not include
+information about estimated confidence intervals or the joint distribution
+of parameter values.
+While we have opted to restrict Demes to static model descriptions,
+allowing parameterized distributions or parameter ranges 
+is a direction of possible future work.
 
@@
         but are much more human-readable.
+        \aprcomment{Need figure panel labels.}
     }
@@
     \caption{
-        \textbf{The \citet{tennessen2012evolution} two-population demographic model in Demes format.}
+        \textbf{The \citet{tennessen2012evolution} two-population
+        demographic model in Demes format.}
+        This model includes a single ancestral population that expands in
+        size in the past, followed by divergence between AFR- and EUR-labeled
+        populations. The two-population phase of the model includes multiple
+        epochs of varying size, and rapid exponential growth over the past
+        five thousand years in each population.
     }

### 2022-05-04 paper.tex
Editing pass through main text

@@
 methods.
-
-\ggcomment{We don't showcase Demes output from any inference software.}
-\aprcomment{You could say that ``showcase'' also refers to the table that
-includes demes support in inference software, not just the illustration
-in the main text.}
 \end{abstract}
@@
 
-Simulation is a core tool in population genetics, and a
+Simulation is a core tool in population genetics, and
 many methods have been developed over the past three
@@
 However, this conciseness means that models of even intermediate complexity
-are difficult for humans to understand, making errors highly likely.
+are difficult for humans to understand, making errors likely.
 Input parameter file formats for simulators allow the model specification
@@
 Individuals can move between populations based on their ancestor-descendant
-relationships (i.e., the topology) or by continuous or discrete migration events.
+relationships  or by continuous or discrete migration events.
 The demographic model itself omits information about genome biology, such as
@@
 (see Appendix~\ref{sec:appendix-spec} for details about our choice of
-YAML formatting for demographic models).
+YAML).
 Demographic models written in YAML format provide information about global
@@
 By maintaining high-quality Demes parsers available as libraries,
-we ensure consistency across simulation
-software in their implementation of Demes-specified models.
+we ensure consistency across simulation and inference
+software.
 In addition to a reference implementation written in Python,
@@
 modelled in the specification, such as interpreting population sizes as
-carrying capacities, implementations of \emph{hard} selection, or layering more
+carrying capacities, implementations of hard selection, or layering more
 complicated mating or spatial structure. The Demes specification therefore
 provides a basic model that can be elaborated on where necessary.
-
-Demes is also designed to interface with a large number of simulation and
-inference methods, so we have restricted the specification to only
-describe demographic features at the population level.
-The underlying discrete demographic structure is the common thread connecting
-many widely-used population genetic methods, so we have omitted many
-features of the underlying genome biology (Appendix~\ref{sec:appendix-scope}).
 
@@
 
-Here, we highlight the interface between Demes and other software, including
+Here, we highlight the interaction between Demes and other software, including
 simulation and model illustration tools. Demes allows us to specify a

### 2022-05-05 paper.tex

@@
 This ability to infer and simulate such complex demographic scenarios, however,
-has highlighted a major shortcoming in community standards.
-The fragmented landscape of different approaches to describing demographic
+has highlighted a major shortcoming in community standards:
+the fragmented landscape of different ways to describe demographic
 models makes it difficult to compare inferences made by different methods
 and to reliably simulate from previously inferred models.
-
-Inference methods for demography usually output a set of
-text files which describe the inferred model,
-and nearly all methods use different file formats.
 Inference results are typically reported in publications
@@
 implementing the precise model inferred for later simulation
-is at best tedious and error prone~\citep{ragsdale2020lessons}
+is at best tedious and error prone~\citep{adrion2020community,ragsdale2020lessons},
 and occasionally impossible because of missing information.
@@
 % APR: I think this is ok here.
-Instead, it would be far better if published results were reported
-in an unambiguous standardised format that could be provided
-directly as input to simulators.
+% Instead, it would be far better if published results were reported
+% in an unambiguous standardised format that could be provided
+% directly as input to simulators.
 
@@
 is used by a number of different
-simulators~\citep[e.g.,][]{ewing2010msms,chen2009fast,staab2015scrm},
-and the POPDemog visualisation software~\citep{zhou2018popdemog}.
+simulators~\citep[e.g.,][]{ewing2010msms,chen2009fast,staab2015scrm}.
 However, this conciseness means that models of even intermediate complexity
@@
 to be less terse and allow for documentation in the form of comments.
-However, there is no standard format shared across simulators,
-and the details of the demographic models are not separated
-from other simulation parameters.
-The APIs used by individual simulators to define demographic models are also
-not appropriate as an interchange standard because of the dependency on both
-the programming language involved
-and the specific details of the simulator in question.
+Several Graphical User Interface (GUI) and visualisation
+methods have been developed, which greatly facilitate
+interpretation~\citep{mailund2005coasim,antao2007modeler4simcoal2,
+parreira_spams_2009,ewing2010msms,parobek_skelesim_2017,zhou2018popdemog}.
+However, these methods
+currently have little traction as they are all either directly coupled
+to an internal simulation method or to the  syntax of a specific simulator.
+There is currently no way in which demographic models inferred by
+different packages can be can be simulated or visualised by downstream software.
 
@@
 \section*{Discussion}
-
-The difficulty of describing complex demographic models
-for population genetic simulations has long been acknowledged
-~\citep[e.g.,][]{antao2007modeler4simcoal2}, % jk - any older refs for this?
-and a number of different methods have been proposed to
-make such simulations more accessible and less error prone.
-Several Graphical User Interface (GUI) and visualisation
-methods have been developed, which greatly facilitate
-interpretation~\citep{mailund2005coasim,antao2007modeler4simcoal2,
-parreira_spams_2009,ewing2010msms,parobek_skelesim_2017,zhou2018popdemog}.
-However, these methods
-currently have little traction as they are all either directly coupled
-to an internal simulation method or to the specific syntax
-of a given simulator. Another approach that has been taken to
-simplify running such simulations is to provide an API in a
-scripting language such as Python or R, providing a flexible
-interface either to a built-in simulation
-engine~\citep{thornton2014cpp,hernandez_sfs_code_2015,kelleher2016efficient,
-haller2017flexible,thornton2019-nu,baumdicker2021-iu}
-% jk: are/were there other simulation frontends out there?
-or as a front end for other simulators~\citep{staab2016coala}.
-While APIs have many advantages and have grown in popularity in
-recent years, the demographic model descriptions are tied
-to the details of both the underlying simulation engines
-and the programming language in question.
-
 Stable and healthy software ecosystems require standard interchange

### 2022-05-18 software-table.tex
Remove stdpopsim from software table
Update dadi entry in Table 1

@@
 \dadi &
-    Summary of what dadi does and how it implements Demes~\citep{gutenkunst2009inferring}.
-    \ggcomment{what is the status of dadi/Demes support?}
+       Simulate SFS from Demes models, infer demographic history and distributions of fitness effects~\citep{gutenkunst2009inferring}.
     \\
@@
 
-\stdpopsim &
-    A community-maintained collection of empirical data for species
-    and demographic models specified in Demes
-    format~\citep{adrion2020community}.
-    \ggcomment{stdpopsim doesn't yet have Demes support}
-    \\
-
 \bottomrule

### 2022-05-24 paper.tex
Update dadi entry in Table 1
Pass through main text

@@
 \newcommand{\moments}[0]{\texttt{moments}}
-\newcommand{\dadi}[0]{\texttt{$\partial$a$\partial$i}}
+\newcommand{\dadi}[0]{\texttt{dadi}}
 \newcommand{\fwdpy}[0]{\texttt{fwdpy11}}
@@
 
-\affil[1]{Lundbeck GeoGenetics Centre, Globe Institute, University of Copenhagen}
+\affil[1]{Section for Molecular Ecology and Evolution, Globe Institute, University of Copenhagen}
 \affil[2]{Department of Integrative Biology, University of Wisconsin--Madison}
@@
 representation, in which all relevant values are readily available.
-Thus, Demes is composed of three entities: the high-level, human-readable
-model specification; the low-level, fully-qualified model
-specification; and the parser, which is responsible for transforming the former
+Thus, Demes is composed of three entities:
+the Human Data Model (HDM) designed for human readability;
+the Machine Data Model (MDM) designed for programmatic input and processing;
+and the parser, which is responsible for transforming the former
 into the latter.
@@
 of a multi-population demographic model.
-Detailed descriptions of the population genetics model, parsers, scope, and
+More detailed descriptions of the population genetics model, parsers, scope, and
 additional examples are given in the Appendix.
+Complete technical details of the MDM and HDM, and the responsibilities
+of the parser are provided in the online Demes specification
+(\url{https://popsim-consortium.github.io/demes-spec-docs/}).
 
@@
 \citep{nunney1993assessing,orr2014population},
-many current forwards- and backwards-in-time simulation software
-make these same assumptions, so that Demes is relevant to a range of
-widely-used simulators.
+many current forwards- and backwards-in-time simulators
+make these same assumptions.
 The basic assumptions of discrete populations connected by instantaneous or continuous
@@
 
-\subsection*{High-level specification}
+\subsection*{Human Data Model}
 
-The high-level description of Demes models is focused on efficient human
+The Demes Human Data Model (HDM) is focused on efficient human
 understanding and avoiding errors.
@@
 (e.g., Figures~\ref{fig:showcase}~and~\ref{fig:IM}).
-The high-level Demes specification encourages human understanding by avoiding
+The HDM encourages human understanding by avoiding
 redundancy in the description where possible and by providing a mechanism for
@@
 
-While the high-level specification is designed for human readability and
+While the HDM is designed for human readability and
 conciseness, the underlying data model suitable for software implementation
+(the Machine Data Model, or MDM)
 is redundant and exhaustive.
-Translation from the high-level specification to the low-level data model
+Translation from the HDM to the MDM
 requires resolving hierarchically-defined default values and verifying
@@
 we provide high-quality parser implementations in the Python and
-C languages (Table~\ref{tab:software}, and discussed in
+C languages (Table~\ref{tab:software}, as discussed in
 Appendix~\ref{sec:appendix-parsers}).
@@
 population sizes, reflecting the recent explosive population size increase in
-human populations.
+humans.
 
-We used this model to simulate 20 haploid genome copies from each EUR and AFR
+We used this model to simulate 20 haploid genome copies from EUR and AFR
 at time zero (i.e., present day) to obtain the joint site-frequency spectrum
@@
 and compared to the \msprime\ simulated data (Figure~\ref{fig:showcase}B,C).
-To reiterate, for both \msprime\ and \moments\ simulations, the same
-demographic model was used as input, avoiding the need to specify the complex
-demographic model multiple times using the unique API of each software.
+Figure~\ref{fig:showcase}D shows the code required to run the
+simulations in \msprime\ and \moments, and demonstrates that
+precisely the same input model, without modification, was provided to both packages.
+Such interoperability is a major gain for researchers, which we
+hope will become the expected norm as more packages adopt the Demes
+format.
 
@@
 by simulation and inference methods and be used for reporting results in
-publications, either as supplemental material or uploaded to a data
-repository in Demes format.
+publications, either as supplemental material or uploaded to a data repository.
 

### 2022-05-25 paper.tex

@@
 The data model is based on the ubiquitous JSON~\citep{bray2017javascript}
-standard, and the structure is formally defined using
+standard, and formally defined using
 JSON Schema~\citep{wright2020json}.
-The Demes JSON schema rigorously defines the data model,
-specifying the hierarchical structure and the properties that are allowed in different
-contexts along with their types and permissible ranges.
-This schema, and its accompanying documentation,
-fully describe the entities in the model and their
-relationships and the required behaviour of implementations.
-Since the schema is definitive,
+Along with the schema, full technical details of the
+of the model are provided in the
+online specification document
+(\url{https://popsim-consortium.github.io/demes-spec-docs/}).
+This specification rigorously defines the data model,
+fully describing the entities and their
+relationships, and the required behaviour of implementations.
+Since the online specification is definitive,
 we will not recapitulate the details
@@
 meaning that the dynamics of population sizes
-changes are \textit{independent} of the details of individual fitnesses.
+changes are independent of the details of individual fitnesses.
 As such, this model excludes scenarios such as ``hard selection,''
@@
 Each has their own unique method of specifying the details of selection and demography.
-The goal of \Demes\ is to provide a common method for the latter task.
+The goal of Demes is to provide a common method for the latter task.
 
@@
 have parents drawn at random from the previous generation.
-Depending on the simulation software, this random drawing of parent may occur
+Depending on the simulator, this random drawing of parent may occur
 either with or without replacement. When drawing occurs with replacement, a small
@@
 and $S=0$, selfing may still occur with probability $1/N$), although
-this effect is negligible in large populations \citep{nordborg1997coalescent}
-Simulation software that allow variable rates of selfing are expected to clearly
+this effect is negligible in large populations \citep{nordborg1997coalescent}.
+Simulators that allow variable rates of selfing are expected to clearly
 document the expected behaviour.
@@
 
-Finally, individuals in a population may have parents from a different
+Individuals in a population may have parents from a different
 population through migrations. These can be defined as continuous migration
@@
 
-This model is, by design, very limited. We do not specify any further details of
-individuals or of the processes that happen \emph{within} populations,
-because such details are necessarily complicated, vary greatly by
-application, and any attempt to encompass such a broad range of biological
-processes is doomed to failure. An ``individual'' (i.e., a discrete
-organism) is an abstraction that should have a well defined interpretation
-in most cases.
-This narrow focus on individuals and their groupings into populations
-should ensure that the \Demes\ standard is stable, and not subject to
-continual development as more and more elaborations of basic processes
-are added. We could never hope to fully encompass all the details
-of every possible simulation (including ploidy,
-genome sizes, recombination, mutation and gene conversion rates,
-selection coefficients, and so on) and so we clearly demarcate the
-role of the specification by including \emph{none} of them.
-Such details can be specified by existing mechanisms used by
-simulators. If they vary according to the demography (for example,
-we have different recombination maps across populations), then
-we can unambiguously define these by referring to the population
-identifiers and times defined by the Demes model. Rather than attempting
-to embed this information \emph{in} the demographic model, we
-can define it externally and refer \emph{to} the demography.
+\section{Human Data Model}\label{sec:appendix-spec}
 
-\section{High-level model specification}\label{sec:appendix-spec}
-
-The high-level description of Demes models is focused on the efficiency of
+The Demes Human Data Model (HDM) is focused on the efficiency of
 human understanding and the avoidance of errors. We have adopted the widely
@@
 
-Structurally, the high-level Demes specification encourages human understanding
+Structurally, the HDM encourages human understanding
 by avoiding redundancy in the description where possible and by providing a
@@
 Figure~\ref{fig:IM} shows an example Demes model expressed in both the
-high and low-level forms (both in terms of YAML syntax). For values that repeat
+HDM and Machine Data Model (MDM) forms (both in terms of YAML syntax).
+For values that repeat
 across fields, defaults may be used to implicitly assign default values to
@@
 The implicit value inference and default value propagation required to
-fully resolve the high-level Demes model description is straightforward
+fully resolve a Demes HDM description into the corresponding
+MDM form is straightforward
 to describe, but still requires significant effort to implement in
@@
 The Demes specification precisely defines the required behaviour of parsers,
-which translate the high-level specifications into unambiguous low-level
-representations where values have been assigned to all parameters and data
+which translate a model described in the HDM into the MDM, where values have been
+assigned to all parameters and data
 constraints have been checked. The output is formally defined as JSON, but in
@@
 information about estimated confidence intervals or the joint distribution
-of parameter values.
-While we have opted to restrict Demes to static model descriptions,
-allowing parameterized distributions or parameter ranges
-is a direction of possible future work.
+of parameter values. In this section we describe the rationale
+for this design decision.
 
@@
 methods, so we have restricted the specification to only describe
-demographic features at the population level. As the underlying discrete
-demographic structure is the common thread connecting many widely used
-population genetic methods, many features of underlying genome biology are
-omitted. Most notably, mutation and recombination rates are left unspecified
-(though they can be encoded as metadata in the data structure), as are
-sequence length, annotations and other genomic features, and individual ploidy.
+demographic features at the population level.
+Features of the underlying genome biology are omitted,
+including mutation and recombination rates,
+genome annotations, ploidy, sequence annotations, and so on.
 Selection and dominance models are absent, as discussed in
-Sec.~\ref{sec:appendix-pop-gen-model}.
-We also omit information about the sampling of individuals, as we view
-the demographic history as independent of the sampling of individuals.
-It is therefore left up to simulation and inference tools how to define
-genomic parameters and sampling strategies, as each software will have its
-own set of constraints and relevant input parameters.
+Sec.~\ref{sec:appendix-pop-gen-model}, and we do not
+specify how individuals are sampled.
 
-In conclusion,
-Demes version 1.0 is intended to provide a stable and minimal specification
-that is compatible with a wide range of simulators. To ensure consistency
-across implementations, the specification is versioned and we commit to
-following semantic versioning guidelines.
+The model is, by design, very limited. We do not specify details of
+individuals or of the processes that happen within populations,
+because such details are necessarily complicated, vary greatly by
+application, and attempts to encompass such a broad range of biological
+processes are unlikely to succeed.
+The narrow focus on individuals and their groupings into populations
+should ensure that the Demes standard is stable, and not subject to
+continual development as more and more elaborations of basic processes
+are added. Demes is intended to provide \emph{part} of a
+simulation model specification; other parts of the specification,
+such as sampling strategies, or parameters that vary by time
+or population can \emph{refer to} this well defined and unambigous
+description of the demography.
 
@@
         \label{fig:IM}
-        Example isolation-with-migration Demes model. (A) The compact non-redundant
-        representation expressed using YAML. (B) The same model in the fully-qualified
-        form. (C) A visual representation of the model using \texttt{demesdraw}.
+        Example isolation-with-migration Demes model. (A) The Human Data Model
+        representation expressed using YAML. (B) The same model in the
+        Machine Data Model form.
+        (C) A visual representation of the model using \texttt{demesdraw}.
         The YAML descriptions in (A) and (B) correspond exactly to a JSON description,

### 2022-05-25 software-table.tex
Update software table

@@
 \dadi &
-       Simulate SFS from Demes models, infer demographic history and distributions of fitness effects~\citep{gutenkunst2009inferring}.
+   Optimizes parameters in models of demographic history and distributions of fitness effects using SFS~\citep{gutenkunst2009inferring}. Can simulate SFS from Demes models.
     \\
 
+\demesslim &
+    Loads Demes models into the \slim\ forward simulator \citep{haller2019slim}.\\
+
 \fwdpy &
-    Forward-time Wright-Fisher simulation \citep{thornton2014cpp,thornton2019-nu}.\\
+   Simulates the Wright-Fisher model forward in time \citep{thornton2014cpp,thornton2019-nu}. Demes is the preferred format for specifying a demographic model.\\
 
 \gadma &
-    Demographic inference \citep{noskova2020gadma}.\\
+    Infers models of demographic history \citep{noskova2020gadma}. Outputs Demes models and visualizations. \\
 
 \moments &
-    Compute expected diversity statistics (SFS, LD) and perform
-    demographic inference using summary statistics
-    \citep{jouganous2017inferring,ragsdale2019models}.\\
+    Optimizes parameters in models of demographic history using SFS and linkage disequilibrium statistics
+    \citep{jouganous2017inferring,ragsdale2019models}. Models to be optimized can be specified in Demes.\\
 
 \msprime &
-    Coalescent simulation of Demes models, including command line
-    interface~\citep{kelleher2016efficient,kelleher2020coalescent,baumdicker2021-iu}.\\
+    Simulates population genetic models using tree sequences~\citep{kelleher2016efficient,kelleher2020coalescent,baumdicker2021-iu}. Demographic history models can be specified using Demes.\\
 

### 2022-05-27 paper.tex
Update software table

@@
 \newcommand{\tskit}[0]{\texttt{tskit}}
+\newcommand{\demesslim}[0]{\texttt{demes-slim}}
 
@@
 have enabled the increasing complexity of genomic studies, with several software
-packages capable handling large sample sizes, many interacting populations, and
+packages capable of handling large sample sizes, many interacting populations, and
 deviations from panmictic random-mating assumptions.
@@
 There is currently no way in which demographic models inferred by
-different packages can be can be simulated or visualised by downstream software.
+different packages can be simulated or visualised by downstream software.
 
@@
 the existence interval of the population. Each epoch specifies the population size
-over that interval, which can be a constant value or function defined by start
+over that interval, which can be a constant value or a function defined by start
 and end sizes that must remain positive.

### 2022-05-27 paper.tex
Americanize the English.

@@
 task of translating the model descriptions in published
-work into input suitable for population genetic simulators is labour intensive
+work into input suitable for population genetic simulators is labor intensive
 and error prone.
@@
 arenas2012simulation,yuan2012overview,hoban2012computer}.
-Simulations are based on highly idealised population models,
+Simulations are based on highly idealized population models,
 and one of the key uses of inferred demographic histories
@@
 to be less terse and allow for documentation in the form of comments.
-Several Graphical User Interface (GUI) and visualisation
+Several Graphical User Interface (GUI) and visualization
 methods have been developed, which greatly facilitate
@@
 There is currently no way in which demographic models inferred by
-different packages can be simulated or visualised by downstream software.
+different packages can be simulated or visualized by downstream software.
 
@@
 and machine readability.
-The specification precisely defines the required behaviour of implementations,
+The specification precisely defines the required behavior of implementations,
 ensuring that there is no ambiguity of interpretation, and includes both a
@@
 and C parser implementations, as well as utilities for verification and
-visualisation of Demes models, and has been implemented in several popular
+visualization of Demes models, and has been implemented in several popular
 inference and simulation methods (Table~\ref{tab:software}).
@@
         for working with Demes models (such as parsing, validation,
-        and visualisation) as well as downstream
+        and visualization) as well as downstream
         software that implement the specification,
@@
 include additional population genetic processes outside of what is explicitly
-modelled in the specification, such as interpreting population sizes as
+modeled in the specification, such as interpreting population sizes as
 carrying capacities, implementations of hard selection, or layering more
@@
 and to date the transfer of inferred models to downstream simulations
-has been ad-hoc, and conversions between the many different ways
-of expressing such models is both labour intensive and fraught with errors.
+has been \textit{ad-hoc}, and conversions between the many different ways
+of expressing such models is both labor intensive and fraught with errors.
 The proposed Demes standard is an attempt to bridge this gap
@@
 for reproducibility is that we must be able to state precisely what
-the result in question is. The lack of standardisation in how
+the result in question is. The lack of standardization in how
 complex demographic models are communicated today, and the lack of
@@
 fully describing the entities and their
-relationships, and the required behaviour of implementations.
+relationships, and the required behavior of implementations.
 Since the online specification is definitive,
@@
 Appendix~\ref{sec:appendix-spec}, demographic model parsers and their
-intended behaviour in Appendix~\ref{sec:appendix-parsers}, and the
+intended behavior in Appendix~\ref{sec:appendix-parsers}, and the
 intended scope of Demes in Appendix~\ref{sec:appendix-scope}.
@@
 or ``demes'', understood to be a collection of individuals that can be
-conveniently modelled using a defined set of rules and parameters
+conveniently modeled using a defined set of rules and parameters
 \citep{gilmour_demes_1939,gilmour_deme_1955}.
@@
 the probability that offspring are generated from one generation to the next by either
-self-fertilisation or cloning of an individual.
+self-fertilization or cloning of an individual.
 More specifically, for a given epoch within a population denote the clonal rate by
@@
 Within the sexually-reproduced offspring,
-$S$ are born via self-fertilisation while the rest
+$S$ are born via self-fertilization while the rest
 have parents drawn at random from the previous generation.
@@
 Simulators that allow variable rates of selfing are expected to clearly
-document the expected behaviour.
+document the expected behavior.
 
@@
 Demes models (e.g., Figures~\ref{fig:IM}~and~\ref{fig:tennessen}). YAML is a data
-serialisation language with an emphasis on simplicity and which interoperates well
-with JSON (indeed, YAML 1.2 is a superset of JSON). We chose YAML over JSON
+serialization language with an emphasis on simplicity and which interoperates well
+with JSON (indeed, YAML 1.2 is a super set of JSON). We chose YAML over JSON
 because although JSON is an excellent format for data interchange, it is
@@
 
-The Demes specification precisely defines the required behaviour of parsers,
+The Demes specification precisely defines the required behavior of parsers,
 which translate a model described in the HDM into the MDM, where values have been
@@
 expected outputs. In addition, we have high-quality parser implementations in
-the Python and C languages (published under liberal open source licences),
+the Python and C languages (published under liberal open source licenses),
 providing a solid foundation for the software ecosystem.
@@
 and possibly non-linear or conditional constraints between parameters.
-Indeed, the choice of how to parameterise a model could be important for
+Indeed, the choice of how to parameterize a model could be important for
 some inference methods (e.g. absolute times versus relative times between events).

### 2022-05-30 paper.tex

--- 2022-05-27 paper.tex
@@
 \usepackage{amsmath}
+\usepackage[labelformat=simple]{subcaption}
+
+% Use upper case letters for subfigure captions.
+\renewcommand{\thesubfigure}{\Alph{subfigure}}
 
@@
     \begin{minipage}{0.44\textwidth}
+    \begin{subfigure}{\textwidth}
+    \caption{}
         \begin{tcolorbox}
@@
         \end{tcolorbox}
+    \end{subfigure}
+    \begin{subfigure}{\textwidth}
+    \caption{}
         \begin{tcolorbox}
@@
         \end{tcolorbox}
+    \end{subfigure}
     \end{minipage}\hfill
     \begin{minipage}{0.54\textwidth}
+    \begin{subfigure}{\textwidth}
+    \caption{}
         \begin{tcolorbox}
@@
         \end{tcolorbox}
+    \end{subfigure}
     \end{minipage}\\
@@
         Example isolation-with-migration Demes model. (A) The Human Data Model
-        representation expressed using YAML. (B) The same model in the
-        Machine Data Model form.
-        (C) A visual representation of the model using \texttt{demesdraw}.
-        The YAML descriptions in (A) and (B) correspond exactly to a JSON description,
+        representation expressed using YAML.
+        (B) A visual representation of the model using \texttt{demesdraw}.
+        (C) The same model in the Machine Data Model form.
+        The YAML descriptions in (A) and (C) correspond exactly to a JSON description,
         but are much more human-readable.
-        \aprcomment{Need figure panel labels.}
     }

### 2022-05-30 paper.tex

@@
 standard models to be defined.
-However, by parameterising selfing and cloning as we have, we assume that these
+However, by parameterizing selfing and cloning as we have, we assume that these
 properties of populations can be specified independently from the genetics.
@@
 
-\subsection{Static models, not parameterised models}
+\subsection{Static models, not parameterized models}
 \label{sec:appendix-static}
@@
 in ABC inference~\citep{beaumont2002approximate}.
-Other inference procedures based on optimising a loss function
+Other inference procedures based on optimizing a loss function
 \citep{gutenkunst2009inferring,kamm2017efficient,jouganous2017inferring,ragsdale2019models,excoffier2021fastsimcoal2}
@@
 such as sampling strategies, or parameters that vary by time
-or population can \emph{refer to} this well defined and unambigous
+or population can \emph{refer to} this well defined and unambiguous
 description of the demography.

### 2022-05-30 paper.tex
A number of -> several
Also latexmk broke my build

@@
 We provide thoroughly tested implementations of Demes parsers in Python
-and C, and showcase initial support in a number of simulators and inference
+and C, and showcase initial support in several simulators and inference
 methods.
@@
 demographic models, and the syntax defined by \ms~\citep{hudson2002generating}
-is used by a number of different
+is used by several
 simulators~\citep[e.g.,][]{ewing2010msms,chen2009fast,staab2015scrm}.

### 2022-05-30 paper.tex
Final readover language tweaks

@@
 to be less terse and allow for documentation in the form of comments.
-Several Graphical User Interface (GUI) and visualization
+Several graphical user interfaces  and visualization
 methods have been developed, which greatly facilitate
@@
 by the community, such that users can expect to simulate directly
-from inferred models with little programming effort.
+from inferred models with little to no programming effort.
 
@@
 Demes therefore serves as ``middleware'' between inference methods and simulation
-software, capturing common assumptions shared across implemented software.
-In Appendix~\ref{sec:appendix-pop-gen-model}, we expand on these concepts
-and provide more details on the population genetics models supported by Demes.
+software, capturing these common assumptions.
+Appendix~\ref{sec:appendix-pop-gen-model} provides more details on the
+population genetics models supported by Demes.
 
@@
 specification to perform this task (the parser),
-which is intended to be shared by programs that support Demes as an input.
+which is intended to be shared by programs that support Demes as input.
 By maintaining high-quality Demes parsers available as libraries,

### 2022-05-30 cover-letter/cover-letter.tex
Update cover letter

@@
-\documentclass[11pt]{article}
-\usepackage[round]{natbib}
+\documentclass{letter}
 
-\usepackage{fullpage}
-\usepackage{url}
+\signature{Jerome Kelleher}
 
-\usepackage{parskip}
-\setlength{\parindent}{0em}
-\setlength{\parskip}{2ex}
-\pagenumbering{gobble}
-
-\usepackage[hang,flushmargin]{footmisc}
-\usepackage[document]{ragged2e}
-
+\address{Big Data Institute\\University of Oxford\\UK}
 \begin{document}
 
-Editors-in-Chief\hfill\today\\
-\emph{Genetics}
+\begin{letter}{GENETICS}
 
-Dear Editors:
+\opening{Dear Editors,}
 
-We are pleased to submit our manuscript “Demes: a standard format for
-demographic models” for your consideration for publication in \emph{Genetics}
-as a Communications article.
+I am writing on behalf of my coauthors to submit our manuscript entitled
+\emph{Demes: a standard format for demographic models}
+for consideration for publication in GENETICS
+as a ``Methods, Technology, and Resources'' Communications article.
 
-The estimation and simulation of demographic models is central in population
-genetics and genomics studies. Realistic demographic models are crucial for
-downstream inferences of selection and genome biology, while understanding
-species histories is itself a primary goal of the field. The amount of genomic
-data and sophistication of inference and simulation tools now allow for the
-estimation of increasingly complex demographic models, often involving dozens
-of parameters with many interacting components.
-
-However, this growing model complexity has revealed major hurdles to community
+The estimation and simulation of demographic models is central to population
+genetics and genomics studies.
+With ever-growing datasets and recent improvements in simulation efficiency,
+demographic models of interest can now involve dozens of parameters.
+This growing model complexity has revealed major hurdles to community
 goals of replicability, robustness, and communication of
-results, e.g.\footnote{Ragsdale, Aaron P., et al. ``Lessons learned from bugs in
+results\footnote{Ragsdale, Aaron P., et al. ``Lessons learned from bugs in
 models of human history.'' \emph{The American Journal of Human Genetics}
-107.4 (2020): 583-588.}. These shortcomings have two primary causes: (1)
+107.4 (2020): 583-588.}.
+These shortcomings have two primary causes: (1)
 simulation and inference tools each have their own format for describing
 demographic models, so that errors are common when translating input between
-software, and (2) there are no agreed-upon guidelines for reporting demographic
-models, so that published results are typically difficult to reconstruct and
-are often presented with missing information.
+software; and (2) there are no agreed-upon guidelines for reporting demographic
+models, so that published results are difficult to reconstruct and
+often presented with missing information.
 
-Here, we present a solution to each of these issues. Demes defines a precise
+Here we present a solution to these issues. Demes defines a precise
 and user-friendly format for demographic model specification, and we include a
@@
 incorporated into popular simulation and inference tools, with growing support
-among population genetics simulation developers.
-
-From a user’s perspective, Demes-supported tools greatly reduce programming
-burden and the possibility for errors and provide a concise and readable format
-for sharing results. In addition to describing Demes, we showcase the
-simplicity of specifying a multi-population model, which is used to illustrate
-the model, simulate sampled genomes using \emph{msprime}, and compare to
-expected patterns of diversity using \emph{moments}. Crucially, this requires
-only a single specification of the demographic model, avoiding the need to
-implement the model multiple time in different input formats.
+among population genetics tool developers.
 
 We foresee Demes gaining wide use in the field, as it improves replicability of
-published demographic inference results, allows interchange between a growing
-number of simulation software and programming languages, and reduces potential
-for user error and frustration of complex model implementation. Demes provides
+published demographic inference results, allows interchange between
+simulation and inference packages, and reduces potential
+for user error. Demes provides
 a link between multiple popular population genetics simulation and inference
-tools, including many that were published in \emph{Genetics} that maintain a
-healthy community of users and developers. This project and paper should
-therefore be of interest to readers of \emph{Genetics}.
+tools, including many that were published in GENETICS that maintain a
+healthy community of users and contibutors. We believe that this manuscript
+will therefore
+be of substantial interest to readers of GENETICS.
 
-Sincerely,\\
-XXX, on behalf of all authors
+\closing{Sincerely,}
 
+\end{letter}
 \end{document}

### 2022-06-08 paper.tex
Linkify urls.
Also Americanize software-table.tex.
Added MH grant info

@@
 % or input from a source file using \inputminted
+
+% Import this last to avoid mysterious errors.
+\usepackage[implicit=false]{hyperref}
 
@@
 publications, either as supplemental material or uploaded to a data repository.
+
+\section*{Acknowledgments}
+Graham Gower was supported by a Villum Fonden Young Investigator award to Fernando Racimo (project no. 00025300).
+Ryan Gutenkunst was supported by the National Institute of General Medical Sciences of the National Institutes of Health (R01GM127348 to RNG).
+Matthew Hartfield is supported by a NERC Independent Research Fellowship (NE/R015686/1).
 

### 2022-07-18 response-to-reviewers.tex
Add response to reviewers framework.

% LaTeX rebuttal letter example.
%
% Copyright 2019 Friedemann Zenke, fzenke.net
%
% Based on examples by Dirk Eddelbuettel, Fran and others from
% https://tex.stackexchange.com/questions/2317/latex-style-or-macro-for-detailed-response-to-referee-report
%
% Licensed under cc by-sa 3.0 with attribution required.

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{lipsum} % to generate some filler text
\usepackage{fullpage}

% import Eq and Section references from the main manuscript where needed
% \usepackage{xr}
% \externaldocument{manuscript}

% package needed for optional arguments
\usepackage{xifthen}
% define counters for reviewers and their points
\newcounter{reviewer}
\setcounter{reviewer}{0}
\newcounter{point}[reviewer]
\setcounter{point}{0}

% This refines the format of how the reviewer/point reference will appear.
\renewcommand{\thepoint}{\thereviewer.\arabic{point}}

% command declarations for reviewer points and our responses
\newcommand{\reviewersection}{\stepcounter{reviewer} \bigskip \hrule
                  \section*{Reviewer \thereviewer}}

\newenvironment{point}
   {\refstepcounter{point} \bigskip \noindent {\textbf{Reviewer~Point~\thepoint} } ---\ }
   {\par }

\newcommand{\shortpoint}[1]{\refstepcounter{point}  \bigskip \noindent
    {\textbf{Reviewer~Point~\thepoint} } ---~#1\par }

\newenvironment{reply}
   {\medskip \noindent \begin{sf}\textbf{Reply}:\  }
   {\medskip \end{sf}}

\newcommand{\shortreply}[2][]{\medskip \noindent \begin{sf}\textbf{Reply}:\  #2
    \ifthenelse{\equal{#1}{}}{}{ \hfill \footnotesize (#1)}%
    \medskip \end{sf}}

\begin{document}

\section*{Response to the editor}
% General intro text goes here
Thank you for considering this manuscript for publication.

\subsection*{Specific comments}
% We address your specific comments below:
% % \reviewersection

\textit{
Demes is an important advance to standardizing the implementation of popgen
models, and so a very welcome contribution. The reviewers give a number of very
useful suggestions. I leave it up to you how you proceed with incorporating
changes based on these comments, as you will have more of a sense of how the
comments overlap those from other users, but please do write a response to all
of the comments.
}

\section*{Response to the reviewers}
% General intro text goes here
We thank the reviewers for their close reading of our manuscript and
insightful comments. In the following we address the points raised
in turn.

\reviewersection

\begin{point}
stuff
\end{point}
\begin{reply}
responds
\end{reply}

\reviewersection

\begin{point}
STUFF
\end{point}
\begin{reply}
Response
\end{reply}

\end{document}

### 2022-07-18 response-to-reviewers.tex
Add reviewer 1s comments

@@
 \begin{point}
-stuff
+This manuscript presents the Demes data model and related libraries ecosystem.
+I have been using it for some time and was happy to review the paper as I find
+it really useful, intuitive and well designed (as are a lot of tools developed
+by those authors and the whole community of the popsim consortium and tskit).
+The main usage I have for it at the moment was to switch effortlessly between
+msprime and SLiM using the same demographic model, loved that. But as described
+in the paper, it has a lot of other qualities. Overall, this format is a really
+useful addition to the field and which I hope will be largely adopted.
+
+I don't have much to add to this manuscript other than a few comments that I
+hope will help answer a few questions that other readers could have. My main
+comment below is that an example of the YAML format should be included in the
+main text.
 \end{point}
 \begin{reply}
-responds
+TODO
+\end{reply}
+
+\begin{point}
+I usually like when there is a link to the software directly in the abstract
+ in those kind of papers, it avoids scrolling through it as links are
+sometimes in different places in those type of manuscripts.
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+
+\begin{point}
+I suggest including the example of the YAML format in the main text.
+This is too sad that the center piece of this paper has been relegated to the appendix :(
+Show us how easy it is to understand the model from the YAML format!
+Maybe FigA2 is a bit too long, but FigA1 A and B could work well.
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+
+\begin{point}
+A fair question that could be asked by any potential user could be: Are there
+  versions to this data format? There is no version argument in the YAML file,
+what if the format changes in the future? How would my model format would be
+recognized by the parser? I get that it is designed with stability in mind and
+it is done well, but we've seen formats change for other things through time
+(e.g. in bioinformatics) to adapt to new usage.
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+
+\begin{point}
+There might be the need to specify how the initialization time of the whole
+  model is dealt with. This is kind of straightforward when thinking about
+coalescent models for people used to it. But not for forward-time simulations.
+Especially that in your examples there are no start\_time for the first
+population. This might be hard for some users to wrap their head around.
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+
+\begin{point}
+p2 you discuss the defaults of CLI and input parameter formats but not of
+  APIs. It could be said that APIs require some learning (of both the interface
+and the given programming language for users not used to it). Learning a new
+API each time you change simulator is also time consuming.
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+
+\begin{point}
+p2 "YAML format" you could specify that it is a data serialization language
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+
+\begin{point}
+p12 "time units is in generations" you could precise it is the default value here, if it is?
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+
+\begin{point}
+p12 "such as years, accompanied by the generation time". Might be worth
+  detailing that it is needed as simulators work with generations and not
+years, so there needs to be some kind of conversion.
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+
+\begin{point}
+p12 "function defined by start and end sizes" You should detail here what are
+  the models of population expansion available and what is the default. Linear,
+exponential, etc.?
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+
+\begin{point}
+p12 "S are born via [...]" missing that S is a proportion
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+
+\begin{point}
+p13-14 "Other implicit values [...]" until end of paragraph, this part with
+  the example could be worth integrating into the main text as this is a major
+thing to understand about the data model in my opinion.
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+
+\begin{point}
+Section A3: you could also detail that having common parsers allows for
+  consistent and informative error messages when it comes to missing values or
+issues in formatting.
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+
+\begin{point}
+p15 what's the difference between "genome annotations" and "sequence annotations"?
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+
+\textit{Random thoughts I had while reading the paper, to discuss (or not)}
+
+\begin{point}
+I wondered if such format couldn't be used as a way to increase (spoken)
+  language inclusivity. I guess it would be fairly easy to have dictionaries
+including multiple translations for each keyword of the format in the parser.
+That way you could write your model in Japanese or Spanish and it could still
+be read without issue.
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+
+\begin{point}
+Do you think such plain text format also provide a better substrate for
+versionning demographic models? Could be mentioned.
+\end{point}
+\begin{reply}
+TODO
 \end{reply}
@@
 \begin{point}
-STUFF
+
 \end{point}
 \begin{reply}
-Response
+TODO
+\end{reply}
+
+\begin{point}
+
+\end{point}
+\begin{reply}
+TODO
 \end{reply}

### 2022-07-18 response-to-reviewers.tex
Add in points from Reviewer 3.

@@
 \begin{point}
+In their paper entitled "Demes: a standard format for demographic models",
+Gower et al. propose a new way to describe (historical) demographic models of
+populations.
 
+The aim of Demes is to provide a standardized way to describe demographic
+models to be used by various programs performing genetic simulations, which
+currently have different ways to specify the simulated demography. The
+intention is to facilitate information exchange and avoid translation problems
+when passing from say a simulation software to a parameter estimation program,
+if both would use the same input file to describe a demographic model.
+
+The Demes data model is implemented in a text file using the YAML serialization
+language, which is often used for configuration files or bioinformatics
+pipeline descriptions. YAML uses a convenient human-readable syntax, that is
+relatively intuitive to read. Two formats are proposed: a Human data Model
+(HDM) supposedly easy to read by human users, and a more elaborate Machine Data
+Model (MDM) used for processing. A parser (implemented as Python or C
+libraries) is provided to translate HDM into MDM. Overall, I find this
+initiative highly welcome as it makes a lot of sense to have the same input
+file that could be used by several programs. One could thus use a given
+demography inferred from neutral markers (e.g. dadi) to specify a model using
+selection implemented using a forward in time simulator (e.g. slim). This is
+possible since the specification of demography is separated from the
+specification of the genetic model (selection, mutation, recombination).
+
+I have found the paper clearly written and I have just a few comments,
+regarding the structure of the paper and the Demes implementation itself.
 \end{point}
@@
 \begin{point}
+I have two major concerns. The first one is about the two formats: HDM and MDM.
+Contrary to what the author say, I have found the HDM more confusing to read
+than the MDM, which is much more explicit (see minor comments below). My
+impression is that naive users would more easily create mis-specified models in
+HDM than MDM, as HDM is using a lot of implicit definitions (to avoid
+redundancy). My relative extensive exposure to the feedback of users reporting
+bugs for my own software tells me that users do not carefully check the
+structure of their demographic models, and explicit definitions help them in
+making it right. Related to that, it is unclear if the default format should
+always be the HDM, which would each time need to be parsed into MDM or if
+programs could use the MDM as the default input file format. Having both
+possibilities would be great, but personally I would just remove the HDM
+format, and add some of its nice features to the MDM (e.g. defaults)
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
 
+\begin{point}
+The second concern is about the use of a backward-in-time timescale (zero =
+present) with a forward in time definition of epochs. It leads to strange
+definitions, like a start time set to infinity (as in Fig. A1C), which will be
+difficult to implement in forward in time simulators, as they always need to
+define a burn-in time. I thus found quite confusing that a start\_time is always
+larger than an end\_time. It would make more sense to use a backward approach
+for both the time definition and the demographic definition of epochs (maybe
+begin with the more recent epochs and finish with the more ancient ones).
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+\begin{point}
+p. 12 Population dynamics. I am not sure that it is necessary to assume a
+  specific population dynamics model (e.g. Wright-Fisher model) for a data
+format. Similarly, it is not necessary to define how allele frequencies would
+change in the population upon selection. Some simulators will implement
+selection differently (viability selection, fecundity selection, and selection
+could be implemented before or after migration) and, again, this does not
+belong to a demography specifying data format. I understand that the authors
+wanted to distinguish soft from hard selection, where population size might
+depend on average fitness in the latter mode, but there ae so many ways to
+implement selection that it should be left to the specific programs reading the
+demography.
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+\begin{point}
+- p. 12 selfing and cloning. These two parameters are making more sense for
+  forward than backward simulators, and they thus do not seem to be fully
+generic.
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+\begin{point}
+p. 13: Defaults and inheritance. It is not entirely clear to me what happens
+  when a default parameter is superseded in a given ancestral population. Is
+the new value inherited in the daughter populations or is it the default value
+that applies?
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+\begin{point}
+p. 14. Parsers. It is a good idea to provide HDM to MDM parsers, but I note
+  that each program reading an MDM will also implement its own parser to read
+it and translate it to program specific parameters. Again, it would make more
+sense to me to have a single format and no parser. Rather some parser making
+sure that the demographic model makes sense could be useful: i.e. checking that
+population ancestors have all been defined, or that end\_times are compatible
+with start\_time.
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+\begin{point}
+p. 16: migration section might become hard to read when there are more than a
+  few migration rates to specify. A migration matrix representation would make
+more sense, or a way to simplify notation, by perhaps specifying some template
+for the parameters identifiers and only specify the values: Template: \{source,
+dest, start\_time, end\_time, rate\} And then just entering values: \{ X, Y, 100,
+0, 50, 0.0001\}
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+\begin{point}
+p. 17 : Whole HDM model is looking strange for a naive user, with a lot of
+  implicit parameters probably set at default values. For instance
+is start time infinity by default for ancestral population?
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+\begin{point}
+AFR and EUR have no starting times, which suggests that they are created when
+  their ancestral population ends. But what if AFR and EUR were branching off
+ANC at different times?
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+\begin{point}
+migrations section looks strangely defined. Is symmetric migration
+  implemented by default? Starting time is not specified for first epoch. Does
+it mean the epoch starts with the creation of AFR and EUR? But again what if
+AFR and EUR would have been created at different times?
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+\begin{point}
+It is good to have a standard way of specifying a given demographic scenario,
+  but it would also be good to have a standard way to specify genetic data to
+be simulated, and to link it to the demography.
+\end{point}
+\begin{reply}
+TODO
+\end{reply}
+\begin{point}
+Finally, I see the point of separating a model from the sampling, but in some
+  cases it could be beneficial to mention them in the same file, for instance
+if you would like to simulate data at different time points from the same
+population. Making sample size part of the properties of a deme would make it
+more efficient than specifying it from outside.
 \end{point}

### 2022-07-18 response-to-reviewers.tex

--- 2022-07-18 response-to-reviewers.tex
@@
 TODO
+
 \end{reply}
@@
 \begin{point}
-
+I have two major concerns. The first one is about the two formats: HDM and MDM.
+Contrary to what the author say, I have found the HDM more confusing to read
+than the MDM, which is much more explicit (see minor comments below). My
+impression is that naive users would more easily create mis-specified models in
+HDM than MDM, as HDM is using a lot of implicit definitions (to avoid
+redundancy). My relative extensive exposure to the feedback of users reporting
+bugs for my own software tells me that users do not carefully check the
+structure of their demographic models, and explicit definitions help them in
+making it right. Related to that, it is unclear if the default format should
+always be the HDM, which would each time need to be parsed into MDM or if
+programs could use the MDM as the default input file format. Having both
+possibilities would be great, but personally I would just remove the HDM
+format, and add some of its nice features to the MDM (e.g. defaults)
 \end{point}
@@
 TODO
+
+\end{reply}
+
+\begin{point}
+The second concern is about the use of a backward-in-time timescale (zero =
+present) with a forward in time definition of epochs. It leads to strange
+definitions, like a start time set to infinity (as in Fig. A1C), which will be
+difficult to implement in forward in time simulators, as they always need to
+define a burn-in time. I thus found quite confusing that a start\_time is always
+larger than an end\_time. It would make more sense to use a backward approach
+for both the time definition and the demographic definition of epochs (maybe
+begin with the more recent epochs and finish with the more ancient ones).
+\end{point}
+\begin{reply}
+TODO
+
+\end{reply}
+\begin{point}
+p. 12 Population dynamics. I am not sure that it is necessary to assume a
+  specific population dynamics model (e.g. Wright-Fisher model) for a data
+format. Similarly, it is not necessary to define how allele frequencies would
+change in the population upon selection. Some simulators will implement
+selection differently (viability selection, fecundity selection, and selection
+could be implemented before or after migration) and, again, this does not
+belong to a demography specifying data format. I understand that the authors
+wanted to distinguish soft from hard selection, where population size might
+depend on average fitness in the latter mode, but there ae so many ways to
+implement selection that it should be left to the specific programs reading the
+demography.
+\end{point}
+\begin{reply}
+TODO
+
+\end{reply}
+\begin{point}
+- p. 12 selfing and cloning. These two parameters are making more sense for
+  forward than backward simulators, and they thus do not seem to be fully
+generic.
+\end{point}
+\begin{reply}
+TODO
+
+\end{reply}
+\begin{point}
+p. 13: Defaults and inheritance. It is not entirely clear to me what happens
+  when a default parameter is superseded in a given ancestral population. Is
+the new value inherited in the daughter populations or is it the default value
+that applies?
+\end{point}
+\begin{reply}
+TODO
+
+\end{reply}
+\begin{point}
+p. 14. Parsers. It is a good idea to provide HDM to MDM parsers, but I note
+  that each program reading an MDM will also implement its own parser to read
+it and translate it to program specific parameters. Again, it would make more
+sense to me to have a single format and no parser. Rather some parser making
+sure that the demographic model makes sense could be useful: i.e. checking that
+population ancestors have all been defined, or that end\_times are compatible
+with start\_time.
+\end{point}
+\begin{reply}
+TODO
+
+\end{reply}
+\begin{point}
+p. 16: migration section might become hard to read when there are more than a
+  few migration rates to specify. A migration matrix representation would make
+more sense, or a way to simplify notation, by perhaps specifying some template
+for the parameters identifiers and only specify the values: Template: \{source,
+dest, start\_time, end\_time, rate\} And then just entering values: \{ X, Y, 100,
+0, 50, 0.0001\}
+\end{point}
+\begin{reply}
+TODO
+
+\end{reply}
+\begin{point}
+p. 17 : Whole HDM model is looking strange for a naive user, with a lot of
+  implicit parameters probably set at default values. For instance
+is start time infinity by default for ancestral population?
+\end{point}
+\begin{reply}
+TODO
+
+\end{reply}
+\begin{point}
+AFR and EUR have no starting times, which suggests that they are created when
+  their ancestral population ends. But what if AFR and EUR were branching off
+ANC at different times?
+\end{point}
+\begin{reply}
+TODO
+
+\end{reply}
+\begin{point}
+migrations section looks strangely defined. Is symmetric migration
+  implemented by default? Starting time is not specified for first epoch. Does
+it mean the epoch starts with the creation of AFR and EUR? But again what if
+AFR and EUR would have been created at different times?
+\end{point}
+\begin{reply}
+TODO
+
+\end{reply}
+\begin{point}
+It is good to have a standard way of specifying a given demographic scenario,
+  but it would also be good to have a standard way to specify genetic data to
+be simulated, and to link it to the demography.
+\end{point}
+\begin{reply}
+TODO
+
+\end{reply}
+\begin{point}
+Finally, I see the point of separating a model from the sampling, but in some
+  cases it could be beneficial to mention them in the same file, for instance
+if you would like to simulate data at different time points from the same
+population. Making sample size part of the properties of a deme would make it
+more efficient than specifying it from outside.
+\end{point}
+\begin{reply}
+TODO
+
 \end{reply}

### 2022-07-21 response-to-reviewers.tex
A few responses to reviewers.

@@
 \usepackage{fullpage}
+\usepackage{url}
 
@@
 \begin{reply}
-TODO
+
+We have had an issue open for this for some time (\url{https://github.com/popsim-consortium/demes-spec/issues/20}).
+The current plan is to add a variable for this when the spec hits version 1.0, which is planned to coincide
+with this publication.
+Files omitting a spec version will assume to default to 1.0.
+
 \end{reply}
@@
 \begin{reply}
-TODO
+
+    Random thoughts from KT:
+
+    The forward time case is pretty easy: find the most ancient
+    thing with a finite time (events and deme start times), and
+    that's your reference point.  A burn-in time pushes that backwards.
+
+    This was straightforward in fwdpy11 and KT has some additional tools
+    "this close" to release to automate the process (for the case of integer-sized
+    clicks in time).
+
 \end{reply}
@@
 \begin{reply}
-TODO
+
+The field names in a YAML file are variables, not words \textit{per se}.
+Thus, allowing different inputs for the field names is like having code trying
+to understand different names for the same piece of data.
+While it would be useful to allow users to generate files with variable names
+that are not English words, and in character sets other than Latin, those files would
+require additional conversion into the variable/value format defined by \textit{demes}.
+We feel that such conversion is beyond the scope of this project.
 

### 2022-07-21 response-to-reviewers.tex
fix wording
Eek, that was some bad wording

@@
 We have had an issue open for this for some time (\url{https://github.com/popsim-consortium/demes-spec/issues/20}).
-The current plan is to add a variable for this when the spec hits version 1.0, which is planned to coincide
+The current plan is to add a variable for this when the spec hits version 1.0, which will coincide
 with this publication.
-Files omitting a spec version will assume to default to 1.0.
+Files omitting a spec version will assume a default value of 1.0.
 

### 2022-07-22 paper.tex
Some repsonses to reviewer comments

@@
 \affil[3]{Department of Molecular and Cellular Biology, University of Arizona}
-\affil[4]{Institute of Evolutionary Biology, The University of Edinburgh}
+\affil[4]{Institute of Ecology and Evolution, The University of Edinburgh}
 \affil[5]{Computer Technologies Laboratory, ITMO University}
@@
 methods.
+An introduction to the file format and a detailed specification are available at:\\
+\url{https://popsim-consortium.github.io/demes-spec-docs/}.
 \end{abstract}
@@
 This data model is implemented in the widely used YAML
-format~\citep{ben2009yaml}, which provides a good balance between human
-and machine readability.
+format~\citep{ben2009yaml}, which is a data serialization language that
+provides a good balance between human and machine readability.
 The specification precisely defines the required behavior of implementations,
@@
 specification for time units is in generations, although other time units are
-permitted, such as years, accompanied by the generation time.
+permitted, such as years, accompanied by the generation time
+so that downstream software may convert times into generations as required.
 

### 2022-07-22 response-to-reviewers.tex

@@
 \usepackage{url}
+\usepackage{xcolor}
 
@@
     \medskip \end{sf}}
+
+\newcommand{\ggcomment}[1]{{\textcolor{yellow!60!red}{GG: #1}}}
 
@@
 \begin{reply}
-TODO
+
+We agree that this is a useful practice.
+However, in our case there are several software packages that are directly
+and indirectly related to the Demes format as summarized in Table 1.
+We have instead added a link to the specification documentation in the abstract.
+This documentation includes a tutorial introduction for users to learn
+how to write a Demes YAML file, and a detailed specification that developers
+may consult when implementing Demes support in their own sofware.
 \end{reply}
@@
 \begin{reply}
-TODO
+
+We have updated the text to include this point.
 \end{reply}
@@
 \begin{reply}
-TODO
+
+There is no default value for the time units. This must be specified explicitly.
 \end{reply}
@@
 \begin{reply}
-TODO
 
+We have updated the text to include this point.
 \end{reply}
@@
 
+\ggcomment{
+Well, we *could* include lots more detail, but I think we want to be clear
+that the manuscript isn't intended to completely document everything---that's
+what the spec is for.
+}
 \end{reply}
@@
 
+\ggcomment{These sound like synonyms to me too.
+Can "sequence annotations" be deleted, or is this a distinct thing?}
 \end{reply}
@@
 \begin{reply}
-TODO
 
+While plain-text formats are ideally suited for use with version control
+systems such as git, almost all competing/custom formats are also
+plain text and thus share this property. However, Demes models are more
+clearly separated from code and CLI directives, which may make model
+changes more obvious when reviewing change requests such as git pull requests.
 \end{reply}
@@
 \begin{reply}
-TODO
+
+\ggcomment{Anything here we should clarify in the manuscript?}
+
+We've spent considerable effort trying to anticipate ways in which users
+could create erroneous demographic models. By design the Demes spec
+makes many classes of model misspecification impossible to write,
+requires errors be raised when models are ambiguous,
+and explicitly documents the required behavior for edge cases.
+We honestly believe that the Demes format is easier to write
+and interpret than the status quo.
+However, we agree that there are some aspects of the HDM that will be
+confusing for new users.
+
+The HDM was explicitly designed to be a less-strict version of the MDM,
+and our initial approach was to simply allow the omission of some fields
+when there was an obvious default.
+The result is that an MDM model is a valid HDM model (but not vice versa),
+so users are free to be more explicit in any part of a model.
+
+We provide the following resources to assist users when writing models.
+\begin{itemize}
+\item
+The Demes tutorial
+(\url{https://popsim-consortium.github.io/demes-spec-docs/main/tutorial.html})
+is a gentle introduction to writing Demes models.
+It covers 99\% of anticipated uses and assumes very little about the
+reader's knowledge of population genetics.
+\item
+The C and Python parsers have command line interfaces that take HDM as
+input and can output an MDM.
+So if a user is in doubt about the interpretation of some aspect of the HDM,
+then they can easily check how their model will be interpreted.
+\item
+Demes models can be visualised using DemesDraw
+(\url{https://github.com/grahamgower/demesdraw}),
+which provides a sketch of how demes are related and their relative sizes.
+\end{itemize}
+
+Finally, there is no requirement for simulation software to include
+an HDM to MDM parser. As an example, the demes-slim
+(\url{https://github.com/grahamgower/demes-slim/})
+converter only accepts MDM models (and only in JSON format).
+It thus relies on users converting HDM models using an
+external parser, which is a small inconvenience for a large gain
+in interoperability.
 
@@
 \begin{reply}
-TODO
+
+\ggcomment{Anything here we should clarify in the manuscript?}
+
+We apreciate that the timescale we've chosen may at first be jarring
+because it uses backwards-time values with forwards-time language.
+But defining the timescale was fundamental choice that we didn't take lightly.
+All decisions about Demes are publicly recored in github issues,
+pull requests, and code experiments. Two discussions relevant to the
+timescale are
+\url{https://github.com/popsim-consortium/demes-python/issues/2}
+and
+\url{https://github.com/popsim-consortium/demes-python/issues/11}.
+
+To summarise:
+\begin{itemize}
+\item
+"backwards time" language is confusing when talking about migrations
+(does one mean migrating individuals, or coalescent lineages moving
+between demes?)
+\item
+Anchoring time in the past and having time values increase towards
+the present would require that users manually translate from "years ago".
+This is tedious and error prone.
+\item
+"Now" is a natural time anchor.
+\item
+There is no perfect choice, but we must make a choice.
+Supporting multiple timescales was considered, but this would be a
+substantial undertaking without a clear benefit.
+\end{itemize}
+
+On a more practical note, changing how the timescale is defined
+at this stage of the Demes project would require substantial changes
+to all of the Demes software and documentation.
+
+\ggcomment{
+Translation of models to forwards time simulators...
+Refer to KT's comment above?
+}
 
@@
 
+\ggcomment{
+Check the spec to see if this is clear there.
+Then refer to the spec.
+}
+
 \end{reply}
@@
 
+\ggcomment{
+\begin{itemize}
+\item
+We provide parsers, so the simulation software doesn't have to
+write this. Thus parsers are shared between simulators.
+\item
+The parser does validation too.
+\item
+A simulator can choose to only accept MDM models.
+\end{itemize}
+}
+
 \end{reply}
@@
 
+\ggcomment{
+Say something about sparse versus dense.
+Then refer to our docs so that we don't have to copy/paste the docs in the manuscript.
+}
+
 \end{reply}
@@
 
+\ggcomment{
+Refer to our docs so that we don't have to copy/paste the docs in the manuscript.
+}
+
 \end{reply}
@@
 TODO
+
+\ggcomment{
+Refer to our docs so that we don't have to copy/paste the docs in the manuscript.
+}
 
@@
 TODO
+
+\ggcomment{
+Not sure what this means, exactly? Likely beyond the scope of Demes.
+}
 

### 2022-07-23 paper.tex
Move Fig A1A and B into main text
See #77. Some text still needs to be added to the new section.

I also
* removed the line numbers, as we don't refer to them
  anywhere and I think this looks less cluttered, and
* improved consistency betwee HDM and MDM yaml in regard
  to indentation and using yaml block style.

@@
 (as continuous ``migrations'' or instantaneous ``pulse'' events)
-(e.g., Figures~\ref{fig:showcase}~and~\ref{fig:IM}).
+(e.g., Figures~\ref{fig:IM}~and~\ref{fig:showcase}).
 The HDM encourages human understanding by avoiding
@@
 specifying default values that are inherited hierarchically.
+
+\subsection*{Example: an isolation with migration model}
+
+\ggcomment{Add some text here describing the YAML in Figure \ref{fig:IM}.}
+
+\begin{figure}[h!]
+    \begin{minipage}{0.45\textwidth}
+    \begin{subfigure}{\textwidth}
+    \caption{}
+        \begin{tcolorbox}[equal height group=IM]
+            \inputminted[fontsize=\scriptsize,numbersep=5pt]{yaml}{models/IM.yaml}
+        \end{tcolorbox}
+    \end{subfigure}
+    \end{minipage}\hfill%
+%
+    \begin{minipage}{0.50\textwidth}
+    \begin{subfigure}{\textwidth}
+    \caption{}
+        \begin{tcolorbox}[equal height group=IM, boxsep=0pt]
+            \includegraphics[width=\textwidth]{fig/IM}
+        \end{tcolorbox}
+    \end{subfigure}
+    \end{minipage}\\
+
+    \caption{
+        \label{fig:IM}
+        Example isolation-with-migration Demes model. (A) The Human Data Model
+        representation expressed using YAML.
+        (B) A visual representation of the model using \texttt{demesdraw}.
+        The same model in the Machine Data Model form is provided in
+        Figure~\ref{fig:IM-MDM}.
+        The YAML descriptions in (A) and Figure~\ref{fig:IM-MDM} correspond
+        exactly to a JSON description, but are much more human-readable.
+    }
+\end{figure}
 
@@
 \begin{figure}[h!]
-    \begin{minipage}{0.44\textwidth}
-    \begin{subfigure}{\textwidth}
-    \caption{}
     \begin{tcolorbox}
-            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM.yaml}
+        \inputminted[fontsize=\scriptsize,numbersep=5pt]{yaml}{models/IM-resolved.yaml}
     \end{tcolorbox}
-    \end{subfigure}
-    \begin{subfigure}{\textwidth}
-    \caption{}
-        \begin{tcolorbox}
-            \includegraphics[width=\textwidth]{fig/IM}
-        \end{tcolorbox}
-    \end{subfigure}
-    \end{minipage}\hfill
-    \begin{minipage}{0.54\textwidth}
-    \begin{subfigure}{\textwidth}
-    \caption{}
-        \begin{tcolorbox}
-            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM-resolved.yaml}
-        \end{tcolorbox}
-    \end{subfigure}
-    \end{minipage}\\
     \caption{
-        \label{fig:IM}
-        Example isolation-with-migration Demes model. (A) The Human Data Model
-        representation expressed using YAML.
-        (B) A visual representation of the model using \texttt{demesdraw}.
-        (C) The same model in the Machine Data Model form.
-        The YAML descriptions in (A) and (C) correspond exactly to a JSON description,
-        but are much more human-readable.
+        \label{fig:IM-MDM}
+        Isolation-with-migration example model from Figure~\ref{fig:IM}
+        in Machine Data Model (MDM) form.
+        The MDM form of the model is complete and explicit, but contains
+        much redundant information that is omitted in the Human Data Model
+        (HDM) form.
     }
@@
     \begin{tcolorbox}
-        \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/tennessen.yaml}
+        \inputminted[fontsize=\scriptsize,numbersep=5pt]{yaml}{models/tennessen.yaml}
     \end{tcolorbox}
@@
         % NOTE: line highlights may need adjusting after changing this code
-        \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt,highlightlines={9,19,52}]{python}{models/tennessen-simulation.py}
+        \inputminted[fontsize=\scriptsize,numbersep=5pt,highlightlines={9,19,52}]{python}{models/tennessen-simulation.py}
     \end{tcolorbox}

### 2022-08-01 response-to-reviewers.tex
clarify that MDM is valid HDM

@@
 
-\ggcomment{Anything here we should clarify in the manuscript?}
-
 We've spent considerable effort trying to anticipate ways in which users
@@
 makes many classes of model misspecification impossible to write,
-requires errors be raised when models are ambiguous,
-and explicitly documents the required behavior for edge cases.
+it requires that errors are raised when models are ambiguous,
+and it explicitly documents the required behavior for edge cases.
 We honestly believe that the Demes format is easier to write
@@
 so users are free to be more explicit in any part of a model.
+This point was not previously mentioned in the manuscript, so we have
+added a paragraph to the Parsers section which explains this.
 
@@
 in interoperability.
+This is not ideal, however, because the software must then
+*validate* the model itself---validation is usually the job of the parser,
+and it's undesirable for users to be manually writing or modifying models
+unless they are subsequently validated according to the spec.
 

### 2022-08-01 paper.tex
clarify that MDM is valid HDM

@@
 Appendix~\ref{sec:appendix-parsers}).
+
+%\subsection*{Machine Data Model}
+A Demes model in MDM form is also in valid HDM form, so parsing an already
+resolved model does not change the model. Furthermore, this allows users to be
+more verbose and write values even when they are not required by the HDM.
+Being explicit may be preferred for reasons of clarity, or if the user is
+unsure about how an HDM model will be resolved into an MDM model.
+Although we note that in the latter case, users can simply look at the
+parser output.
 

### 2022-08-01 response-to-reviewers.tex
Add section explaining Figure 1
Closes #77.

@@
 \begin{reply}
-TODO
+We thank the reviewer for the suggestion.
+We have moved the example model in Figure A1.A (the HDM YAML) and
+Figure A1.B (the model diagram) into the main text as Figures 1A and B,
+while leaving the MDM form of the same model in the Appendix.
+We have also added a section to the text that decribes the main features
+of the model in relation to the YAML document.
 \end{reply}

### 2022-08-01 paper.tex
Add section explaining Figure 1
Closes #77.

@@
 
-\ggcomment{Add some text here describing the YAML in Figure \ref{fig:IM}.}
-
 \begin{figure}[h!]
@@
         representation expressed using YAML.
-        (B) A visual representation of the model using \texttt{demesdraw}.
+        (B) A visual representation of the model using \demesdraw.
         The same model in the Machine Data Model form is provided in
@@
 \end{figure}
+
+In Figure \ref{fig:IM} we provide an example model in Demes YAML format.
+Models typically start with a concise description, followed by the mandatory
+\texttt{time\_units} field.
+This model uses the \texttt{defaults} section to provide a default \texttt{start\_size}
+of 1000 individuals for each epoch of each deme.
+There are three demes in the model, an ancestral deme named ``A'' which ceases
+to exist at 100 generations ago, and demes ``X'' and ``Y'' that derive their
+ancestry from ``A'' at this time.
+Demes A and X have only one epoch, in which the population sizes are
+constant, whereas deme Y has two epochs. Deme Y's second epoch has a different
+\texttt{end\_size} than its \texttt{start\_size}, which indicates the size
+grows exponentially from 1000 individuals at 50 generations ago to 3000
+individuals at time 0 (the present).
+The migration section lists one migration stanza, between demes X and Y.
+This migration stanza doesn't indicate a source or destination deme,
+so the migration is symmetric. No migration times are specified, so migrations
+occur continuously at the given rate during the time interval over which
+both demes exist (from 100 generations ago until the present).
+We do not attempt a detailed explanation of all Demes features here,
+and readers are instead directed to the tutorial included
+in the online Demes specification
+(\url{https://popsim-consortium.github.io/demes-spec-docs/}).
 

### 2022-08-01 response-to-reviewers.tex
More clarifications and responses

@@
 \setcounter{point}{0}
+
+% Import this last to avoid mysterious errors.
+\usepackage[implicit=false]{hyperref}
 
@@
 We agree that this is a useful practice.
-However, in our case there are several software packages that are directly
-and indirectly related to the Demes format as summarized in Table 1.
-We have instead added a link to the specification documentation in the abstract.
+In our case there are several software packages that are directly
+and indirectly related to the Demes format as summarized in Table 1,
+so we have instead added a link to the specification documentation in the abstract.
 This documentation includes a tutorial introduction for users to learn
@@
 
-    Random thoughts from KT:
-
-    The forward time case is pretty easy: find the most ancient
-    thing with a finite time (events and deme start times), and
-    that's your reference point.  A burn-in time pushes that backwards.
-
-    This was straightforward in fwdpy11 and KT has some additional tools
-    "this close" to release to automate the process (for the case of integer-sized
-    clicks in time).
+We thank the reviewer for bringing this omission to our attention.
+It's important that different forward simulators use the same strategy
+for converting from times in the past (as used by Demes) into their
+corresponding values when using a forwards-time convention.
+The trick is to find the oldest non-infinite time in the model
+and use this as the forwards-time reference point, suitably adjusted
+for a burn-in. This same trick is used by DemesDraw for the model diagrams.
+We have added a section to the specification detailing how this
+conversion should be done
+(\url{https://github.com/popsim-consortium/demes-spec/pull/159}),
+and added a brief description to the "Time units" section in the manuscript's
+appendix.
 
@@
 There is no default value for the time units. This must be specified explicitly.
+We mention that the time units field is mandatory in the new Example section
+that describes the new Figure 1.
 \end{reply}
@@
 
+We thank the reviewer for the useful feedback, and for relaying their
+experience.
 We've spent considerable effort trying to anticipate ways in which users
@@
 
-\ggcomment{Anything here we should clarify in the manuscript?}
+We thank the reviewer for the critical feedback and suggestions.
+On a practical note, changing how the timescale is defined
+at this stage of the Demes project would require substantial changes
+to all of the Demes software and documentation.
 
-We apreciate that the timescale we've chosen may at first be jarring
-because it uses backwards-time values with forwards-time language.
-But defining the timescale was fundamental choice that we didn't take lightly.
-All decisions about Demes are publicly recored in github issues,
-pull requests, and code experiments. Two discussions relevant to the
-timescale are
+We appreciate that the timescale we've chosen may at first be jarring
+because Demes time values increase towards the past, but Demes uses
+forwards-time language and ordering conventions.
+Defining the timescale was a fundamental choice that we didn't take lightly.
+All decisions about Demes are publicly recorded in github issues,
+pull requests, and many experiments in both writing models and implementing
+Demes-related software.
+Two discussions relevant to the timescale are
 \url{https://github.com/popsim-consortium/demes-python/issues/2}
@@
 
-On a more practical note, changing how the timescale is defined
-at this stage of the Demes project would require substantial changes
-to all of the Demes software and documentation.
+Demes uses forwards-time conventions in many places, such as for
+"source" and "dest" migration fields (migrants leave a "source"
+deme and have offspring in a "dest" deme),
+epoch ordering (from the past towards the present),
+and a deme must be listed in the input file before its
+descendant demes.
+This latter requirement means that older demes are listed first and that
+demes are topologically sorted according to ancestry relationships.
+Most non-trivial models of biological relevance are treelike with one
+"root" deme that stretches back in time towards infinity.
+The ordering in Demes naturally aligns with how rooted trees are typically drawn:
+top-down as in computer science (\ggcomment{cite Knuth, vol 1, section 2.3}),
+or left-to-right as in phylogenetics
+(\ggcomment{cite Felsenstein, Inferring Phylogenies, chapter 34}).
 
-\ggcomment{
-Translation of models to forwards time simulators...
-Refer to KT's comment above?
-}
+Finally, we note that reviewer 2 also raised the issue of how times should
+be interpreted in a forwards-in-time simulator.
+We have added a section to the specification that describes this,
+and added a paragraph to the "Time units" section of the manuscript's
+appendix which gives backwards-time and forwards-time interpretations
+for infinite time values.
 

### 2022-08-01 paper.tex
More clarifications and responses

@@
 are difficult for humans to understand, making errors likely.
-Input parameter file formats for simulators allow the model specification
+APIs are more verbose, but require a substantial time investment to
+learn, and are tied to a specific tool, so this knowledge is not portable
+to other simulators.
+Like APIs, input parameter file formats for simulators allow the model specification
 to be less terse and allow for documentation in the form of comments.
@@
 of 1000 individuals for each epoch of each deme.
-There are three demes in the model, an ancestral deme named ``A'' which ceases
-to exist at 100 generations ago, and demes ``X'' and ``Y'' that derive their
-ancestry from ``A'' at this time.
+There are three demes in the model, an ancestral deme named ``A'' which exists
+arbitrarily far back into the past then ceases to exist at 100 generations ago,
+and demes ``X'' and ``Y'' that derive their
+ancestry from A when A goes extinct.
 Demes A and X have only one epoch, in which the population sizes are
@@
 values greater than zero with larger values corresponding to times in the more
-distant past. By having time units increase into the past, we avoid the need to
+distant past. By having time values increase into the past, we avoid the need to
 choose an arbitrary point in history as ``time zero''. A natural
@@
 so that downstream software may convert times into generations as required.
+
+There must be at least one population with an infinite \texttt{start\_time}.
+An infinite start time may be interpreted differently depending on the simulator.
+In a coalescent setting, there is no upper bound for the coalescent time
+of lineages in this population.
+In a forwards-time setting, the interval of time between infinity and
+the oldest non-infinite model time (i.e. the ``first event") is approximated
+by the simulator's burn-in phase.
 
@@
 fields of the given type.  A default is superseded by an explicitly provided
-value if given. Other implicit values are inherited naturally following the
+value if given. Size values are inherited naturally following the
 progression of time. For example, if an epoch \texttt{start\_size} is not
-provided, it is assumed to be equal to the \texttt{end\_size} of the previous
+provided (neither directly, nor via a defaults section),
+it is assumed to be equal to the \texttt{end\_size} of the previous
 epoch. This also means that the first epoch of each population must specify the
-initial size.
+initial size (or it must be provided in a defaults section).
 

### 2022-08-01 paper.tex
Apply Jerome's suggestions

@@
 APIs are more verbose, but require a substantial time investment to
-learn, and are tied to a specific tool, so this knowledge is not portable
+learn, and as they are tied to a specific tool this knowledge is not portable
 to other simulators.
@@
 
-In Figure \ref{fig:IM} we provide an example model in Demes YAML format.
+In Figure \ref{fig:IM} we provide an example isolation-with-migration model in Demes YAML format.
 Models typically start with a concise description, followed by the mandatory
@@
 We do not attempt a detailed explanation of all Demes features here,
-and readers are instead directed to the tutorial included
-in the online Demes specification
+and readers are instead directed to the tutorial and detailed specification
+in the online documentation
 (\url{https://popsim-consortium.github.io/demes-spec-docs/}).

### 2022-08-01 response-to-reviewers.tex

@@
 corresponding values when using a forwards-time convention.
-The trick is to find the oldest non-infinite time in the model
+The technique is to find the oldest non-infinite time in the model
 and use this as the forwards-time reference point, suitably adjusted
-for a burn-in. This same trick is used by DemesDraw for the model diagrams.
+for a burn-in. This same approach is used by DemesDraw for the model diagrams.
 We have added a section to the specification detailing how this
@@
 (\url{https://github.com/popsim-consortium/demes-spec/pull/159}),
-and added a brief description to the "Time units" section in the manuscript's
+and added a brief description to the ``Time units'' section in the manuscript's
 appendix.
@@
 \item
-"backwards time" language is confusing when talking about migrations
+``Backwards time'' language is confusing when talking about migrations
 (does one mean migrating individuals, or coalescent lineages moving
@@
 Anchoring time in the past and having time values increase towards
-the present would require that users manually translate from "years ago".
+the present would require that users manually translate from ``years ago''.
 This is tedious and error prone.
 \item
-"Now" is a natural time anchor.
+``Now'' is a natural time anchor.
 \item
@@
 Demes uses forwards-time conventions in many places, such as for
-"source" and "dest" migration fields (migrants leave a "source"
-deme and have offspring in a "dest" deme),
+``source'' and ``dest'' migration fields (migrants leave a ``source''
+deme and have offspring in a ``dest'' deme),
 epoch ordering (from the past towards the present),
@@
 Most non-trivial models of biological relevance are treelike with one
-"root" deme that stretches back in time towards infinity.
+``root'' deme that stretches back in time towards infinity.
 The ordering in Demes naturally aligns with how rooted trees are typically drawn:
@@
 We have added a section to the specification that describes this,
-and added a paragraph to the "Time units" section of the manuscript's
+and added a paragraph to the ``Time units'' section of the manuscript's
 appendix which gives backwards-time and forwards-time interpretations

### 2022-08-02 response-to-reviewers.tex

@@
 useful addition to the field and which I hope will be largely adopted.
+\end{point}
+\begin{reply}
+Thank you for the kind words, we are delighted that Demes has been useful
+to you!
+\end{reply}
 
+\begin{point}
 I don't have much to add to this manuscript other than a few comments that I
@@
 \begin{reply}
-TODO
-
+We have addressed each of the comments in turn below.
 \end{reply}
@@
 \begin{reply}
-
 We agree that this is a useful practice.
@@
 \begin{reply}
-We thank the reviewer for the suggestion.
+Thank you for this excellent suggestion.
 We have moved the example model in Figure A1.A (the HDM YAML) and
@@
 \begin{reply}
-TODO
-
-\ggcomment{
-Well, we *could* include lots more detail, but I think we want to be clear
-that the manuscript isn't intended to completely document everything---that's
-what the spec is for.
-}
+We have updated the text to clarify that only the exponential population
+size function is currently supported, but the specification may be updated
+to include more.
 \end{reply}
@@
 \begin{reply}
-TODO
-
+Fixed.
 \end{reply}
@@
 \begin{reply}
-TODO
-
+We have rephrased this, and illustrated in the new ``Example'' section.
 \end{reply}

### 2022-08-02 paper.tex

@@
 and end sizes that must remain positive.
+Only exponential population size changes are currently supported,
+but other functions may be added to the specification over time.
 
@@
 Within the sexually-reproduced offspring,
+a proportion
 $S$ are born via self-fertilization while the rest

### 2022-08-02 response-to-reviewers.tex
Finish responses to Reviewer 1

@@
 \begin{reply}
-TODO
-
+We have updated the Parsers main text section to make this point.
 \end{reply}
@@
 \begin{reply}
-TODO
-
-\ggcomment{These sound like synonyms to me too.
-Can "sequence annotations" be deleted, or is this a distinct thing?}
+Good point; we have deleted the ``sequence annotations''.
 \end{reply}
 
-\textit{Random thoughts I had while reading the paper, to discuss (or not)}
+\subsection*{Random thoughts I had while reading the paper, to discuss (or
+not)}
 

### 2022-08-02 paper.tex
Finish responses to Reviewer 1

@@
 software.
+Having common parsers also benefits users by providing
+consistent and informative error messages for missing values or
+issues in formatting.
 In addition to a reference implementation written in Python,
@@
 including mutation and recombination rates,
-genome annotations, ploidy, sequence annotations, and so on.
+genome annotations, ploidy, and so on.
 Selection and dominance models are absent, as discussed in

### 2022-08-02 paper.tex
Move Example to its own section
Didn't make sense to have this within the Demes section
Remove Parsers appendix
Closes #84

@@
 
-\subsection*{Example: an isolation with migration model}
+\subsection*{Parsers}
+
+While the HDM is designed for human readability and
+conciseness, the underlying data model suitable for software implementation
+(the Machine Data Model, or MDM)
+is redundant and exhaustive.
+Translation from the HDM to the MDM
+requires resolving hierarchically-defined default values and verifying
+relationships between populations and the validity of specified parameter
+values.
+Because this translation and validation requires significant
+programming effort, we define a standard software entity as part of the
+specification to perform this task (the parser),
+which is intended to be shared by programs that support Demes as input.
+The Demes specification precisely defines the required behavior of parsers,
+and we provide a reference implementation written in Python to resolve any
+potential ambiguities, as well as an extensive test suite of examples and the
+expected outputs. In addition, we have high-quality parser implementations in
+the Python and C languages (Table~\ref{tab:software})
+providing a solid foundation for the software ecosystem.
+By maintaining high-quality Demes parsers available as libraries,
+we ensure consistency across simulation and inference
+software.
+Having common parsers also benefits users by providing
+consistent and informative error messages for missing values or
+issues in formatting.
+
+%\subsection*{Machine Data Model}
+A Demes model in MDM form is also in valid HDM form, so parsing an already
+resolved model does not change the model. Furthermore, this allows users to be
+more verbose and write values even when they are not required by the HDM.
+Being explicit may be preferred for reasons of clarity, or if the user is
+unsure about how an HDM model will be resolved into an MDM model.
+Although we note that in the latter case, users can simply look at the
+parser output.
+
+\subsection*{The scope of Demes}
+
+The Demes specification is static by design---we wish to unambiguously
+describe a demographic model with a concrete set of parameters.
+This simplicity means that we cannot directly specify parameter distributions
+or estimated confidence intervals for those parameters.
+While it is not difficult to imagine extending the specification in ways that
+would allow this (Appendix~\ref{sec:appendix-scope}), it is not immediately
+clear that the benefits are worth the increased parser complexity.
+It is important to note, however, that Demes may be used in applications that
+include additional population genetic processes outside of what is explicitly
+modeled in the specification, such as interpreting population sizes as
+carrying capacities, implementations of hard selection, or layering more
+complicated mating or spatial structure. The Demes specification therefore
+provides a basic model that can be elaborated on where necessary.
+
+\section*{Example: an isolation with migration model}
 
@@
 (\url{https://popsim-consortium.github.io/demes-spec-docs/}).
-
-\subsection*{Parsers}
-
-While the HDM is designed for human readability and
-conciseness, the underlying data model suitable for software implementation
-(the Machine Data Model, or MDM)
-is redundant and exhaustive.
-Translation from the HDM to the MDM
-requires resolving hierarchically-defined default values and verifying
-relationships between populations and the validity of specified parameter
-values.
-Because this translation and validation requires significant
-programming effort, we define a standard software entity as part of the
-specification to perform this task (the parser),
-which is intended to be shared by programs that support Demes as input.
-By maintaining high-quality Demes parsers available as libraries,
-we ensure consistency across simulation and inference
-software.
-Having common parsers also benefits users by providing
-consistent and informative error messages for missing values or
-issues in formatting.
-In addition to a reference implementation written in Python,
-we provide high-quality parser implementations in the Python and
-C languages (Table~\ref{tab:software}, as discussed in
-Appendix~\ref{sec:appendix-parsers}).
-
-%\subsection*{Machine Data Model}
-A Demes model in MDM form is also in valid HDM form, so parsing an already
-resolved model does not change the model. Furthermore, this allows users to be
-more verbose and write values even when they are not required by the HDM.
-Being explicit may be preferred for reasons of clarity, or if the user is
-unsure about how an HDM model will be resolved into an MDM model.
-Although we note that in the latter case, users can simply look at the
-parser output.
-
-\subsection*{The scope of Demes}
-
-The Demes specification is static by design---we wish to unambiguously
-describe a demographic model with a concrete set of parameters.
-This simplicity means that we cannot directly specify parameter distributions
-or estimated confidence intervals for those parameters.
-While it is not difficult to imagine extending the specification in ways that
-would allow this (Appendix~\ref{sec:appendix-scope}), it is not immediately
-clear that the benefits are worth the increased parser complexity.
-It is important to note, however, that Demes may be used in applications that
-include additional population genetic processes outside of what is explicitly
-modeled in the specification, such as interpreting population sizes as
-carrying capacities, implementations of hard selection, or layering more
-complicated mating or spatial structure. The Demes specification therefore
-provides a basic model that can be elaborated on where necessary.
 
@@
 
-\section{Parsers}\label{sec:appendix-parsers}
-
-The implicit value inference and default value propagation required to
-fully resolve a Demes HDM description into the corresponding
-MDM form is straightforward
-to describe, but still requires significant effort to implement in
-software. Moreover, there are many constraints on the data model
-(for example, epoch \texttt{end\_time} values should be strictly decreasing
-within a deme), which must be enforced.
-It would not be reasonable to require every program that
-takes Demes models as input to implement this logic, as the programming
-effort would be significant (limiting adoption)
-and it is likely that if there were many implementations some would differ
-in their details (harming the software ecosystem).
-% GG: this is currently the case for programs accepting ms command lines. E.g.
-%  - MaCS interprets -es proportion as 1-proportion compared to ms.
-%  - SCRM uses different options for -ema command.
-We therefore define
-a standard software entity as part of the specification (the parser),
-which performs this task and can be shared by programs that
-support Demes as an input. By having relatively few high-quality Demes
-parsers available as libraries (ideally, one per programming language),
-the probability of divergences from the standard is greatly reduced.
-
-The Demes specification precisely defines the required behavior of parsers,
-which translate a model described in the HDM into the MDM, where values have been
-assigned to all parameters and data
-constraints have been checked. The output is formally defined as JSON, but in
-practice parsers output an object model that is suitable for the particular
-programming language and that can be used directly by the implementing program.
-We provide a reference implementation written in Python to resolve any
-potential ambiguities and to provide a helpful template for other
-implementations, as well as an extensive test suite of examples and the
-expected outputs. In addition, we have high-quality parser implementations in
-the Python and C languages (published under liberal open source licenses),
-providing a solid foundation for the software ecosystem.
-
 \section{Scope of the specification}

### 2022-08-02 paper.tex
Move content into main text and rejig

@@
 
-Below we provide a brief overview of the population genetics models
+Here we provide a brief overview of the population genetics models
 that Demes supports and the components of the Demes infrastructure.
-We then highlight an example usage, including visualization and simulation
-of a multi-population demographic model.
-More detailed descriptions of the population genetics model, parsers, scope, and
-additional examples are given in the Appendix.
 Complete technical details of the MDM and HDM, and the responsibilities
@@
 (\url{https://popsim-consortium.github.io/demes-spec-docs/}).
+This specification rigorously defines the data model,
+fully describing the entities and their
+relationships, and the required behavior of implementations.
+Since the online specification is definitive,
+we will not recapitulate the details
+here, but instead focus on the high level properties of the model and
+the rationale behind key design decisions.
 
@@
 primary interface for writing and interchanging demographic models
-(see Appendix~\ref{sec:appendix-spec} for details about our choice of
-YAML).
-Demographic models written in YAML format provide information about global
+(see Appendix~\ref{sec:appendix-spec} for rationale).
+Demographic models provide information about global
 features of the model (such as time units and generation times),
@@
 and gene flow between populations
-(as continuous ``migrations'' or instantaneous ``pulse'' events)
-(e.g., Figures~\ref{fig:IM}~and~\ref{fig:showcase}).
-The HDM encourages human understanding by avoiding
-redundancy in the description where possible and by providing a mechanism for
-specifying default values that are inherited hierarchically.
+(as continuous ``migrations'' or instantaneous ``pulse'' events).
+Fig~\ref{fig:IM} shows an isolation-with-migration in HDM format.
+
+Structurally, the HDM encourages human understanding
+by avoiding redundancy in the description where possible and by providing a
+mechanism for specifying default values that are inherited hierarchically.
+For values that repeat
+across fields, defaults may be used to implicitly assign default values to
+fields of the given type.  A default is superseded by an explicitly provided
+value if given. Size values are inherited naturally following the
+progression of time. For example, if an epoch \texttt{start\_size} is not
+provided (neither directly, nor via a defaults section),
+it is assumed to be equal to the \texttt{end\_size} of the previous
+epoch. This also means that the first epoch of each population must specify the
+initial size (or it must be provided in a defaults section).
+
+Avoiding redundancy in this way reduces the cognitive load on readers,
+by highlighting necessary parameters which may be otherwise be obscured.
+It is not necessary---or indeed recommended---that all models are expressed
+in a maximally concise form, and we wholeheartedly endorse
+the explicit statement of parameters where it increases model legibility.
 
@@
 
-%\subsection*{Machine Data Model}
-A Demes model in MDM form is also in valid HDM form, so parsing an already
-resolved model does not change the model. Furthermore, this allows users to be
-more verbose and write values even when they are not required by the HDM.
-Being explicit may be preferred for reasons of clarity, or if the user is
-unsure about how an HDM model will be resolved into an MDM model.
-Although we note that in the latter case, users can simply look at the
-parser output.
+\subsection*{Scope of the specification}
 
-\subsection*{The scope of Demes}
+A primary design goal of Demes is to provide a means of
+unambiguously communicating the results of demographic
+model inferences to population genetic simulators.
+Since demography is defined in terms of groups of individuals
+and these groupings are influenced by genetics,
+it is difficult to find a simple definition that separates
+the two. Thus, we have attempted to be pragmatic,
+limiting the features that we include in Demes to
+those that are (or can reasonably be expected to)
+in practise regarded as part of a demographic model.
+
+Thus, the model is limited to features that we can expect many different
+demographic inference and simulation methods to share. The specification only
+describes demographic features at the population level. Features of genome
+biology are out of scope, including mutation and recombination rates,
+genome annotations, ploidy, and so on. Selection and dominance models are
+absent, as discussed in Appendix.~\ref{sec:appendix-pop-gen-model}.
+It is important to note, however, that Demes may be used in applications that
+include additional population genetic processes outside of what is explicitly
+modelled in the specification, such as interpreting population sizes as
+carrying capacities, implementations of hard selection, or layering more
+complicated mating or spatial structure. The Demes specification
+is intended to provide a basic model that can be elaborated on where necessary.
+
+Demes is not a standard population genetic simulation specification,
+although it could be \emph{part} of one.
+Since the standard is based on JSON, and JSON documents can be
+arbitrarily nested, it is possible to imagine a simple specification
+of genome features such as mutation and recombination rates
+in which the demography is defined by an embedded Demes specification.
+Features of the simulation specification (such as defining
+the time and location of samples) can then \emph{refer to} the Demes
+model. This design, in which we embed the Demographic model
+\emph{within} a larger specification rather than adding arbitrary
+and unrelated complexities \emph{to} the demography
+is an essential simplification and separation of duties.
 
@@
 While it is not difficult to imagine extending the specification in ways that
-would allow this (Appendix~\ref{sec:appendix-scope}), it is not immediately
-clear that the benefits are worth the increased parser complexity.
-It is important to note, however, that Demes may be used in applications that
-include additional population genetic processes outside of what is explicitly
-modeled in the specification, such as interpreting population sizes as
-carrying capacities, implementations of hard selection, or layering more
-complicated mating or spatial structure. The Demes specification therefore
-provides a basic model that can be elaborated on where necessary.
+would allow this, it is not
+clear that the benefits are worth the increased parser complexity
+(see Appendix~\ref{sec:appendix-static}).
 
-\section*{Example: an isolation with migration model}
+\section*{Example: an isolation-with-migration model}
 
@@
 built as part of developing Demes, but many other useful tools
-can be envisaged that consume, transform, or produce this format.
+can be envisaged that produce, consume, or transform this format.
 
@@
 Matthew Hartfield is supported by a NERC Independent Research Fellowship (NE/R015686/1).
+Jerome Kelleher is supported by the Robertson Foundation.
 
@@
 (\url{https://popsim-consortium.github.io/demes-spec-docs/}).
-This specification rigorously defines the data model,
-fully describing the entities and their
-relationships, and the required behavior of implementations.
-Since the online specification is definitive,
-we will not recapitulate the details
-here, but instead focus on the high level properties of the model and
-the rationale behind key design decisions.
-
-Below we detail the population genetics model in
-Appendix~\ref{sec:appendix-pop-gen-model},
-the high-level, human-readable model specification in
-Appendix~\ref{sec:appendix-spec}, demographic model parsers and their
-intended behavior in Appendix~\ref{sec:appendix-parsers}, and the
-intended scope of Demes in Appendix~\ref{sec:appendix-scope}.
-The extended figures in Appendix~\ref{sec:appendix-figures} provide
-additional examples and illustrations of Demes usage.
 
@@
 
-\section{Human Data Model}\label{sec:appendix-spec}
+\section{Rationale for YAML}\label{sec:appendix-spec}
 
-The Demes Human Data Model (HDM) is focused on the efficiency of
-human understanding and the avoidance of errors. We have adopted the widely
+We have adopted the widely
 used YAML format~\citep{ben2009yaml} as the recommended means of interchanging
@@
 
-Structurally, the HDM encourages human understanding
-by avoiding redundancy in the description where possible and by providing a
-mechanism for specifying default values that are inherited hierarchically.
-Figure~\ref{fig:IM} shows an example Demes model expressed in both the
-HDM and Machine Data Model (MDM) forms (both in terms of YAML syntax).
-For values that repeat
-across fields, defaults may be used to implicitly assign default values to
-fields of the given type.  A default is superseded by an explicitly provided
-value if given. Size values are inherited naturally following the
-progression of time. For example, if an epoch \texttt{start\_size} is not
-provided (neither directly, nor via a defaults section),
-it is assumed to be equal to the \texttt{end\_size} of the previous
-epoch. This also means that the first epoch of each population must specify the
-initial size (or it must be provided in a defaults section).
-
-\section{Scope of the specification}
-\label{sec:appendix-scope}
-
-\subsection{Static models, not parameterized models}
+\section{Rationale for static models}
 \label{sec:appendix-static}
@@
 
-\subsection{Population-level features, not genome features}
-\label{sec:appendix-features}
-
-Demes is designed to interface with a large number of simulation and inference
-methods, so we have restricted the specification to only describe
-demographic features at the population level.
-Features of the underlying genome biology are omitted,
-including mutation and recombination rates,
-genome annotations, ploidy, and so on.
-Selection and dominance models are absent, as discussed in
-Sec.~\ref{sec:appendix-pop-gen-model}, and we do not
-specify how individuals are sampled.
-
-The model is, by design, very limited. We do not specify details of
-individuals or of the processes that happen within populations,
-because such details are necessarily complicated, vary greatly by
-application, and attempts to encompass such a broad range of biological
-processes are unlikely to succeed.
-The narrow focus on individuals and their groupings into populations
-should ensure that the Demes standard is stable, and not subject to
-continual development as more and more elaborations of basic processes
-are added. Demes is intended to provide \emph{part} of a
-simulation model specification; other parts of the specification,
-such as sampling strategies, or parameters that vary by time
-or population can \emph{refer to} this well defined and unambiguous
-description of the demography.
-
 \section{Extended data and figures}

### 2022-08-03 paper.tex
Tweak from review

@@
 limiting the features that we include in Demes to
-those that are (or can reasonably be expected to)
-in practise regarded as part of a demographic model.
+those that are in practise regarded as part of a demographic model.
 

### 2022-08-03 response-to-reviewers.tex
Remove Parsers appendix
Closes #84

@@
 Thank you for considering this manuscript for publication.
+
+Things to mention:
+
+\begin{itemize}
+\item We compressed the text and improved the presentation by removing
+the Parsers appendix (there was little extra information that was not
+already in the existing main text section).
+\item Reviewer 2 made some good and fair points, but which were mainly
+(we believe) missing the underlying rationale for design decisions.
+We have rearranged the text, bringing some content that was in the
+appendix into the main text to try and address this up front
+and avoid similar misunderstandings.
+\end{itemize}
 
@@
 
-The HDM was explicitly designed to be a less-strict version of the MDM,
+The HDM was explicitly designed to be a less-verbose version of the MDM,
 and our initial approach was to simply allow the omission of some fields
@@
 so users are free to be more explicit in any part of a model.
-This point was not previously mentioned in the manuscript, so we have
-added a paragraph to the Parsers section which explains this.
+We have also pointed out that it is not necessary to be maximally
+concise when describing models, and we encourage redundancy where
+it increases legibility.
 

### 2022-08-03 paper.tex
Light editing pass

@@
 migrations are also shared by many inference methods
-\citep[e.g.,][]{gutenkunst2009inferring,gravel2012population,kamm2017efficient,
+\citep[e.g.,][]{gutenkunst2009inferring,li2011inference,gravel2012population,kamm2017efficient,
 jouganous2017inferring,ragsdale2019models,excoffier2021fastsimcoal2}.
@@
 (as continuous ``migrations'' or instantaneous ``pulse'' events).
-Fig~\ref{fig:IM} shows an isolation-with-migration in HDM format.
+Fig~\ref{fig:IM} shows an example isolation-with-migration model in HDM format.
 
@@
 For values that repeat
-across fields, defaults may be used to implicitly assign default values to
+across fields, the ``defaults'' mechanism may be used to implicitly assign default values to
 fields of the given type.  A default is superseded by an explicitly provided
@@
 progression of time. For example, if an epoch \texttt{start\_size} is not
-provided (neither directly, nor via a defaults section),
+provided (either directly, or via a defaults section),
 it is assumed to be equal to the \texttt{end\_size} of the previous
@@
 
-Thus, the model is limited to features that we can expect many different
+The model is therefore limited to features that we can expect many different
 demographic inference and simulation methods to share. The specification only
@@
 genome annotations, ploidy, and so on. Selection and dominance models are
-absent, as discussed in Appendix.~\ref{sec:appendix-pop-gen-model}.
+absent, as discussed in Appendix~\ref{sec:appendix-pop-gen-model}.
 It is important to note, however, that Demes may be used in applications that
@@
 the time and location of samples) can then \emph{refer to} the Demes
-model. This design, in which we embed the Demographic model
+model. This design, in which we embed the demographic model
 \emph{within} a larger specification rather than adding arbitrary
@@
 would allow this, it is not
-clear that the benefits are worth the increased parser complexity
+clear that the benefits are worth the greatly increased parser complexity
 (see Appendix~\ref{sec:appendix-static}).
-
-\section*{Example: an isolation-with-migration model}
 
@@
         Figure~\ref{fig:IM-MDM}.
-        The YAML descriptions in (A) and Figure~\ref{fig:IM-MDM} correspond
-        exactly to a JSON description, but are much more human-readable.
     }
@@
 
-In Figure \ref{fig:IM} we provide an example isolation-with-migration model in Demes YAML format.
+\section*{Example: an isolation-with-migration model}
+In Figure \ref{fig:IM} we provide an example isolation-with-migration model.
 Models typically start with a concise description, followed by the mandatory
@@
 and demes ``X'' and ``Y'' that derive their
-ancestry from A when A goes extinct.
+ancestry from A when it goes extinct.
 Demes A and X have only one epoch, in which the population sizes are
@@
 has been \textit{ad-hoc}, and conversions between the many different ways
-of expressing such models is both labor intensive and fraught with errors.
+of expressing such models is both labor intensive and error-prone.
 The proposed Demes standard is an attempt to bridge this gap
@@
 for reproducibility is that we must be able to state precisely what
-the result in question is. The lack of standardization in how
+the result in question \emph{is}. The lack of standardization in how
 complex demographic models are communicated today, and the lack of
@@
 the oldest non-infinite model time (i.e. the ``first event") is approximated
-by the simulator's burn-in phase.
+by the simulator's burn-in phase---detailed guidance
+is provided in the online specification.
 
@@
 this effect is negligible in large populations \citep{nordborg1997coalescent}.
-Simulators that allow variable rates of selfing are expected to clearly
-document the expected behavior.
 
@@
 \citet{nordborg1997coalescent,hartfield2016facsexcoal}.
-
-%\mhcomment{If requested
-%I can quickly draw up a schematic diagram to illustrate these partitions.}
 
@@
 serialization language with an emphasis on simplicity and which interoperates well
-with JSON (indeed, YAML 1.2 is a super set of JSON). We chose YAML over JSON
+with JSON (indeed, YAML 1.2 is a superset of JSON). We chose YAML over JSON
 because although JSON is an excellent format for data interchange, it is
-ill-suited for human understanding or manipulation. We also considered other
+ill-suited for human understanding and manipulation. We also considered other
 declarative data exchange formats such as TOML,
-but chose YAML because of equivalence with JSON,
-its popularity and good software support.
+but chose YAML because of its equivalence with JSON,
+popularity, and good software support.
 Since the Demes data model is defined in JSON Schema,
@@
 complexity of parsers, with relatively limited benefit to most users.
-It's unlikely that Demes could be made sufficiently
-flexible without considerable effort to implement many features of
+It is unlikely that Demes could be made sufficiently
+flexible without implementing many features of
 general-purpose programming languages, such as variables, arithmetic,
@@
         passed to \moments\ to compute the expected joint SFS.
-        To compare the SFS in Figure~\ref{fig:showcase}, we \emph{marginalize} the
+        To compare the SFS in Figure~\ref{fig:showcase}, we marginalize the
         joint SFS to obtain the single-population SFS for both AFR and EUR populations.

### 2022-08-03 paper.tex
Rejig popgen model
Move content up from appendix, and try to be clearer on the motivation

@@
 \subsection*{Population genetics model}
+For inference and simulation software to meaning fully interoperate there
+must be a shared understanding of what a demographic model \emph{is}.
+Population genetics is a large field, and rather than attempting to
+capture all possible within- and between- population processes, we have
+instead adopted a pragmatic approach of identifying a common set
+of assumptions shared by many methods. We outline the processes
+and assumptions briefly here and in
+Appendix~\ref{sec:appendix-pop-gen-model}.
 
@@
 relationships  or by continuous or discrete migration events.
-The demographic model itself omits information about genome biology, such as
-genome size and architecture, ploidy, and mutation and recombination rates.
-It also omits any information about selection operating within populations.
-Within populations, allele-frequency dynamics are assumed to
-follow the Wright-Fisher model, so that generations are discrete and
-non-overlapping and population sizes are independent of mean fitness
-(i.e., ``soft'' selection~\citep{christiansen1975hard}).
-While these restrictions preclude some population-genetic scenarios of
-interest, such as continuous spatial structure
-\citep{wright1943isolation,barton2002neutral,barton2010new,
-ringbauer2017inferring,battey2020space}
-or stochastic population size dynamics
-\citep{nunney1993assessing,orr2014population},
-many current forwards- and backwards-in-time simulators
-make these same assumptions.
-The basic assumptions of discrete populations connected by instantaneous or continuous
-migrations are also shared by many inference methods
-\citep[e.g.,][]{gutenkunst2009inferring,li2011inference,gravel2012population,kamm2017efficient,
-jouganous2017inferring,ragsdale2019models,excoffier2021fastsimcoal2}.
+As described in the Scope of the Specification section below, the demographic
+model does not, as a deliberate simplification and separation of
+duties, include any information about genome biology or selection.
+Within a population, we assume Wright-Fisher dynamics
+(see Appendix~\ref{sec:appendix-population-dynamics} for more
+precise details).
+
+These basic assumptions of discrete Wright-Fisher populations connected by
+instantaneous or continuous
+migrations are shared by many inference methods
+\citep[e.g.,][]{gutenkunst2009inferring,li2011inference,
+gravel2012population,kamm2017efficient,
+jouganous2017inferring,ragsdale2019models,
+excoffier2021fastsimcoal2}, and
+forwards- and backwards-time simulators~\citep[e.g.,][]{hudson2002generating,
+gutenkunst2009inferring,
+excoffier2011fastsimcoal,kelleher2016efficient,
+jouganous2017inferring,haller2019slim,thornton2019-nu}.
 Demes therefore serves as ``middleware'' between inference methods and simulation
 software, capturing these common assumptions.
-Appendix~\ref{sec:appendix-pop-gen-model} provides more details on the
-population genetics models supported by Demes.
+
+It is important to note that the goal of describing
+the basic population processes precisely is not to be
+proscriptive about what methods may or may not use the specification,
+but so that we can be clear on what situations we can
+expect methods to agree exactly. Arbitrary population
+processes
+---for example, within-deme
+continuous spatial structure
+\citep{wright1943isolation,barton2002neutral,barton2010new,
+ringbauer2017inferring,battey2020space}---may
+be layered on top of this basic description,
+but as dynamics diverge from the core assumptions,
+then of course we can expect results to differ
+accordingly.
 
@@
 \subsection{Population dynamics}
-
+\label{sec:appendix-population-dynamics}
 Within a population, we assume that allele frequency dynamics can
@@
 \citep[e.g.,][]{hudson2002generating,gutenkunst2009inferring,excoffier2011fastsimcoal,kelleher2016efficient,jouganous2017inferring,haller2019slim,thornton2019-nu}.
-Each has their own unique method of specifying the details of selection and demography.
-The goal of Demes is to provide a common method for the latter task.
 

### 2022-08-03 response-to-reviewers.tex
Respond to the rest of Laurent's comments

@@
 \begin{reply}
-TODO
+Dear Laurent,
+
+Thank you for the kind words and the careful and insightful review.
+We have found your comments extremely useful, helping us to see
+parts of the manuscript where we had failed to sufficiently explain
+key points and motivations. We have addressed each of your points
+in turn below, and would be grateful for any further thoughts
+you may have.
+
+On a practical note, we have waited until Demes and the surrounding
+software ecosystem is quite mature before submitting this paper
+for publication. On the upside, this means that the features
+we are describing are battle-hardened, and the abstractions that
+we are using have worked well in practice for a range of
+applications. On the downside, however, this means that our ability
+to change the specification is rather limited, as we hope you
+will appreciate.
+
 \end{reply}
@@
 \end{point}
+
 \begin{reply}
+This is a reasonable point, which we have addressed by
+(a) making the rationale for the HDM clearer in the text; and (b) adding an
+example IM model in HDM format, which we briefly explain. Hopefully
+this will be enough for readers to understand the basic elements of the
+model.
 
-We thank the reviewer for the useful feedback, and for relaying their
-experience.
+A key goal of the HDM is human \emph{understanding} --- we hope that
+users can look at a Demes model (and perhaps the demesdraw rendering)
+as illustrated in the new figure, and understand key features of it.
+Perhaps people may even include models used in the text of their
+papers (as we often see for \texttt{ms} command lines!).
+Conciseness, and avoiding the repetition of ``obvious'' values,
+is a key part of making this representation human-readable.
+
+In terms of user-friendliness, we think that
+the tools available for working with
+Demes really are substantially better than the status quo.
 We've spent considerable effort trying to anticipate ways in which users
@@
 and it explicitly documents the required behavior for edge cases.
-We honestly believe that the Demes format is easier to write
-and interpret than the status quo.
-However, we agree that there are some aspects of the HDM that will be
-confusing for new users.
+Of course, there will be some aspects of the HDM that will be
+confusing for new users --- nothing is perfect.
 
@@
 in interoperability.
-This is not ideal, however, because the software must then
-*validate* the model itself---validation is usually the job of the parser,
-and it's undesirable for users to be manually writing or modifying models
-unless they are subsequently validated according to the spec.
-
 \end{reply}
@@
 \begin{reply}
-
-We thank the reviewer for the critical feedback and suggestions.
-On a practical note, changing how the timescale is defined
-at this stage of the Demes project would require substantial changes
-to all of the Demes software and documentation.
-
-We appreciate that the timescale we've chosen may at first be jarring
-because Demes time values increase towards the past, but Demes uses
-forwards-time language and ordering conventions.
-Defining the timescale was a fundamental choice that we didn't take lightly.
+This is an entirely reasonable point, and a question that
+we agonised over at length. There is no correct answer, but
+we believe that the current choice is the ``least-bad'' option.
 All decisions about Demes are publicly recorded in github issues,
@@
 The ordering in Demes naturally aligns with how rooted trees are typically drawn:
-top-down as in computer science (\ggcomment{cite Knuth, vol 1, section 2.3}),
+top-down as in computer science (e.g.~Knuth)
 or left-to-right as in phylogenetics
-(\ggcomment{cite Felsenstein, Inferring Phylogenies, chapter 34}).
-
-Finally, we note that reviewer 2 also raised the issue of how times should
-be interpreted in a forwards-in-time simulator.
-We have added a section to the specification that describes this,
-and added a paragraph to the ``Time units'' section of the manuscript's
-appendix which gives backwards-time and forwards-time interpretations
-for infinite time values.
+(e.g.~Felsenstein)
 
 \end{reply}
+
 \begin{point}
@@
 \begin{reply}
-TODO
+Excellent point, and we entirely agree. We have tried to clarify by simply
+stating in the main text that dynamics are assumed to be Wright-Fisher,
+with a more precise statement in the Appendix. We have also tried to clarify
+the motivation for stating these assumptions in the main text.
+\end{reply}
 
-\end{reply}
 \begin{point}
@@
 \begin{reply}
-TODO
+Indeed---this is a point that we had very long and detailed discussions on.
+In the end, the decision to include them is pragmatic, since several of the
+initial software implementations consider these parameters to be part of
+the demography. We have updated the text in the paper to reflect this
+choice, and stressed the pragmatic approach to defining the line between
+genetics and demography.
+\end{reply}
 
-\end{reply}
 \begin{point}
@@
 \begin{reply}
-TODO
+Default values are inherited hierarchically through the \emph{document},
+rather than the graph of demes.
 
-\ggcomment{
-Check the spec to see if this is clear there.
-Then refer to the spec.
-}
+Please see the specification
+(\url{https://popsim-consortium.github.io/demes-spec-docs/main/specification.html#defaults})
+and tutorial
+(\url{https://popsim-consortium.github.io/demes-spec-docs/main/tutorial.html#setting-defaults})
+for the detailed documentation on this.
+\end{reply}
 
-\end{reply}
 \begin{point}
@@
 \begin{reply}
-TODO
+It's not necessarily true that a program that accepts the MDM will need to
+write its own Demes (MDM) parser. For example, msprime uses the \texttt{demes}
+Python module to accept demographic models in Demes format. In effect,
+what msprime does is take the object model provided by the parser
+which is identical to an MDM document. Msprime then processes this document,
+converting it into its own internal Demography format. Once the conversion
+into msprime's internal format is complete, \emph{this} object
+model is validated using precisely the same code as if the model
+had been produced by msprime's own demography APIs.
+The only validation that msprime does on the Demes MDM document
+is to check for parameters that it doesn't support (like selfing and cloning
+rates).
 
-\ggcomment{
-\begin{itemize}
-\item
-We provide parsers, so the simulation software doesn't have to
-write this. Thus parsers are shared between simulators.
-\item
-The parser does validation too.
-\item
-A simulator can choose to only accept MDM models.
-\end{itemize}
-}
+Thus, from msprime's perpective, there is no difference between what we
+have currently (the HDM converted to MDM via the parser) and what
+you propose (the MDM validated by the parser).
 
+If the HDM did not exist, we would still require parsers because validating
+the full MDM is a lengthy and tedious task.
+A key goal that we set was for the ecosystem to be ``batteries-included''
+from the outset, which is why we have implemented several high-quality
+parsers (Python, and C with Rust, Julia in development) which should
+cover the vast majority of use-cases.
 \end{reply}
+
 \begin{point}
@@
 \begin{reply}
-TODO
-
-\ggcomment{
-Say something about sparse versus dense.
-Then refer to our docs so that we don't have to copy/paste the docs in the manuscript.
-}
+YAML supports a similar type of shorthand, and there are various
+examples in the tutorial where we use it to increase legibility,
+e.g.
+\url{https://popsim-consortium.github.io/demes-spec-docs/main/tutorial.html#exponential-size-changes}
 
@@
 \begin{reply}
-TODO
-
-\ggcomment{
-Refer to our docs so that we don't have to copy/paste the docs in the manuscript.
-}
+We have updated the main text to note that the goal of the paper is not to give the reader
+a detailed understanding of the Demes model, but to argue for its necessity
+and to explain the rationale for key design decisions.
+However, to give the reader a very basic
+understanding of how Demes works, we have added a new Example section that
+gives a simple IM model, with a walk-through in the text.
+Hopefully this, along with the extensive online materials
+(in particular the tutorial), will be sufficient to address specific
+points like this on the structure of the model.
 
@@
 \begin{reply}
-TODO
+See response to point 2.9.
+\end{reply}
 
-\ggcomment{
-Refer to our docs so that we don't have to copy/paste the docs in the manuscript.
-}
-
-\end{reply}
 \begin{point}
@@
 \begin{reply}
-TODO
+We absolutely agree! However, as we stress in several places in the updated
+text this is outside the scope of defining a demographic model, and this
+is a vital simplification. We note in the updated ``Scope of the
+Specification'' section that Demes could be part of a standard simulation
+specification by embedding the demographic model within it.
+\end{reply}
 
-\ggcomment{
-Not sure what this means, exactly? Likely beyond the scope of Demes.
-}
-
-\end{reply}
 \begin{point}
@@
 \begin{reply}
-TODO
-
+See the previous response for the general point; it is
+more general and extensible to embed a demes model \emph{within} an
+overall simulation description, than it is to include details required
+for a simulation in the Demes model.
 \end{reply}

### 2022-08-03 cover-letter/cover-letter2.tex
Add cover letter

\documentclass{letter}

\signature{Jerome Kelleher}

\address{Big Data Institute\\University of Oxford\\UK}
\begin{document}

\begin{letter}{GENETICS}

\opening{Dear Editors,}

I am writing on behalf of my coauthors to resubmit our manuscript entitled
\emph{Demes: a standard format for demographic models}
for consideration for publication in GENETICS
as a ``Methods, Technology, and Resources'' article.

We thank the reviewers for their insightful and helpful reviews. We
believe that we have answered all questions raised, leading to a more
clearly structured manuscript. We have attached
a detailed point-by-point response in the \texttt{response-to-reviewers.pdf}
file.

We have also attached a file showing the changes made to the manuscript
in response to reviewers in \texttt{review-diff.pdf}.

\closing{Sincerely,}

\end{letter}
\end{document}

### 2022-08-03 paper.tex
Respond to the rest of Laurent's comments
Minor updates

@@
 \subsection*{Population genetics model}
-For inference and simulation software to meaning fully interoperate there
+For inference and simulation software to meaningfully interoperate there
 must be a shared understanding of what a demographic model \emph{is}.
@@
 relationships  or by continuous or discrete migration events.
+Within a population, we assume Wright-Fisher dynamics
+(see Appendix~\ref{sec:appendix-population-dynamics} for more
+precise details).
 As described in the Scope of the Specification section below, the demographic
@@
 duties, include any information about genome biology or selection.
-Within a population, we assume Wright-Fisher dynamics
-(see Appendix~\ref{sec:appendix-population-dynamics} for more
-precise details).
 
@@
 value if given. Size values are inherited naturally following the
+the document. Size values are inherited naturally following the
 progression of time. For example, if an epoch \texttt{start\_size} is not

### 2022-08-03 response-to-reviewers.tex

@@
 % General intro text goes here
-Thank you for considering this manuscript for publication.
+Thank you for considering this manuscript for publication, for your
+positive comments and in particular for the helpful and insightful reviews.
 
-Things to mention:
+The changes that we have made mostly reflect an attempt to better
+explain the purpose of Demes, its scope, and the rationale for
+key design decisions. While the changes to the text
+look fairly extensive (see the attached latex-diff), they are mostly
+a function of moving text from the appendix to the main text, and
+removing redundancy.
+
+Specifically:
 
@@
 already in the existing main text section).
-\item Reviewer 2 made some good and fair points, but which were mainly
-(we believe) missing the underlying rationale for design decisions.
+\item Laurent made some good and fair points, but which were mainly
+(we believe) reflecting our failure at communicating
+the underlying rationale for design decisions effectively.
 We have rearranged the text, bringing some content that was in the
@@
 
-\subsection*{Specific comments}
-% We address your specific comments below:
-% % \reviewersection
-
-\textit{
-Demes is an important advance to standardizing the implementation of popgen
-models, and so a very welcome contribution. The reviewers give a number of very
-useful suggestions. I leave it up to you how you proceed with incorporating
-changes based on these comments, as you will have more of a sense of how the
-comments overlap those from other users, but please do write a response to all
-of the comments.
-}
+We have responded to all comments point-by-point, below.
 
@@
 \begin{reply}
-
 We have had an issue open for this for some time (\url{https://github.com/popsim-consortium/demes-spec/issues/20}).
@@
 Files omitting a spec version will assume a default value of 1.0.
-
 \end{reply}
@@
 \begin{reply}
-
-We thank the reviewer for bringing this omission to our attention.
+Thank you for bringing this omission to our attention.
 It's important that different forward simulators use the same strategy
@@
 appendix.
-
 \end{reply}
@@
 \begin{reply}
-
 We have updated the text to include this point.
@@
 \begin{point}
-p2 "YAML format" you could specify that it is a data serialization language
+p2 ``YAML format" you could specify that it is a data serialization language
 \end{point}
 \begin{reply}
-
 We have updated the text to include this point.
@@
 \begin{point}
-p12 "time units is in generations" you could precise it is the default value here, if it is?
+p12 ``time units is in generations" you could precise it is the default value here, if it is?
 \end{point}
 \begin{reply}
-
 There is no default value for the time units. This must be specified explicitly.
@@
 \begin{point}
-p12 "such as years, accompanied by the generation time". Might be worth
+p12 ``such as years, accompanied by the generation time". Might be worth
   detailing that it is needed as simulators work with generations and not
@@
 \begin{point}
-p12 "function defined by start and end sizes" You should detail here what are
+p12 ``function defined by start and end sizes" You should detail here what are
   the models of population expansion available and what is the default. Linear,
@@
 \begin{point}
-p12 "S are born via [...]" missing that S is a proportion
+p12 ``S are born via [...]" missing that S is a proportion
 \end{point}
@@
 \begin{point}
-p13-14 "Other implicit values [...]" until end of paragraph, this part with
+p13-14 ``Other implicit values [...]" until end of paragraph, this part with
   the example could be worth integrating into the main text as this is a major
@@
 \begin{point}
-p15 what's the difference between "genome annotations" and "sequence annotations"?
+p15 what's the difference between ``genome annotations" and ``sequence annotations"?
 \end{point}
@@
 \begin{reply}
-
 The field names in a YAML file are variables, not words \textit{per se}.
@@
 We feel that such conversion is beyond the scope of this project.
-
 \end{reply}
@@
 \begin{reply}
-
 While plain-text formats are ideally suited for use with version control
@@
 \begin{point}
-In their paper entitled "Demes: a standard format for demographic models",
+In their paper entitled ``Demes: a standard format for demographic models",
 Gower et al. propose a new way to describe (historical) demographic models of

### 2022-08-03 response-to-reviewers.tex
very minor changes

@@
 from the outset, which is why we have implemented several high-quality
-parsers (Python, and C with Rust, Julia in development) which should
+parsers (Python, C, Rust, and Julia) which should
 cover the vast majority of use-cases.

### 2022-08-03 paper.tex
fix typo
very minor changes

@@
 simplicity and clarity.
-We provide thoroughly tested implementations of Demes parsers in Python
-and C, and showcase initial support in several simulators and inference
+We provide thoroughly tested implementations of Demes parsers
+in multiple languages including Python and C,
+and showcase initial support in several simulators and inference
 methods.
@@
 Population genetics is a large field, and rather than attempting to
-capture all possible within- and between- population processes, we have
+capture all possible within- and between-population processes, we have
 instead adopted a pragmatic approach of identifying a common set
@@
 value if given. Size values are inherited naturally following the
-the document. Size values are inherited naturally following the
 progression of time. For example, if an epoch \texttt{start\_size} is not
@@
 expected outputs. In addition, we have high-quality parser implementations in
-the Python and C languages (Table~\ref{tab:software})
+the Python, C, Rust, and Julia languages (Table~\ref{tab:software})
 providing a solid foundation for the software ecosystem.
@@
 Since the standard is based on JSON, and JSON documents can be
-arbitrarily nested, it is possible to imagine a simple specification
+arbitrarily nested, we can imagine a simple specification
 of genome features such as mutation and recombination rates

### 2022-08-03 software-table.tex
Linkify urls.
Also Americanize software-table.tex.
very minor changes

@@
 \texttt{demes-c} &
-    A C library for parsing Demes YAML descriptions.
+    A C library for parsing Demes YAML descriptions
     (\url{https://github.com/grahamgower/demes-c}). \\
 
+\texttt{demes-rust} &
+    A Demes parser in Rust
+    (\url{https://github.com/molpopgen/demes-rs}). \\
+
+\texttt{demes-julia} &
+    A parser in Julia
+    (\url{https://github.com/apragsdale/Demes.jl}). \\
+
 \texttt{demesdraw} &
-    A Python library for visualising Demes models
+    A Python library for visualizing Demes models
     (\url{https://github.com/grahamgower/demesdraw}). \\

### 2022-08-03 fig/showcase.py

@@ -117,8 +117,8 @@ if __name__ == "__main__":
     ax4 = plt.subplot2grid(grid, (2, 4), colspan=6, rowspan=3)
     img = mpimg.imread("minted-snippet.png")
     yy, xx, _ = img.shape
-    ax4.set_xlim(0, xx)
-    ax4.set_ylim(yy, 0)
+    ax4.set_xlim(-xx * 0.01, xx * 1.01)
+    ax4.set_ylim(yy * 1.01, -yy * 0.01)
     ax4.imshow(img, interpolation="none")
     ax4.set_frame_on(False)
     ax4.set_xticks([])

### 2022-08-08 software-table.tex
added author and software for MSMC
fixed table error on software-table, added citation

@@
 
+\MSMC & A script provided in the MSMC-tools repository (https://github.com/stschiff/msmc-tools)
+    converts MSMC \citep{SchiffelsDurbin2014,WangSchiffels2020} output to the demes format.
+Instructions are available in the Guide (https://github.com/stschiff/msmc-tools/blob/master/msmc-tutorial/guide.md).\\
+
 \msprime &

### 2022-08-08 software-table.tex

--- 2022-08-03 software-table.tex
@@
 
+\MSMC & A script provided in the MSMC-tools repository \url{https://github.com/stschiff/msmc-tools}
+    converts MSMC \citep{SchiffelsDurbin2014,WangSchiffels2020} output to the demes format.
+Instructions are available in the Guide \url{https://github.com/stschiff/msmc-tools/blob/master/msmc-tutorial/guide.md}.\\
+
 \msprime &

### 2022-08-08 paper.tex
added author and software for MSMC

@@
 \author[5]{Ekaterina Noskova}
+\author[8]{Stephan Schiffels}
 \author[3]{Travis J. Struck}
@@
     Discovery, University of Oxford}
-\affil[7]{Ecology and Evolutionary Biology, University of California at Irvine}
+\affil[7]{Ecology and Evolutionary Biology, University of California, Irvine}
+\affil[8]{Max Planck Institute for Evolutionary Anthropology, Leipzig, Germany}
 
@@
 Jerome Kelleher is supported by the Robertson Foundation.
+Stephan Schiffels was supported by funding from the European Research Council (ERC) under the European Union’s Horizon 2020 research and innovation programme (grant agreement No 851511).
 

### 2022-08-08 paper.tex
Fix build and trim MSMC text

@@
 \newcommand{\tskit}[0]{\texttt{tskit}}
+\newcommand{\MSMC}[0]{\texttt{MSMC}}
 \newcommand{\demesslim}[0]{\texttt{demes-slim}}
@@
     Discovery, University of Oxford}
-\affil[7]{Ecology and Evolutionary Biology, University of California, Irvine}
+\affil[7]{Ecology and Evolutionary Biology, University of California at Irvine}
 \affil[8]{Max Planck Institute for Evolutionary Anthropology, Leipzig, Germany}
@@
 \citep[e.g.,][]{gutenkunst2009inferring,li2011inference,
-gravel2012population,kamm2017efficient,
+gravel2012population,
+SchiffelsDurbin2014,
+kamm2017efficient,
 jouganous2017inferring,ragsdale2019models,

### 2022-08-08 paper.tex
Ack reviewers and tell editor about new author

@@
     Discovery, University of Oxford}
-\affil[7]{Ecology and Evolutionary Biology, University of California at Irvine}
+\affil[7]{Ecology and Evolutionary Biology, University of California, Irvine}
 \affil[8]{Max Planck Institute for Evolutionary Anthropology, Leipzig, Germany}
@@
 \section*{Acknowledgments}
+We would like to thank the editor and reviewers for helpful
+comments that have significantly improved this manuscript.
 Graham Gower was supported by a Villum Fonden Young Investigator award to Fernando Racimo (project no. 00025300).

### 2022-08-08 paper.tex

--- 2022-08-08 paper.tex
@@
 Graham Gower was supported by a Villum Fonden Young Investigator award to Fernando Racimo (project no. 00025300).
-Ryan Gutenkunst was supported by the National Institute of General Medical Sciences of the National Institutes of Health (R01GM127348 to RNG).
+Ryan Gutenkunst and Travis Struck were supported by the National Institute of General Medical Sciences of the National Institutes of Health (R01GM127348 to RNG).
 Matthew Hartfield is supported by a NERC Independent Research Fellowship (NE/R015686/1).

### 2022-08-09 paper.tex

@@
 \newcommand{\gadma}[0]{\texttt{GADMA}}
+\newcommand{\gIMble}[0]{\texttt{gIMble}}
 \newcommand{\tskit}[0]{\texttt{tskit}}
@@
 % Second Authors
-\author[3]{Ryan N. Gutenkunst}
-\author[4]{Matthew Hartfield}
+\author[3]{Gertjan Bisschop}
+\author[4]{Ryan N. Gutenkunst}
+\author[3]{Matthew Hartfield}
 \author[5]{Ekaterina Noskova}
 \author[8]{Stephan Schiffels}
-\author[3]{Travis J. Struck}
+\author[4]{Travis J. Struck}
 %Senior Authors
@@
 \affil[2]{Department of Integrative Biology, University of Wisconsin--Madison}
-\affil[3]{Department of Molecular and Cellular Biology, University of Arizona}
-\affil[4]{Institute of Ecology and Evolution, The University of Edinburgh}
+\affil[3]{Institute of Ecology and Evolution, The University of Edinburgh}
+\affil[4]{Department of Molecular and Cellular Biology, University of Arizona}
 \affil[5]{Computer Technologies Laboratory, ITMO University}
@@
 Stephan Schiffels was supported by funding from the European Research Council (ERC) under the European Union’s Horizon 2020 research and innovation programme (grant agreement No 851511).
+Gertjan Bisschop was supported by funding from the ERC(ModelGenomLand, 757648).
 

### 2022-08-09 software-table.tex
Fix build and trim MSMC text
Brackets for msmc url

--- 2022-08-03 software-table.tex
@@
 
+\gIMble &
+    Fits IM-type demographic models and infers genomic barriers to geneflow \citep{laetsch2022gimble}. Outputs inferred models in Demes format. \\
+
 \moments &
@@
     \citep{jouganous2017inferring,ragsdale2019models}. Models to be optimized can be specified in Demes.\\
+
+\MSMC & A script provided in the MSMC-tools repository
+(\url{https://github.com/stschiff/msmc-tools})
+converts MSMC \citep{SchiffelsDurbin2014,WangSchiffels2020} output to the demes format. 
+\\
 

### 2022-08-10 paper.tex
Basic changes for Genetics

@@
-\documentclass[11pt]{article}
-\usepackage[round,comma,authoryear]{natbib}
-\usepackage{fullpage}
-\usepackage{authblk}
-\usepackage{graphicx}
+\documentclass[9pt,twocolumn,twoside]{gsajnl}
+\articletype{inv} % article type
+
 \usepackage{booktabs}
@@
 % Import this last to avoid mysterious errors.
-\usepackage[implicit=false]{hyperref}
+% \usepackage[implicit=false]{hyperref}
 
@@
 \newcommand{\demesslim}[0]{\texttt{demes-slim}}
-
-\newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
-\newcommand{\jkcomment}[1]{{\textcolor{red}{JK: #1}}}
-\newcommand{\ggcomment}[1]{{\textcolor{yellow!60!red}{GG: #1}}}
-\newcommand{\krtcomment}[1]{{\textcolor{purple}{KRT: #1}}}
-\newcommand{\mhcomment}[1]{{\textcolor{cyan}{MH: #1}}}
 
@@
 %Senior Authors
-\author[6,$\dagger$,$\aleph$]{Jerome Kelleher}
+\author[6,$\dagger$,$\ddagger$]{Jerome Kelleher}
 \author[7,$\dagger$]{Kevin R. Thornton}
@@
 
-\affil[$\star$]{Denotes shared first authorship, listed alphabetically}
-\affil[$\dagger$]{Denotes shared senior authorship, listed alphabetically}
-\affil[$\aleph$]{Denotes corresponding author}
+% \affil[$\star$]{Denotes shared first authorship, listed alphabetically}
+% \affil[$\dagger$]{Denotes shared senior authorship, listed alphabetically}
+% \affil[$\aleph$]{Denotes corresponding author}
 
-\maketitle
+\keywords{Demographic models, inference, simulation}
+
+\runningtitle{Demes: a standard format for demographic models}
+\runningauthor{Gower \textit{et al.}}
 
@@
 
+\maketitle
+\thispagestyle{firststyle}
+\marginmark
+\firstpagefootnote
+
+% Use the \equalcontrib command to mark authors with equal
+% contributions, using the relevant superscript numbers
+% \equalcontrib{1}
+% \equalcontrib{2}
+\makeatletter
+\newcommand{\equalcontribfirst}[1]{\@authfootnote{#1}{These authors contributed equally to this work, as joint first authors.}}
+\newcommand{\equalcontriblast}[1]{\@authfootnote{#1}{These authors contributed equally to this work, as joint last authors.}}
+\makeatother
+
+\equalcontribfirst{$\star$} % {Denotes shared first authorship, listed alphabetically}
+\equalcontriblast{$\dagger$} % {Denotes shared senior authorship, listed alphabetically}
+% \equalcontrib{$\ddagger$}{Denotes corresponding author}
+
+\correspondingauthoraffiliation{$\ddagger$}{\url{jerome.kelleher@bdi.ox.ac.uk}}
+\vspace{-33pt}% Only used for adjusting extra space in the left column of the first page
 \section*{Introduction}
@@
 
-\bibliographystyle{genetics}
+% \bibliographystyle{genetics}
 \bibliography{paper}

### 2022-08-10 paper.tex
inline software table

@@
 \newcommand{\demesslim}[0]{\texttt{demes-slim}}
-
-\usepackage{catchfile}
-\begin{filecontents*}{software-table.tex}
-\end{filecontents*}
 
@@
 
-\CatchFileDef{\softwaretable}{software-table.tex}{}
 \renewcommand{\arraystretch}{1.5}
@@
         \begin{tabular}{lp{12cm}}
-            \softwaretable
+%%% Start software table
+
+\toprule
+\multicolumn{2}{l}{Software infrastructure}\\
+\midrule
+\texttt{demes-python} &
+    A Python library for loading, saving, and working with
+    Demes models. Includes support for converting to and from
+    \texttt{ms}~\citep{hudson2002generating}
+    (\url{https://github.com/popsim-consortium/demes-python}).\\
+
+\texttt{demes-c} &
+    A C library for parsing Demes YAML descriptions
+    (\url{https://github.com/grahamgower/demes-c}). \\
+
+\texttt{demes-rust} &
+    A Demes parser in Rust
+    (\url{https://github.com/molpopgen/demes-rs}). \\
+
+\texttt{demes-julia} &
+    A parser in Julia
+    (\url{https://github.com/apragsdale/Demes.jl}). \\
+
+\texttt{demesdraw} &
+    A Python library for visualizing Demes models
+    (\url{https://github.com/grahamgower/demesdraw}). \\
+
+\midrule
+\multicolumn{2}{l}{Methods using Demes as input/output format}\\
+\midrule
+
+% TOOL AUTHORS: please add a row to this table to summarise what the software
+% does, and how it implements demes.
+
+\dadi &
+   Optimizes parameters in models of demographic history and distributions of fitness effects using SFS~\citep{gutenkunst2009inferring}. Can simulate SFS from Demes models.
+    \\
+
+\demesslim &
+    Loads Demes models into the \slim\ forward simulator \citep{haller2019slim}.\\
+
+\fwdpy &
+   Simulates the Wright-Fisher model forward in time \citep{thornton2014cpp,thornton2019-nu}. Demes is the preferred format for specifying a demographic model.\\
+
+\gadma &
+    Infers models of demographic history \citep{noskova2020gadma}. Outputs Demes models and visualizations. \\
+
+\gIMble &
+    Fits IM-type demographic models and infers genomic barriers to geneflow \citep{laetsch2022gimble}. Outputs inferred models in Demes format. \\
+
+\moments &
+    Optimizes parameters in models of demographic history using SFS and linkage disequilibrium statistics
+    \citep{jouganous2017inferring,ragsdale2019models}. Models to be optimized can be specified in Demes.\\
+
+\MSMC & A script provided in the MSMC-tools repository
+(\url{https://github.com/stschiff/msmc-tools})
+converts MSMC \citep{SchiffelsDurbin2014,WangSchiffels2020} output to the demes format.
+\\
+
+\msprime &
+    Simulates population genetic models using tree sequences~\citep{kelleher2016efficient,kelleher2020coalescent,baumdicker2021-iu}. Demographic history models can be specified using Demes.\\
+
+\bottomrule
+
+%%% end software table
         \end{tabular}

### 2022-08-10 paper.tex
Tweaked software table width
Update figure 1

@@
     \begin{center}
-        \begin{tabular}{lp{12cm}}
+        \begin{tabular}{lp{6.5cm}}
 %%% Start software table
@@
 \bottomrule
-
 %%% end software table
@@
 expect methods to agree exactly. Arbitrary population
-processes
----for example, within-deme
+processes---for example, within-deme
 continuous spatial structure
@@
     \caption{}
-        \begin{tcolorbox}[equal height group=IM]
+        \begin{tcolorbox}
             \inputminted[fontsize=\scriptsize,numbersep=5pt]{yaml}{models/IM.yaml}
@@
 %
-    \begin{minipage}{0.50\textwidth}
+    \begin{minipage}{0.45\textwidth}
     \begin{subfigure}{\textwidth}
     \caption{}
-        \begin{tcolorbox}[equal height group=IM, boxsep=0pt]
+        \begin{tcolorbox}
             \includegraphics[width=\textwidth]{fig/IM}
@@
         Example isolation-with-migration Demes model. (A) The Human Data Model
-        representation expressed using YAML.
+        representation expressed as YAML.
         (B) A visual representation of the model using \demesdraw.
@@
 
-\begin{figure}[tb!]
+\begin{figure}
     \centering

### 2022-08-10 paper.tex
Remove line break in abstract

@@
 methods.
-An introduction to the file format and a detailed specification are available at:\\
-\url{https://popsim-consortium.github.io/demes-spec-docs/}.
+An introduction to the file format and a detailed specification are available
+at \url{https://popsim-consortium.github.io/demes-spec-docs/}.
 \end{abstract}
@@
 
-\begin{figure}
+\begin{figure*}
     \centering
@@
     \label{fig:showcase}
-\end{figure}
+\end{figure*}
 
@@
         % NOTE: line highlights may need adjusting after changing this code
-        \inputminted[fontsize=\scriptsize,numbersep=5pt,highlightlines={9,19,52}]{python}{models/tennessen-simulation.py}
+        \inputminted[fontsize=\scriptsize,numbersep=5pt,highlightlines={6,14,40}]{python}{models/tennessen-simulation.py}
     \end{tcolorbox}

### 2022-08-26 paper.tex
Addresses and fix affil ordering

@@
 \author[5]{Ekaterina Noskova}
-\author[8]{Stephan Schiffels}
+\author[6]{Stephan Schiffels}
 \author[4]{Travis J. Struck}
 %Senior Authors
-\author[6,$\dagger$,$\ddagger$]{Jerome Kelleher}
-\author[7,$\dagger$]{Kevin R. Thornton}
-
-\affil[1]{Section for Molecular Ecology and Evolution, Globe Institute, University of Copenhagen}
-\affil[2]{Department of Integrative Biology, University of Wisconsin--Madison}
-\affil[3]{Institute of Ecology and Evolution, The University of Edinburgh}
-\affil[4]{Department of Molecular and Cellular Biology, University of Arizona}
-\affil[5]{Computer Technologies Laboratory, ITMO University}
-\affil[6]{Big Data Institute, Li Ka Shing Centre for Health Information and
-    Discovery, University of Oxford}
-\affil[7]{Ecology and Evolutionary Biology, University of California, Irvine}
-\affil[8]{Max Planck Institute for Evolutionary Anthropology, Leipzig, Germany}
+\author[7,$\dagger$,$\ddagger$]{Jerome Kelleher}
+\author[8,$\dagger$]{Kevin R. Thornton}
+\affil[1]{Section for Molecular Ecology and Evolution, Globe Institute, University of Copenhagen, 1350 Copenhagen K, Denmark}
+\affil[2]{Department of Integrative Biology, University of Wisconsin--Madison, WI 53706, USA}
+\affil[3]{Institute of Ecology and Evolution, The University of Edinburgh, EH9 3FL, UK}
+\affil[4]{Department of Molecular and Cellular Biology, University of Arizona, Tucson, AZ 85721, USA}
+\affil[5]{Computer Technologies Laboratory, ITMO University, 197101, Saint-Petersburg, Russia}
+\affil[6]{Max Planck Institute for Evolutionary Anthropology, 04103 Leipzig, Germany}
+\affil[7]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, OX3 7LF, UK}
+\affil[8]{Ecology and Evolutionary Biology, University of California, Irvine, CA 92697, USA}
 

### 2022-08-26 paper.tex
Add Data availability section as requested
New sections for funding and conlict of interest

@@
 
+\section*{Data Availability}
+The Demes specification and documentation are available at
+\url{https://popsim-consortium.github.io/demes-spec-docs/}
+
 \section*{Acknowledgments}
@@
 comments that have significantly improved this manuscript.
+
+\section*{Funding}
 Graham Gower was supported by a Villum Fonden Young Investigator award to Fernando Racimo (project no. 00025300).
 Ryan Gutenkunst and Travis Struck were supported by the National Institute of General Medical Sciences of the National Institutes of Health (R01GM127348 to RNG).
-Matthew Hartfield is supported by a NERC Independent Research Fellowship (NE/R015686/1).
-Jerome Kelleher is supported by the Robertson Foundation.
+Matthew Hartfield was supported by a NERC Independent Research Fellowship (NE/R015686/1).
+Jerome Kelleher was supported by the Robertson Foundation.
 Stephan Schiffels was supported by funding from the European Research Council (ERC) under the European Union’s Horizon 2020 research and innovation programme (grant agreement No 851511).
 Gertjan Bisschop was supported by funding from the ERC(ModelGenomLand, 757648).
+
+\section*{Conflicts of interest}
+The authors declare that there is no conflict of interest.
 
